---
title: "Backends: brms, cmdstanr, rstanarm, and survey"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  How to feed svyder from fitted model objects instead of a raw draws matrix.
  Covers the extract_draws() and extract_design() generics, what each backend
  (brms, cmdstanr, rstanarm, posterior, survey) auto-extracts, and what you
  must always supply yourself --- above all, the cluster aggregation unit.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backends: brms, cmdstanr, rstanarm, and survey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", out.width = "85%",
                      warning = FALSE, message = FALSE)
library(svyder)
# Consistent colour palette across all svyder vignettes (colour-blind aware)
svyder_pal <- c(
  `I-a`  = "#D6604D",  # within-group fixed effects (survey-dominated)
  `I-b`  = "#4393C3",  # between-group fixed effects (shielded)  [brand blue]
  `II`   = "#5AAE61",  # random effects (shrinkage-protected)
  `III`  = "#9E9E9E",  # hyperparameters (excluded)
  flag   = "#B2182B",  # flagged
  target = "#1A1A1A"   # DER = 1 / threshold reference lines
)
```

## Overview

svyder computes Design Effect Ratios from a plain matrix of posterior draws.
This vignette is for practitioners who have already fitted a hierarchical model
with **brms**, **cmdstanr**, or **rstanarm**, and would rather hand svyder the
fit object than reassemble the draws by hand. Two S3 generics do the wiring:
`extract_draws()` pulls a standardized draws matrix out of a fit object or a
`posterior::draws_*` container, and `extract_design()` pulls survey weights,
cluster ids, and strata out of a `survey` design object. The `der_compute()`
methods for each backend call these generics for you.

One theme runs through every backend: svyder can auto-detect a model's likelihood
and parameters, but it can **never** infer the survey design from a Stan fit ---
a `brmsfit` does not know which primary sampling units produced its rows. You
therefore always pass `weights` and, most importantly, `cluster` yourself. And
the plain-matrix path always works with no extra packages installed, so it is the
reliable fallback whenever a backend is fussy.

We cover, in order:

- the matrix path (always available), which is exactly what the bundled demo uses;
- the three Stan front-ends --- brms, cmdstanr, rstanarm --- and what each does and does not auto-extract;
- the `survey` package via `extract_design()`, shown live on demo pieces;
- `posterior` draws objects;
- how to map your own model onto svyder's `X`, `group`, and `param_types`.

> **Key insight.** No backend supplies `cluster`. The aggregation unit is part
> of the *estimand* --- the DER changes materially between a design-PSU target
> and a model-group target --- so svyder requires you to declare it explicitly.
> This is a deliberate design decision, not a limitation of any one backend.

This vignette is about integration mechanics only. For the meaning of the DER,
the four-tier classification, and the correction algorithm, see
[Getting Started](getting-started.html) and
[The Compute--Classify--Correct Pipeline](pipeline.html).

## The matrix path (always available)

Every `der_compute()` method eventually funnels into `der_compute.matrix()`,
which takes a raw draws matrix (`S` samples $\times$ `d` parameters, columns
ordered `beta[1..p]` then `theta[1..J]`) plus the data and design pieces. The
bundled `nsece_demo` object **is** the matrix path pre-assembled: it stores the
draws alongside `y`, `X`, `group`, `weights`, `psu`, `family`, and
`sigma_theta`, so no model fitting or Stan installation is required.

Here is the canonical matrix call, run live on the demo. Note `cluster =
nsece_demo$psu` --- the design PSU ids --- which is required and has no default:

```{r matrix-path}
fit <- der_compute(
  nsece_demo$draws,                 # S x d draws matrix: [beta, theta]
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  cluster     = nsece_demo$psu,     # REQUIRED aggregation unit (design PSUs)
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types
)

fit
```

That is the whole contract for the matrix path: **you supply everything**. The
backend methods below exist only to fill in as many of these arguments as can be
recovered from a fit object, so you type less. What they can recover differs by
backend, but `weights` and `cluster` are never among them.

> **Note.** If you can produce a draws matrix in `[beta, theta]` column order
> --- from any sampler, including one svyder has no method for --- you never
> need a backend at all. Pass the matrix directly, as above.

## brms

`der_compute()` has a method for `brmsfit` objects. Because brms stores the full
model specification (data, formula, family, priors) inside the fit, this method
auto-extracts almost everything. Concretely, `der_compute.brmsfit()`:

- reads the **response** `y` and **design matrix** `X` from the stored model
  data and `brms::standata()`;
- infers **`family`** from `stats::family()` (brms `bernoulli`/`binomial`
  $\rightarrow$ `"binomial"`, `gaussian` $\rightarrow$ `"gaussian"`);
- detects the **`group`** factor from the first random-effect grouping term;
- estimates **`sigma_theta`** as the posterior mean of the first `sd_*`
  parameter (and `sigma_e` from `sigma`, for gaussian);
- extracts the **draws** via `extract_draws.brmsfit()`, which keeps population-level
  (`b_*`) and group-level (`r_*`) parameters, drops variance and correlation
  components (`sd_*`, `cor_*`, `sigma`, `lp__`, `lprior`), and renames them to
  svyder's `beta[k]` / `theta[j]` convention.

What you must still pass: **`weights`** (survey weights) and **`cluster`** (the
aggregation unit) --- brms fit objects carry no survey design. You may also pass
`group` to override the auto-detected grouping factor, or `sigma_theta` /
`family` to override the auto-detected values.

The following is illustrative and not evaluated here (it requires a compiled
Stan model):

```{r brms, eval = FALSE}
library(brms)

# A hierarchical logistic model fitted to survey data
bfit <- brm(
  y ~ poverty_cwc + tiered_reim + (1 | state),
  data   = provider_df,
  family = bernoulli()
)

# svyder auto-extracts y, X, group, family, sigma_theta from bfit;
# you supply the survey weights and the cluster aggregation unit.
der <- der_compute(
  bfit,
  weights = provider_df$w,          # survey weights (NOT in the fit)
  cluster = provider_df$psu         # REQUIRED: design PSU ids
)

der_diagnose(der)                   # classify + correct in one step
```

> **Key insight.** brms auto-detection is a convenience, not magic. It reads the
> *model*, never the *sample design*. If your PSUs are not the same as your model
> grouping factor, be deliberate: `cluster = provider_df$psu` gives a design-based
> target, `cluster = provider_df$state` gives a model-aligned one.

## cmdstanr

The `CmdStanMCMC` method is the most manual of the three, because CmdStan keeps
**no model data** in the fit object --- only the draws. `extract_draws.CmdStanMCMC()`
converts `fit$draws()` to a matrix and strips the diagnostic columns `lp__` and
`lp_approx__`; that is all cmdstanr can give svyder automatically.

Consequently `der_compute.CmdStanMCMC()` requires you to supply **every** data
argument yourself: `y`, `X`, `group`, `weights`, `cluster`, and `sigma_theta`
(plus `sigma_e` for gaussian, and optional `strata`, `normalize`, `param_types`).
Crucially, the draws matrix you fitted **must already be in svyder column order**,
`[beta_1, ..., beta_p, theta_1, ..., theta_J]`, because svyder cannot reorder
columns it cannot name.

Not evaluated (requires CmdStan):

```{r cmdstanr, eval = FALSE}
library(cmdstanr)

mod <- cmdstan_model("hierarchical_logit.stan")
cfit <- mod$sample(data = stan_data, parallel_chains = 4)

# cmdstanr stores no data, so supply ALL pieces explicitly.
der <- der_compute(
  cfit,
  y           = stan_data$y,
  X           = stan_data$X,            # columns match beta[1..p]
  group       = stan_data$group,
  weights     = stan_data$w,            # survey weights
  cluster     = stan_data$psu,          # REQUIRED aggregation unit
  family      = "binomial",
  sigma_theta = 0.66                    # or a posterior summary of your sd param
)
```

> **Note.** With cmdstanr the burden of getting `X` and the draw columns into the
> same order as `beta[1..p]` is on you. If in doubt, extract the draws yourself,
> reorder to `[beta, theta]`, and use the matrix path --- it is identical in
> result and easier to verify.

## rstanarm

The `stanreg` method sits between brms and cmdstanr in convenience. rstanarm
stores the data, formula, and family, so `der_compute.stanreg()` auto-extracts:

- **`y`** from `x$y` and **`X`** from `stats::model.matrix(x)`;
- **`family`** from `stats::family()` (`binomial` or `gaussian`);
- the **`group`** factor from `x$glmod$reTrms$flist` (the first grouping term);
- **`sigma_theta`** as `sqrt(mean(.))` of the first `Sigma[...]` variance
  component (rstanarm stores group-level *variances*, so svyder takes the square
  root), and **`sigma_e`** from `sigma` for gaussian;
- the **draws** via `extract_draws.stanreg()`, which keeps fixed effects
  (plain names such as `(Intercept)`, `x1`) and random effects (`b[...]`), drops
  `Sigma[...]`, `sigma`, and `log-posterior`, and renames to `beta[k]` /
  `theta[j]`.

As always, you provide **`weights`** and **`cluster`** yourself.

Not evaluated (requires a compiled Stan model):

```{r rstanarm, eval = FALSE}
library(rstanarm)

sfit <- stan_glmer(
  y ~ poverty_cwc + tiered_reim + (1 | state),
  data   = provider_df,
  family = binomial()
)

# rstanarm auto-extracts y, X, group, family, sigma_theta (and sigma_e);
# you supply weights and the cluster aggregation unit.
der <- der_compute(
  sfit,
  weights = provider_df$w,
  cluster = provider_df$psu           # REQUIRED
)
```

## The survey package

The `survey` package is the natural source of the two things every Stan backend
lacks: survey **weights** and the **cluster** (PSU) aggregation unit.
`extract_design()` has a method for `survey.design2` objects that returns exactly
those pieces --- plus strata --- ready to hand to `der_compute()`.

Because the `survey` package is available here, we can demonstrate this **live**.
We build a minimal single-stage design from `nsece_demo` pieces: a data frame
holding the PSU id and the weight, wrapped by `survey::svydesign()`. (In a real
analysis this is the design object you already have; we reconstruct a small one
purely so the example runs without external data.)

```{r survey-build}
library(survey)

# A data frame carrying the two design columns svyder needs
df_des <- data.frame(
  psu = nsece_demo$psu,
  w   = nsece_demo$weights
)

des <- svydesign(ids = ~psu, weights = ~w, data = df_des)
class(des)
```

Now call `extract_design()` on it and inspect what comes back:

```{r survey-extract}
di <- extract_design(des)

str(di, max.level = 1)
```

`extract_design.survey.design2()` returns a three-element list:

- **`weights`** --- the survey weights (`stats::weights(des)`), one per row;
- **`cluster`** --- the PSU identifiers from the first stage
  (`des$cluster[[1]]`); svyder later coerces these to consecutive integer ids;
- **`strata`** --- a data frame of stratum variables, or (for this unstratified
  design) a single column of constant values.

```{r survey-extract-detail}
cat(sprintf("weights: %d values, mean %.3f\n",
            length(di$weights), mean(di$weights)))
cat(sprintf("cluster: %d unique PSUs\n", length(unique(di$cluster))))
cat(sprintf("strata : %d column(s)\n", ncol(di$strata)))
```

That gives 644 unique PSUs --- exactly the design-based aggregation unit for
`nsece_demo`. You could stop here and pass `di$weights` and `di$cluster` into the
matrix call by hand. But `der_compute()` accepts the design object directly
through its `design =` argument and does the extraction internally, which is the
tidiest route. This runs **live** on the demo draws:

```{r survey-compute}
fit_des <- der_compute(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types,
  design      = des                   # supplies weights + cluster + strata
)

# The three fixed-effect DERs, recovered via the survey design object
round(fit_des$der[1:3], 3)
```

These are the design-based DERs for the three fixed effects: the intercept and
tiered-reimbursement coefficients (both between-state) sit well below 1, while
`beta[2]`, the within-state poverty coefficient, is flagged. Passing `design =
des` reproduces the same fixed-effect DERs as passing `cluster = nsece_demo$psu`
explicitly, because the design's PSUs *are* those cluster ids --- the design
object is simply a convenient carrier for them.

What the design records is stored in `fit_des$target`:

```{r survey-target}
fit_des$target[c("cluster_n", "strata_n", "normalize")]
```

> **Note.** When a `survey.design2` carries genuine strata (more than one
> stratum), `der_compute()` picks them up automatically and switches the meat
> estimator to its stratified form. `nsece_demo` has no strata, so `strata_n` is
> 1 here and the estimator is unstratified. Weights still follow the `normalize`
> convention (`"unit_mean"` by default), recorded alongside the cluster count.

## Posterior draws objects

If you already hold your draws as a `posterior` package object, you do not need a
model-specific backend at all. `extract_draws()` has methods for every
`posterior::draws_*` format --- `draws_matrix`, `draws_df`, `draws_array`,
`draws_list`, and `draws_rvars` --- each of which converts to a plain matrix and
drops the `.chain`, `.iteration`, and `.draw` bookkeeping columns. The
`matrix` method is the identity: a matrix is already draws.

```{r posterior, eval = FALSE}
library(posterior)

# Suppose you assembled your draws as a posterior draws_matrix,
# with columns already ordered [beta[1..p], theta[1..J]].
dm <- as_draws_matrix(my_draws)

der <- der_compute(
  as.matrix(dm),                      # or feed dm through extract_draws()
  y = y, X = X, group = group,
  weights = w, cluster = psu,         # REQUIRED
  family = "binomial", sigma_theta = 0.66
)
```

Because these methods only reshape draws, they carry the same requirement as the
matrix path: the columns must be in svyder order, and you still supply all the
data and design arguments, `cluster` included.

## Mapping your model to X / group / param_types

Whichever backend you use, svyder ultimately needs three model-structure inputs
lined up correctly. Getting them right matters more than which front-end
produced the fit.

**`X` --- the fixed-effect design matrix.** Its columns must correspond, in
order, to the `beta[1..p]` draws. brms and rstanarm build `X` for you from the
formula; with cmdstanr or the matrix path you pass it yourself and are
responsible for the column ordering.

**`group` --- the grouping factor.** An integer (or coercible) vector, one value
per observation, giving each row's group. It defines the `theta[1..J]` random
effects and the shrinkage structure. brms and rstanarm read it from the first
random-effect term; cmdstanr and the matrix path take it directly.

**`param_types` --- how each fixed effect is identified.** A character vector,
one entry per column of `X`, taking values `"fe_within"` or `"fe_between"`:

- `"fe_within"` --- the covariate varies **within** groups (e.g. an
  individual-level poverty measure). Under Theorem 1 its DER equals the full
  design effect (Tier I-a) and it is the primary correction candidate.
- `"fe_between"` --- the covariate is **constant within** each group (e.g. a
  state-level policy indicator). Its DER is $\mathrm{DEFF}\,(1 - B)$, shielded by
  shrinkage (Tier I-b), and it is usually below the threshold.

If you leave `param_types = NULL`, svyder infers it from `X` --- any column that
is constant within every group is labelled `fe_between`, the rest `fe_within` ---
and emits a message so you can check the inference, for example:

```
param_types inferred from the design matrix: intercept=fe_between,
poverty_cwc=fe_within, tiered_reim=fe_between. Supply 'param_types' to override.
```

Passing `param_types` explicitly suppresses the message. Inference is reliable
for the demo, but a within-group covariate that happens to be constant within
every realized group in *your* sample would be mislabelled `fe_between`; when in
doubt, state `param_types` yourself. In `nsece_demo` the mapping is:

```{r param-types-demo}
data.frame(
  column     = colnames(nsece_demo$X),
  param_type = nsece_demo$param_types
)
```

> **Key insight.** `param_types` is about the covariate's *variation structure*,
> not about statistical significance or effect size. A within-group covariate is
> "survey-exposed" because clustered sampling directly inflates the variance of
> the contrasts that identify it; a between-group covariate is shielded because
> hierarchical shrinkage absorbs part of the design effect. This is the crux of
> the whole diagnostic, and it is set here, at the mapping step.

## What's next?

You can now feed svyder from any of its supported backends, or bypass them with a
plain draws matrix. The natural next steps:

| Vignette | What you will learn |
|:---------|:--------------------|
| [Getting Started](getting-started.html) | The full diagnostic in one call, read from `print`, `tidy`, `glance`, and one profile plot. |
| [The Compute--Classify--Correct Pipeline](pipeline.html) | Step through `der_compute` $\rightarrow$ `der_classify` $\rightarrow$ `der_correct`, tune the threshold, and see why selective correction beats blanket correction. |

For the design-target decision that the `cluster =` argument encodes --- design
PSUs versus model groups, weight conventions, and strata --- see
`vignette("choosing-the-target")`. For the theory behind `param_types` and the
four tiers, see `vignette("theory-decomposition")` and
`vignette("theory-sandwich")`.
