---
title: "Case Study: NSECE-Style Survey Data"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A complete, start-to-finish DER analysis on the bundled synthetic
  NSECE-style survey data: explore the design, run the diagnostic, interpret
  fixed and random effects, read the decomposition, correct selectively, compare
  variance targets, and write up the result for publication.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: NSECE-Style Survey Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", out.width = "85%",
                      warning = FALSE, message = FALSE)
library(svyder)
# Consistent colour palette across all svyder vignettes (colour-blind aware)
svyder_pal <- c(
  `I-a`  = "#D6604D",  # within-group fixed effects (survey-dominated)
  `I-b`  = "#4393C3",  # between-group fixed effects (shielded)  [brand blue]
  `II`   = "#5AAE61",  # random effects (shrinkage-protected)
  `III`  = "#9E9E9E",  # hyperparameters (excluded)
  flag   = "#B2182B",  # flagged
  target = "#1A1A1A"   # DER = 1 / threshold reference lines
)
```

## Overview

This vignette walks through a full Design Effect Ratio (DER) analysis from
first look to publication-ready write-up, using the survey data bundled with
**svyder**. It is written for practitioners: by the end you will be able to
take a fitted Bayesian hierarchical model on complex survey data, decide which
parameters the design actually distorts, correct exactly those, and report the
result defensibly.

We work throughout on `nsece_demo`, a **synthetic** dataset modelled on the
2019 National Survey of Early Care and Education (NSECE).

> **Note.** `nsece_demo` is *not* the real NSECE data. It is a synthetic,
> NSECE-*like* dataset bundled so every example runs with no Stan installation
> and no data-use agreement. Its structure mirrors the paper's application (6,785
> providers, 51 states), but its posterior draws, weights, and therefore its DER
> values differ from the published analysis. Every number this vignette
> *computes* is a live demo number; whenever we quote the paper's real NSECE
> findings we say so explicitly and never present them as demo output. For the
> real analysis, see Lee, Williams, and Savitsky (2026).

We cover, in order:

- **The design** --- providers, states, PSUs, unequal weights, and the global
  Kish design effect.
- **Exploration** --- weight distribution, PSU structure, and group sizes.
- **The diagnostic** --- one call to `der_diagnose()` and how to read it.
- **Fixed effects** --- why the within-state coefficient is flagged and the two
  between-state coefficients are not.
- **Random effects** --- why "protection" of the state effects depends on the
  declared target.
- **Decomposition** --- reading `der_decompose()` to see *why* each DER lands
  where it does.
- **Selective correction** --- widening only the flagged interval.
- **Comparing targets** --- a robustness report across two estimands.
- **Reporting** --- a template paragraph you can adapt.

Where the theory is only sketched here, we link out to the method track:
`vignette("theory-decomposition")` and
`vignette("theory-classification-correction")`.

## Background: the design

The model is a hierarchical logistic regression. Each provider $i$ in state
$j[i]$ contributes a binary outcome, and each state carries a random intercept:

$$
y_i \mid \beta, \theta_{j[i]} \sim
\mathrm{Bernoulli}\!\left(\mathrm{logit}^{-1}\!\left(x_i^\top \beta + \theta_{j[i]}\right)\right),
\qquad \theta_j \sim N\!\left(0, \sigma_\theta^2\right).
$$

The design matrix $X$ has three columns: an `intercept`, `poverty_cwc` (a
cluster-mean-centred poverty measure that varies *within* states), and
`tiered_reim` (a state policy indicator that varies only *between* states). The
distinction between within- and between-state covariates is the hinge of the
whole analysis, so `nsece_demo` records it explicitly:

```{r design-facts}
data(nsece_demo)

data.frame(
  covariate  = colnames(nsece_demo$X),
  param_type = nsece_demo$param_types,
  varies     = c("(constant)", "within states", "between states")
)

N <- nsece_demo$N          # providers
J <- nsece_demo$J          # states
n_psu <- length(unique(nsece_demo$psu))
sprintf("N = %d providers | J = %d states | %d PSUs | sigma_theta = %.2f",
        N, J, n_psu, nsece_demo$sigma_theta)
```

The survey weights are unequal, and unequal weights are the reason a
design-based variance can diverge from the model-based one. The single most
useful one-number summary of that inequality is the **Kish global design
effect**,

$$
\mathrm{DEFF} = \frac{N \sum_i w_i^2}{\left(\sum_i w_i\right)^2},
$$

which we can read straight off the weights:

```{r kish-deff}
w <- nsece_demo$weights
deff_kish <- length(w) * sum(w^2) / sum(w)^2
round(deff_kish, 3)
```

A Kish DEFF of `r round(deff_kish, 3)` means the weighted sample carries only
about `r round(100 / deff_kish)`% of the information an equally sized *simple
random sample* would provide --- a roughly `r round(deff_kish, 1)`-fold variance
inflation before any model is fitted. That inflation is the raw material the DER
redistributes across parameters.

> **Key insight.** The global DEFF is a *ceiling*, not a verdict. Theorem 1
> (see `vignette("theory-decomposition")`) shows a within-state fixed effect
> absorbs the full DEFF while a between-state effect absorbs only
> $\mathrm{DEFF}\,(1-B)$. So a single design can leave one coefficient badly
> under-covered and another perfectly calibrated --- exactly what we will find.

**How this compares with the real NSECE.** For orientation only: the published
analysis (Lee, Williams, & Savitsky, 2026) uses the *actual* NSECE 2019 file ---
6,785 providers and 51 states, as here, but drawn from **415 PSUs nested in 30
strata** with a global **DEFF of 3.76**. The synthetic `nsece_demo` collapses
that two-stage design to 644 single-stage PSUs and a milder DEFF of
`r round(deff_kish, 2)`, and has no strata slot. Treat the numbers below as an
illustration of the *workflow*, not a replication of the paper.

## Explore the design

Before running any diagnostic, look at the ingredients that drive the DER:
the spread of the weights, the size of the PSUs, and the size of the states.

### Weight distribution

```{r fig-weights, fig.height = 4, fig.cap = "Distribution of survey weights (rescaled to mean 1). The long right tail is what inflates the design-based variance; a perfectly equal-weight design would be a single bar at 1."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  ggplot(data.frame(w = w), aes(x = w)) +
    geom_histogram(bins = 40, fill = svyder_pal[["I-b"]],
                   colour = "white", linewidth = 0.2) +
    geom_vline(xintercept = 1, linetype = "dashed",
               colour = svyder_pal[["target"]]) +
    labs(x = "Survey weight (mean-1 scale)", y = "Providers",
         title = "Unequal weights drive the design effect",
         subtitle = sprintf("Kish DEFF = %.2f; dashed line marks weight = 1",
                            deff_kish)) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
```

```{r weight-summary}
round(summary(w), 3)
```

The weights range from near zero to well above 20. It is that inequality --- not
the mean --- that `DEFF` measures.

> **Note.** Because `nsece_demo`'s weights are already within-group normalised
> with mean 1, the three `normalize` conventions (`"unit_mean"`, `"group_size"`,
> `"none"`) give *identical* DERs on this dataset. So this vignette does **not**
> use the demo to illustrate weight-convention sensitivity. In the real NSECE,
> the convention genuinely matters --- it can even reverse which target flags
> more parameters (Lee et al., 2026); see
> [Choosing the Variance Target](choosing-the-target.html).

### PSU structure

The primary sampling units (PSUs) are the aggregation unit for the design-based
variance: the sandwich "meat" sums squared *cluster* score totals, so PSU size
governs how much the clustering costs.

```{r psu-structure}
psu_sizes <- as.numeric(table(nsece_demo$psu))
cat(sprintf("%d PSUs | providers per PSU: min %d, median %d, max %d\n",
            length(psu_sizes), min(psu_sizes),
            as.integer(median(psu_sizes)), max(psu_sizes)))
table(psu_sizes)
```

Most PSUs hold about a dozen providers. This tidy, roughly balanced clustering
is deliberate: it keeps the demo's per-cluster score totals well behaved so the
numbers are stable and reproducible.

### Group (state) sizes

The 51 states are the model's grouping factor and define the random effects.
Their sizes are far more skewed than the PSUs:

```{r fig-groups, fig.height = 4, fig.cap = "Providers per state. A handful of large states sit alongside many small ones; small states have noisy per-state DERs because each contributes only one cluster-score contrast."}
group_sizes <- as.numeric(table(nsece_demo$group))

if (requireNamespace("ggplot2", quietly = TRUE)) {
  gdf <- data.frame(state = seq_along(group_sizes),
                    n = sort(group_sizes, decreasing = TRUE))
  ggplot(gdf, aes(x = reorder(factor(state), -n), y = n)) +
    geom_col(fill = svyder_pal[["II"]], width = 0.8) +
    labs(x = "State (ordered by size)", y = "Providers",
         title = "State sizes are highly unequal") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())
}

cat(sprintf("Providers per state: min %d, median %d, max %d\n",
            min(group_sizes), as.integer(median(group_sizes)),
            max(group_sizes)))
```

The imbalance matters twice over: large states have high shrinkage factors $B$
(their random effects lean on their own data), and small states yield noisy
per-state DERs --- worth remembering when we reach the random effects.

## Run the diagnostic

The full pipeline --- compute the sandwich and posterior variances, classify
each parameter into a tier, and correct the flagged ones --- runs in a single
call to `der_diagnose()`. The one argument that defines the *estimand* is
`cluster`: it names the aggregation unit of the design-based variance.

> **Key insight.** `cluster` is **required** and has no default. Pass the design
> PSU ids for a design-based target (as here), or `group` for a model-aligned
> target. Never use the deprecated `psu =` argument. We compare the two targets
> in the [robustness section](#compare-targets) below.

```{r diagnose}
result <- der_diagnose(
  x           = nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  cluster     = nsece_demo$psu,          # design-based target (PSU ids)
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types,
  tau         = 1.2
)
```

The `print` method is the diagnostic at a glance:

```{r print-result}
result
```

Read it top to bottom: the target is `r result$target$cluster_n` clusters with a
centred, DF-corrected meat and mean-1 weights; the DER range spans the flagged
and shrinkage-protected extremes; `sigma_theta` sits in **Tier III** (its DER is
undefined by construction, so it is reported but never certified); and
`r sum(result$classification$flagged)` of `r nrow(result$classification)`
parameters clear the threshold $\tau = 1.2$.

For a one-row model summary, use `glance()`:

```{r glance}
glance(result)
```

The `mean_deff` (`r round(glance(result)$mean_deff, 3)`) is the average
*per-state* design effect, close to the global Kish `r round(deff_kish, 3)`;
`mean_B` (`r round(glance(result)$mean_B, 3)`) is the average shrinkage
factor. Together these two numbers explain the entire DER profile, as the
decomposition will make explicit.

## Fixed effects

Look first at the three fixed effects. The `tidy()` method returns one row per
parameter; we take the first three:

```{r fe-tidy}
td <- tidy(result)
fe <- td[1:3, c("term", "der", "tier", "action")]
fe$covariate <- colnames(nsece_demo$X)
fe[, c("term", "covariate", "der", "tier", "action")]
```

The pattern is stark and it is not an accident:

- **`beta[2]`** --- `poverty_cwc`, the *within*-state covariate --- has DER
  **`r round(td$der[2], 3)`**, classified **Tier I-a**, and is **flagged for
  correction**. Its model-based posterior variance is only about
  `r round(100 / td$der[2])`% of what the design-based sandwich says it should
  be.
- **`beta[1]`** --- the intercept, a *between*-state quantity --- has DER
  **`r round(td$der[1], 3)`**, **Tier I-b**, and is **retained**.
- **`beta[3]`** --- `tiered_reim`, the *between*-state policy covariate --- has
  DER **`r round(td$der[3], 3)`**, **Tier I-b**, and is **retained**.

This is Theorem 1 made visible. A within-group covariate draws all its
identifying variation from contrasts *inside* clusters, so it is fully exposed
to the clustering and $\mathrm{DER} \approx \mathrm{DEFF}$. A between-group
covariate is identified by comparisons *across* clusters, which the hierarchical
model already shields through shrinkage, so $\mathrm{DER} \approx
\mathrm{DEFF}(1-B)$ --- comfortably below 1 here.

> **Key insight.** In the *same* model, the survey design badly under-covers the
> within-state coefficient while leaving the two between-state coefficients
> already conservative. A blanket correction that rescaled all three toward the
> sandwich target would *narrow* the two between-state intervals --- the opposite
> of what a design correction is for. The full theory is in
> `vignette("theory-decomposition")`.

## Random effects

Now the 51 state random effects. Under this design-based target, a large
fraction are flagged:

```{r re-summary}
re <- td[grepl("^theta", td$term), ]
cat(sprintf("Random effects flagged (design-PSU target): %d of %d\n",
            sum(re$flagged), nrow(re)))
cat(sprintf("Median random-effect DER: %.3f\n", median(re$der)))
```

`r sum(re$flagged)` of `r nrow(re)` state effects clear the threshold, with a
median DER of `r round(median(re$der), 3)`. A profile plot shows the whole
model at once --- fixed effects and every state effect, coloured by tier, with
the threshold as a dashed line:

```{r fig-profile, fig.height = 5, fig.cap = "DER profile. Each point is one parameter, coloured by tier; the dashed line is tau = 1.2. beta[2] (Tier I-a) sits well above the threshold, the two between-state fixed effects (Tier I-b) sit far below, and the state random effects (Tier II) straddle it under this design-based target."}
plot(result, type = "profile")
```

Whether these state effects are "protected" is not a fixed property of the
model --- it depends on what you declared the target to be. Re-run the diagnostic
with the *model-aligned* target (`cluster = group`) and the picture shifts:

```{r re-model-target}
result_grp <- der_diagnose(
  x           = nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  cluster     = nsece_demo$group,        # model-aligned target
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types,
  tau         = 1.2
)

td_grp <- tidy(result_grp)
re_grp <- td_grp[grepl("^theta", td_grp$term), ]
cat(sprintf("Random effects flagged (model-group target): %d of %d\n",
            sum(re_grp$flagged), nrow(re_grp)))
cat(sprintf("Median random-effect DER: %.3f\n", median(re_grp$der)))
```

Under the model-group target only `r sum(re_grp$flagged)` of `r nrow(re_grp)`
state effects are flagged, and the median RE DER drops to
`r round(median(re_grp$der), 3)` --- around the value where the design and the
model agree.

> **Key insight.** This is the paper's central message. Random effects are
> *conditionally* shrinkage-protected only when the aggregation unit matches the
> model's grouping. Align the target to the design PSUs and the same effects can
> become design-sensitive. There is no target-free answer to "are the random
> effects protected?" --- the estimand has to be declared first.

> **Note.** Per-state DERs are **noisy**. Each state contributes exactly one
> cluster-score-total contrast to its own random-effect variance, so individual
> $\theta_j$ DERs --- especially for small states --- carry real Monte Carlo and
> finite-sample error. Read the random effects as a *distribution* (how many
> flagged, the median), not as 51 precise per-state verdicts.

## Decomposition

The `der_decompose()` function opens up *why* each DER lands where it does. It
reports, per parameter, the design effect used, the shrinkage factor $B$, the
between-group variation share $R_k$, the finite-group correction $\kappa$, and
the DER the closed-form theorems *predict*:

```{r decompose-fe}
dec <- der_decompose(result)
dec_fe <- dec[dec$param_type %in% c("fe_within", "fe_between"), ]
round(dec_fe[, c("der", "deff_used", "R_k", "der_predicted")], 3)
```

Read across the `R_k` column --- the between-group share of each covariate's
identifying variation:

- `beta[2]` (poverty, within): $R_k = `r round(dec_fe$R_k[dec_fe$param == "beta[2]"], 3)`$
  --- essentially all its information is *within* states, so
  $\mathrm{DER} \approx \mathrm{DEFF}$.
- `beta[1]` and `beta[3]` (between): $R_k \approx `r round(dec_fe$R_k[dec_fe$param == "beta[3]"], 2)`$
  --- almost all *between* states, so the shrinkage shields them and the
  predicted DER falls well below 1.

That single column cleanly separates the two covariate types: $R_k \approx 0.01$
for the within-state coefficient versus $R_k \approx 0.93$ for the between-state
ones. The predicted DERs sit right on top of the empirical ones, which is the
decomposition doing its job.

The random-effect rows populate $B$ and $\kappa$ instead of $R_k$ (Theorem 2
governs them):

```{r decompose-re}
theta1 <- dec[dec$param == "theta[1]", ]
round(theta1[, c("der", "deff_used", "B_used", "kappa", "der_predicted")], 3)
```

For `theta[1]`, the marginal random-effect DER is
$B \cdot \mathrm{DEFF} \cdot \kappa(J)$ --- the shrinkage factor and the
finite-group correction together pull the raw design effect down toward the
predicted value. The mechanics are derived in
`vignette("theory-decomposition")`.

## Selective correction and intervals

`der_diagnose()` already applied the default block-Cholesky correction to the
flagged set (Algorithm 1 in the paper). Its effect on any single parameter is a
widening of the credible interval by $\sqrt{\mathrm{DER}}$ --- and, crucially,
the unflagged parameters are left **bitwise untouched**.

Compare the original and corrected 90% intervals for the flagged fixed effect
`beta[2]` against an unflagged one, `beta[1]`:

```{r ci-compare}
draws_corr <- as.matrix(result)              # corrected draws (flagged block only)
draws_orig <- result$original_draws

ci <- function(x) quantile(x, c(0.05, 0.95))

tab <- rbind(
  `beta[1] original`  = ci(draws_orig[, 1]),
  `beta[1] corrected` = ci(draws_corr[, 1]),
  `beta[2] original`  = ci(draws_orig[, 2]),
  `beta[2] corrected` = ci(draws_corr[, 2])
)
round(tab, 3)
```

```{r ci-widths}
w2_orig <- diff(ci(draws_orig[, 2]))
w2_corr <- diff(ci(draws_corr[, 2]))
cat(sprintf("beta[2] 90%% CI width: %.3f -> %.3f (ratio %.2f)\n",
            w2_orig, w2_corr, w2_corr / w2_orig))
cat(sprintf("beta[1] draws identical after correction: %s\n",
            identical(draws_orig[, 1], draws_corr[, 1])))
```

The flagged `beta[2]` interval widens by a factor of about
`r round(w2_corr / w2_orig, 2)` --- exactly its scale factor
$\sqrt{`r round(td$der[2], 3)`}$ --- while the unflagged `beta[1]` draws come
back identical to the input.

```{r fig-ci, fig.height = 3, fig.cap = "Original vs corrected 90% credible interval for the flagged coefficient beta[2]. Correction shifts nothing and widens the interval to match the design-based variance; unflagged parameters (not shown) are unchanged."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  cidf <- data.frame(
    version = factor(c("Original", "Corrected"),
                     levels = c("Original", "Corrected")),
    lo = c(ci(draws_orig[, 2])[1], ci(draws_corr[, 2])[1]),
    md = c(median(draws_orig[, 2]), median(draws_corr[, 2])),
    hi = c(ci(draws_orig[, 2])[2], ci(draws_corr[, 2])[2])
  )
  ggplot(cidf, aes(x = md, y = version, colour = version)) +
    geom_pointrange(aes(xmin = lo, xmax = hi), linewidth = 0.9, size = 0.6) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = svyder_pal[["target"]]) +
    scale_colour_manual(values = c(Original = svyder_pal[["III"]],
                                   Corrected = svyder_pal[["I-a"]]),
                        guide = "none") +
    labs(x = "beta[2] (poverty_cwc), 90% CI", y = NULL,
         title = "Selective correction widens only the flagged interval") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
```

These corrected draws are drop-in: use them for posterior summaries, credible
intervals, or any downstream derived quantity, knowing the design correction has
been applied to precisely the parameters that needed it and to no others.

## Compare targets {#compare-targets}

Because the target is part of the estimand, good practice is to report the DER
under *both* plausible aggregation units and let the reader see the sensitivity.
`der_compare()` does this in one call:

```{r compare}
cmp <- der_compare(
  result,
  clusters = list(
    design_psu  = nsece_demo$psu,
    model_group = nsece_demo$group
  )
)

# reshape the fixed effects to a side-by-side view
fe_cmp <- cmp[cmp$param %in% c("beta[1]", "beta[2]", "beta[3]"), ]
fe_wide <- reshape(fe_cmp, idvar = "param", timevar = "cluster_name",
                   direction = "wide")
names(fe_wide) <- sub("der\\.", "", names(fe_wide))
fe_wide$covariate <- colnames(nsece_demo$X)
round_cols <- c("design_psu", "model_group")
fe_wide[round_cols] <- lapply(fe_wide[round_cols], round, 3)
fe_wide[, c("param", "covariate", "design_psu", "model_group")]
```

The flagged within-state coefficient `beta[2]` has DER
**`r round(cmp$der[cmp$param == "beta[2]" & cmp$cluster_name == "design_psu"], 3)`**
under the design-PSU target versus
**`r round(cmp$der[cmp$param == "beta[2]" & cmp$cluster_name == "model_group"], 3)`**
under the model-group target --- both above the threshold, so the *headline
finding is robust to the target choice* even though the magnitude moves. The
between-state coefficients stay well below 1 under either target.

> **Note.** In the real NSECE the target choice is even more consequential: the
> paper reports **1 of 54** parameters flagged under the model-group target
> versus **28 of 54** under the design-PSU target (Lee et al., 2026). The
> synthetic demo compresses that gap (30 vs 25), but the *lesson* is the same:
> declare the target, and report the comparison. The dedicated treatment is in
> [Choosing the Variance Target](choosing-the-target.html).

## Reporting for publication

A DER analysis is only as credible as its write-up. Because the DER depends on
choices a reader cannot see in a coefficient table --- the declared target, the
weight convention, the threshold --- those choices belong in the methods text.
Here is a template paragraph you can adapt, filled with this vignette's live
demo numbers:

> *We assessed the design sensitivity of each model parameter with the Design
> Effect Ratio (DER), the ratio of a declared design-based (sandwich) variance
> to the model-based posterior variance (Lee, Williams, & Savitsky, 2026). We
> declared a design-based target aggregating scores to the*
> `r result$target$cluster_n` *survey PSUs, with survey weights rescaled to mean
> one and a stratum-centred, degrees-of-freedom-corrected meat. Parameters with
> DER exceeding* `r result$tau` *were flagged as survey-sensitive. Of the*
> `r nrow(result$classification)` *parameters,*
> `r sum(result$classification$flagged)` *were flagged --- most prominently the
> within-state poverty coefficient (DER =*
> `r round(td$der[2], 3)`*), which is identified by within-cluster contrasts and
> therefore fully exposed to the clustering. We applied a selective
> block-Cholesky correction to the flagged set only, widening the poverty
> coefficient's 90% credible interval by a factor of*
> `r round(w2_corr / w2_orig, 2)` *while leaving the well-calibrated
> between-state coefficients and the shrinkage-protected majority of state
> random effects unchanged. As a robustness check we recomputed all DERs under a
> model-aligned target (aggregating to states); the poverty coefficient remained
> flagged under both targets.*

Adapt it by stating: (1) the **declared target** and why (design PSUs vs model
groups); (2) the **weight convention** (`normalize`); (3) the **threshold** and
any sensitivity check (`der_sensitivity()`); (4) the **flagged set** and the
**selective** nature of the correction; and (5) that unflagged parameters were
left untouched. If your design has strata or a known empty-PSU universe, report
`strata` and `cluster_strata` too (see `vignette("theory-sandwich")`).

## What's next?

You now have the full applied loop: explore, diagnose, interpret, decompose,
correct, compare, and report.

| Where to go | What you will find |
|:------------|:-------------------|
| [Choosing the Variance Target](choosing-the-target.html) | The target decision in depth: `cluster`, `normalize`, `strata`, `cluster_strata` |
| [Backends](backends.html) | Feeding real `brms` / `cmdstanr` / `rstanarm` fits and `survey` designs into svyder |
| `vignette("theory-decomposition")` | The math behind the fixed- and random-effect DERs: Theorems 1 and 2, $R_k$, $B$, $\kappa(J)$, and the conservation law |
| `vignette("theory-classification-correction")` | The four tiers, the threshold rationale, and why blanket correction is harmful |

For the real NSECE 2019 analysis this synthetic case study is patterned on, see
Lee, Williams, and Savitsky (2026).
