---
title: "Case Study: NSECE Survey Data Analysis"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A comprehensive case study demonstrating the full DER diagnostic workflow on
  NSECE-like survey data: explore the survey design, run diagnostics, interpret
  fixed and random effect results, verify decomposition theorems, compare
  selective vs blanket correction, and report results for publication.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: NSECE Survey Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(svyder)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
set.seed(42)

# Project colour palette
pal <- c(
  tier_ia    = "#D6604D",
  tier_ib    = "#4393C3",
  tier_ii    = "#1B7837",
  threshold  = "#762A83"
)
```

## Overview

This vignette walks through a complete DER analysis on the `nsece_demo`
dataset, a synthetic dataset modeled after the 2019 National Survey of Early
Care and Education (NSECE). By the end, you will be able to:

1. **Complete a full DER workflow** from data exploration through selective
   correction.
2. **Interpret design effects** in the context of early childhood policy
   research.
3. **Diagnose which parameters** are design-sensitive and explain why using
   the decomposition theorems.
4. **Apply selective correction** and compare it with naive and blanket
   alternatives.
5. **Report DER results** in a publication-ready format.

The analysis proceeds in seven stages: explore the survey design, run the
diagnostic, interpret fixed and random effects, verify the decomposition
theorems and conservation law, compare correction strategies, assess
sensitivity to the threshold, and format results for publication.


## Background: The NSECE Survey

The National Survey of Early Care and Education (NSECE), administered by the
U.S. Department of Health and Human Services, is the primary data source for
understanding the nation's childcare landscape. The 2019 wave surveyed
childcare providers across all 50 states plus the District of Columbia using a
complex multi-stage design:

- **Stratification** by state ensures adequate representation of each
  jurisdiction.
- **Cluster sampling** within states (primary sampling units, PSUs) captures
  geographic correlation.
- **Unequal selection probabilities** oversample small states and rare
  provider types.

The resulting survey weights can vary by a factor of 10 or more, producing
a substantial Kish design effect. The `nsece_demo` dataset is a synthetic
analog that preserves these structural features.

```{r load-data}
data(nsece_demo)
```

| Feature | Value |
|:--------|:------|
| Observations (N) | `r nsece_demo$N` |
| States (J) | `r nsece_demo$J` |
| Fixed effects (p) | `r nsece_demo$p` |
| Posterior draws | `r nrow(nsece_demo$draws)` |
| Parameters (d = p + J) | `r ncol(nsece_demo$draws)` |
| Family | `r nsece_demo$family` |


## The Research Question

The model examines a binary outcome: whether each childcare provider
participates in the state's Infant-Toddler (IT) quality improvement system.
Three fixed-effect covariates enter the hierarchical logistic regression:

- $\beta_0$ (intercept): overall log-odds of IT system participation.
- $\beta_1$ (`poverty_cwc`): provider-level poverty, centered within each
  state. This is a **within-cluster** covariate: it varies across providers
  within the same state.
- $\beta_2$ (`tiered_reim`): a binary indicator for whether the state
  operates a tiered reimbursement policy. This is a **between-cluster**
  covariate: it takes a single value for all providers in a given state.

In addition, the model includes 51 state-level random effects
$\theta_1, \ldots, \theta_{51}$, for a total of 54 parameters.

The DER analysis will reveal which of these 54 parameters are sensitive to
the survey design and which are protected by the hierarchical structure.

```{r param-types}
nsece_demo$param_types
```


## Exploring the Survey Design

Before running the DER diagnostic, it is instructive to examine the survey
design features that drive the design effect.

### Survey weight distribution

```{r weight-summary}
w <- nsece_demo$weights
cat(sprintf("Weight range:  [%.2f, %.2f]\n", min(w), max(w)))
cat(sprintf("Weight CV:      %.3f\n", sd(w) / mean(w)))
cat(sprintf("Kish DEFF:      %.3f\n", 1 + (sd(w) / mean(w))^2))
```

The coefficient of variation of the weights is substantial, yielding a Kish
design effect well above 1. This DEFF quantifies the information loss from
unequal weighting: an effective sample size of $N / \text{DEFF}$ rather than
$N$.

```{r fig-weights, fig.height = 4, fig.cap = "Distribution of survey weights in the NSECE-like dataset. The rightward tail indicates oversampled observations (e.g., small-state providers), which inflate the design effect."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  wt_df <- data.frame(weight = w)
  cv_w  <- sd(w) / mean(w)
  deff  <- 1 + cv_w^2

  ggplot(wt_df, aes(x = weight)) +
    geom_histogram(bins = 40, fill = pal["tier_ib"], colour = "white",
                   linewidth = 0.3, alpha = 0.85) +
    geom_vline(xintercept = mean(w), linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.8) +
    annotate("text", x = mean(w) * 1.15, y = Inf, vjust = 1.5,
             label = sprintf("mean = %.2f", mean(w)),
             colour = pal["threshold"], size = 4, fontface = "italic") +
    annotate("text", x = max(w) * 0.7, y = Inf, vjust = 3,
             label = sprintf("DEFF = %.2f", deff),
             colour = pal["tier_ia"], size = 5, fontface = "bold") +
    labs(x = "Survey weight", y = "Count",
         title = "Survey Weight Distribution") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
```

### PSU structure

Each state contains multiple PSUs. The number of PSUs per state affects the
precision of the cluster-robust variance estimate.

```{r psu-structure}
psu_per_state <- tapply(nsece_demo$psu, nsece_demo$group,
                        function(x) length(unique(x)))
cat(sprintf("PSUs per state: min = %d, median = %.0f, max = %d\n",
            min(psu_per_state), median(psu_per_state), max(psu_per_state)))
cat(sprintf("Total PSUs:     %d\n", length(unique(nsece_demo$psu))))
```

### Group sizes

```{r group-sizes}
n_j <- table(nsece_demo$group)
cat(sprintf("Providers per state: min = %d, median = %.0f, max = %d\n",
            min(n_j), median(n_j), max(n_j)))
```

The variation in group sizes means that the shrinkage factor $B_j$ will
differ across states: larger states experience less shrinkage (their data
dominate the prior), while smaller states are pulled more strongly toward
the grand mean.


## Running the DER Diagnostic

The full DER pipeline --- compute, classify, correct --- runs in a single
call to `der_diagnose()`:

```{r diagnose}
result <- der_diagnose(
  x           = nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  psu         = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types,
  tau         = 1.2
)
```

### Print output

The `print()` method provides a concise summary of the diagnosis:

```{r print-result}
result
```

Out of 54 parameters, only 1 exceeds the threshold $\tau = 1.2$. The
diagnostic immediately identifies the within-state poverty coefficient as the
sole parameter requiring design-based correction.

### Summary output

The `summary()` method displays the full classification table:

```{r summary-result}
summary(result)
```

### Model-level summary

The `glance()` method returns a one-row data frame suitable for reporting:

```{r glance-result}
glance.svyder(result)
```


## Understanding the Fixed Effects

The three fixed effects illustrate the three-tier classification in action.

```{r fe-der}
td <- tidy.svyder(result)
fe <- td[1:3, c("term", "estimate", "std.error", "der", "tier", "action")]
fe
```

The DER values span nearly two orders of magnitude:

- **$\beta_1$ (intercept)**: DER well below 1.0. This between-cluster
  parameter draws its information from between-state comparisons. The
  hierarchical prior absorbs much of the design effect through shrinkage.
  Classification: **Tier I-b** (protected; do not correct).

- **$\beta_2$ (poverty\_cwc)**: DER substantially above 1.0. This
  within-cluster covariate is identified from variation among providers
  *within* the same state. The random effects cannot absorb intra-cluster
  correlation at this level, so the full design effect passes through.
  Classification: **Tier I-a** (exposed; correct).

- **$\beta_3$ (tiered\_reim)**: DER well below 1.0. Like the intercept, this
  between-cluster policy indicator is identified from between-state
  differences and is substantially protected by hierarchical shrinkage.
  Classification: **Tier I-b** (protected; do not correct).

The 100-fold difference in DER between $\beta_2$ and the other fixed effects
is not an anomaly --- it is the predicted consequence of the decomposition
theorem. The information source determines design sensitivity.

### Visualizing the fixed effects

```{r fig-fe-profile, fig.height = 4, fig.cap = "DER profile for the three fixed effects. The within-state poverty coefficient (red, Tier I-a) is the only parameter exceeding the threshold (dashed purple line). The between-state parameters (blue, Tier I-b) are protected by hierarchical shrinkage."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  fe_plot <- data.frame(
    param = c("Intercept", "Poverty (CWC)", "Tiered Reim."),
    der   = td$der[1:3],
    tier  = td$tier[1:3]
  )
  fe_plot$param <- factor(fe_plot$param, levels = fe_plot$param)

  tier_colors <- c("I-a" = unname(pal["tier_ia"]),
                   "I-b" = unname(pal["tier_ib"]),
                   "II"  = unname(pal["tier_ii"]))

  ggplot(fe_plot, aes(x = param, y = der, colour = tier)) +
    geom_hline(yintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.7) +
    geom_point(size = 4) +
    geom_segment(aes(xend = param, y = 0, yend = der), linewidth = 0.8) +
    scale_colour_manual(values = tier_colors, name = "Tier") +
    annotate("text", x = 3.3, y = 1.2, label = expression(tau == 1.2),
             colour = pal["threshold"], size = 3.5, vjust = -0.5) +
    labs(x = NULL, y = "Design Effect Ratio",
         title = "Fixed Effect DER Values") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "bottom")
}
```


## Understanding the Random Effects

All 51 state random effects fall into Tier II (protected). Their DER values
are consistently below 1.0.

```{r re-der}
re <- td[td$tier == "II", ]
cat(sprintf("Random effect DER: min = %.4f, mean = %.4f, max = %.4f\n",
            min(re$der), mean(re$der), max(re$der)))
cat(sprintf("All below threshold: %s\n",
            ifelse(all(re$der < 1.2), "yes", "no")))
```

### Caterpillar plot of state DER values

```{r fig-re-caterpillar, fig.height = 6, fig.cap = "DER values for 51 state random effects, ordered by magnitude. All values fall well below the threshold (dashed purple line), confirming that hierarchical shrinkage protects the random effects from design-induced distortion. States with larger groups (weaker shrinkage) show slightly higher DER values."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  re_plot <- re[order(re$der), ]
  re_plot$rank <- seq_len(nrow(re_plot))
  state_labels <- gsub("theta\\[|\\]", "", re_plot$term)
  re_plot$state_label <- paste0("State ", state_labels)

  ggplot(re_plot, aes(x = der, y = reorder(state_label, der))) +
    geom_vline(xintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.7) +
    geom_vline(xintercept = 1.0, linetype = "dotted",
               colour = "grey50", linewidth = 0.5) +
    geom_point(colour = pal["tier_ii"], size = 1.8) +
    labs(x = "Design Effect Ratio", y = NULL,
         title = "State Random Effect DER Values") +
    theme_minimal(base_size = 10) +
    theme(panel.grid.minor = element_blank(),
          axis.text.y = element_text(size = 6))
}
```

### Shrinkage and DER: why random effects are protected

The DER for a random effect is approximately $B_j \times \text{DEFF}_j
\times \kappa_j(J)$, where $B_j$ is the group-specific shrinkage factor.
Since $B_j < 1$ for all groups, the prior directly attenuates the design
effect before it reaches the posterior.

```{r fig-B-vs-DER, fig.height = 4.5, fig.cap = "Relationship between the shrinkage factor B_j and DER for 51 state random effects. Each point is one state, colored by its per-group design effect (DEFF_j). States with larger B (less shrinkage) show higher DER, but even the least-shrunk states remain well below 1.0."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  re_diag <- data.frame(
    B_j    = result$B_j,
    deff_j = result$deff_j,
    der    = td$der[4:54]
  )

  ggplot(re_diag, aes(x = B_j, y = der, colour = deff_j)) +
    geom_point(size = 2.5, alpha = 0.8) +
    scale_colour_gradient(low = pal["tier_ib"], high = pal["tier_ia"],
                          name = expression(DEFF[j])) +
    geom_hline(yintercept = 1.0, linetype = "dotted",
               colour = "grey50", linewidth = 0.5) +
    labs(x = expression("Shrinkage factor" ~ B[j]),
         y = "Design Effect Ratio",
         title = "Shrinkage vs. DER for State Random Effects") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "right")
}
```


## Decomposition Analysis

The `der_decompose()` function reveals the constituent factors behind each
parameter's DER value.

```{r decompose}
decomp <- der_decompose(result)
```

### Fixed effect decomposition

For fixed effects, $\text{DER}_{\beta_k} = \text{DEFF} \times (1 - R_k)$,
where $R_k$ measures the fraction of identifying variation from
between-group differences:

```{r decompose-fe}
decomp[decomp$param_type != "re",
       c("param", "param_type", "der", "deff_mean", "R_k", "der_predicted")]
```

The within-cluster poverty covariate has $R_k \approx 0$: nearly all its
identifying variation comes from within-state differences, so the full DEFF
passes through. The between-cluster parameters have $R_k$ close to $B$,
confirming that the hierarchical structure absorbs much of the design effect.

### Random effect decomposition

For random effects, $\text{DER}_j = B_j \times \text{DEFF}_j \times
\kappa_j(J)$:

```{r decompose-re}
re_decomp <- decomp[decomp$param_type == "re",
                    c("param", "der", "B_mean", "deff_mean",
                      "kappa", "der_predicted")]
head(re_decomp, 10)
```

### Built-in decomposition plot

The built-in decomposition plot compares observed DER values against their
predicted values from the decomposition formulas:

```{r fig-decomposition, fig.cap = "DER decomposition: observed versus predicted values. Points near the diagonal indicate good agreement between the theoretical formulas and the full sandwich computation. The fixed effect outlier is the within-cluster poverty covariate."}
plot(result, type = "decomposition")
```


## The Conservation Law in Action

An elegant structural property of the DER framework is the conservation
law: in the balanced case,

$$\text{DER}_\mu + \text{DER}_\theta^{\text{cond}} = \text{DEFF}.$$

The hierarchical prior does not create or destroy design sensitivity. It
*redistributes* it. Every unit of design sensitivity removed from the random
effects reappears in the fixed effects (specifically the grand mean).

```{r conservation}
thm_check <- der_theorem_check(result)

conservation <- attr(thm_check, "conservation_law")
if (!is.null(conservation)) {
  cat(sprintf("DER_mu (intercept)       = %.4f\n", conservation$der_mu))
  cat(sprintf("DER_theta (mean RE)      = %.4f\n", conservation$der_theta_mean))
  cat(sprintf("Sum                      = %.4f\n", conservation$conservation_sum))
  cat(sprintf("Mean DEFF                = %.4f\n", conservation$deff_mean))
  cat(sprintf("Relative error           = %.4f\n", conservation$relative_error))
}
```

The conservation law holds approximately because the NSECE-like data has
unequal group sizes. The relative error quantifies the discrepancy; in
practice, errors below 0.20 indicate that the balanced-case approximation
provides useful intuition.

### Visualizing the design sensitivity budget

```{r fig-conservation, fig.height = 4, fig.cap = "The design sensitivity budget. The total DEFF is partitioned between the grand mean (DER_mu) and the random effects (mean DER_theta). The prior transfers sensitivity from the random effects to the grand mean, but the total budget is approximately conserved."}
if (requireNamespace("ggplot2", quietly = TRUE) && !is.null(conservation)) {
  library(ggplot2)

  budget_df <- data.frame(
    component = c("Grand mean\n(DER_mu)",
                  "Random effects\n(mean DER_theta)"),
    value     = c(conservation$der_mu, conservation$der_theta_mean)
  )
  budget_df$component <- factor(budget_df$component,
                                levels = budget_df$component)

  ggplot(budget_df, aes(x = "DEFF Budget", y = value, fill = component)) +
    geom_col(width = 0.5, colour = "white", linewidth = 0.5) +
    geom_hline(yintercept = conservation$deff_mean, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.8) +
    annotate("text", x = 1.4, y = conservation$deff_mean,
             label = sprintf("DEFF = %.2f", conservation$deff_mean),
             colour = pal["threshold"], size = 4, fontface = "italic",
             hjust = 0) +
    scale_fill_manual(values = c(pal["tier_ib"], pal["tier_ii"]),
                      name = "Component") +
    labs(x = NULL, y = "Design Effect Ratio",
         title = "Conservation of Design Sensitivity") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          legend.position = "right")
}
```


## Selective vs. Blanket Correction

The central practical contribution of the DER framework is **selective
correction**: adjusting only the flagged parameters while leaving the rest
untouched. This section compares selective correction with two alternatives:
naive (no correction) and blanket (correct everything).

### Extracting corrected draws

```{r corrected-draws}
draws_corrected <- as.matrix(result)
draws_original  <- result$original_draws

cat(sprintf("Draws matrix: %d draws x %d parameters\n",
            nrow(draws_corrected), ncol(draws_corrected)))
```

### Credible intervals for the flagged parameter

The within-state poverty coefficient ($\beta_2$) is the only flagged
parameter. Compare its naive and corrected credible intervals:

```{r ci-comparison}
# 95% credible intervals for the flagged parameter (column 2)
ci_naive     <- quantile(draws_original[, 2],  probs = c(0.025, 0.975))
ci_corrected <- quantile(draws_corrected[, 2], probs = c(0.025, 0.975))

width_naive     <- diff(ci_naive)
width_corrected <- diff(ci_corrected)
width_ratio     <- width_corrected / width_naive

cat(sprintf("Naive CI:     [%.3f, %.3f]  (width = %.3f)\n",
            ci_naive[1], ci_naive[2], width_naive))
cat(sprintf("Corrected CI: [%.3f, %.3f]  (width = %.3f)\n",
            ci_corrected[1], ci_corrected[2], width_corrected))
cat(sprintf("Width ratio:  %.1f%%\n", width_ratio * 100))
```

The correction widens the interval, reflecting the additional uncertainty
from the survey design that the naive model-based posterior failed to
capture.

### Unflagged parameters are unchanged

A critical property of selective correction is that unflagged parameters
retain their original draws:

```{r unchanged-check}
# Check that unflagged columns are identical
unflagged_cols <- which(result$scale_factors == 1.0)
all_identical  <- all(draws_corrected[, unflagged_cols] ==
                      draws_original[, unflagged_cols])
cat(sprintf("Unflagged parameters unchanged: %s (%d of %d parameters)\n",
            all_identical, length(unflagged_cols), ncol(draws_corrected)))
```

### The blanket correction problem

What would happen if we applied blanket correction to all 54 parameters? We
can assess this by examining the scale factors that *would* be applied:

```{r blanket-hypothetical}
# Hypothetical blanket scale factors: sqrt(DER) for all parameters
blanket_sf <- sqrt(td$der)

# For DER < 1, blanket correction would NARROW the posterior
n_narrowed   <- sum(td$der < 1.0)
n_widened    <- sum(td$der > 1.0)
n_unchanged  <- sum(td$der == 1.0)

cat(sprintf("Blanket correction would:\n"))
cat(sprintf("  Widen:   %d parameters (DER > 1)\n", n_widened))
cat(sprintf("  Narrow:  %d parameters (DER < 1)\n", n_narrowed))

# Worst case: how much would the most-protected RE shrink?
min_sf <- min(blanket_sf)
cat(sprintf("\nSmallest scale factor: %.4f\n", min_sf))
cat(sprintf("That RE's CI would shrink to %.1f%% of original width\n",
            min_sf * 100))
```

Blanket correction would inappropriately narrow the credible intervals of
53 parameters. For the most-protected random effects, the interval would
collapse to a fraction of its original width, destroying the precision gains
from hierarchical shrinkage.

### Built-in comparison plot

```{r fig-comparison, fig.height = 3.5, fig.cap = "Credible interval comparison for the flagged parameter (poverty_cwc). Gray shows the naive model-based interval; the colored interval shows the DER-corrected interval. The correction appropriately widens the interval to account for the survey design."}
plot(result, type = "comparison")
```

### Custom CI comparison across all fixed effects

```{r fig-ci-custom, fig.height = 4, fig.cap = "Fixed-effect credible intervals: naive (gray) vs. selectively corrected (colored). Only the within-state poverty coefficient changes; the between-state parameters are left untouched."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  fe_names <- c("Intercept", "Poverty (CWC)", "Tiered Reim.")
  ci_list  <- lapply(1:3, function(k) {
    naive_ci <- quantile(draws_original[, k],  probs = c(0.025, 0.50, 0.975))
    corr_ci  <- quantile(draws_corrected[, k], probs = c(0.025, 0.50, 0.975))
    data.frame(
      param  = fe_names[k],
      type   = c("Naive", "Corrected"),
      lo     = c(naive_ci[1], corr_ci[1]),
      mid    = c(naive_ci[2], corr_ci[2]),
      hi     = c(naive_ci[3], corr_ci[3]),
      stringsAsFactors = FALSE
    )
  })
  ci_df <- do.call(rbind, ci_list)
  ci_df$param <- factor(ci_df$param, levels = rev(fe_names))
  ci_df$type  <- factor(ci_df$type, levels = c("Naive", "Corrected"))

  ggplot(ci_df, aes(x = mid, y = param, colour = type)) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = "grey60", linewidth = 0.5) +
    geom_pointrange(aes(xmin = lo, xmax = hi),
                    position = position_dodge(width = 0.4),
                    size = 0.5, linewidth = 0.7) +
    scale_colour_manual(
      values = c(Naive = "grey55", Corrected = pal["tier_ia"]),
      name = "Estimate"
    ) +
    labs(x = "Coefficient value", y = NULL,
         title = "Fixed Effect Credible Intervals: Naive vs. Corrected") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "bottom")
}
```

### Computational cost comparison

Selective correction operates on a $|S| \times |S|$ submatrix, where $|S|$
is the number of flagged parameters. Blanket correction operates on the full
$d \times d$ matrix.

```{r cost-comparison}
d <- ncol(draws_corrected)
n_flagged <- sum(result$classification$flagged)

cat(sprintf("Selective: O(%d^3) = O(%d) operations\n",
            n_flagged, n_flagged^3))
cat(sprintf("Blanket:   O(%d^3) = O(%d) operations\n",
            d, d^3))
cat(sprintf("Speedup:   %.0fx\n", d^3 / max(n_flagged^3, 1)))
```


## Sensitivity and Robustness

The classification depends on the threshold $\tau$. A robust diagnostic
should not change its recommendation dramatically when $\tau$ varies over a
reasonable range.

```{r sensitivity}
sens <- der_sensitivity(result, tau_range = seq(0.5, 3.0, by = 0.1))
sens[, c("tau", "n_flagged", "pct_flagged")]
```

The within-state poverty coefficient is flagged at every threshold from 0.5
up to approximately its DER value. No other parameter is flagged at any
threshold above 1.0. This stability confirms a **sharp classification
boundary**: there is a clean separation between the design-sensitive
within-cluster covariate and all other parameters.

```{r fig-sensitivity, fig.height = 4, fig.cap = "Sensitivity of the DER classification to the threshold tau. The number of flagged parameters drops from 1 to 0 at the DER value of the poverty coefficient (~2.6). The stable plateau at 1 flagged parameter across a wide range of thresholds indicates a robust classification."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  ggplot(sens, aes(x = tau, y = n_flagged)) +
    geom_step(colour = pal["tier_ia"], linewidth = 0.9) +
    geom_point(colour = pal["tier_ia"], size = 2) +
    geom_vline(xintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.6) +
    annotate("text", x = 1.2, y = max(sens$n_flagged) + 0.3,
             label = expression(tau == 1.2 ~ "(default)"),
             colour = pal["threshold"], size = 3.5, hjust = -0.05) +
    scale_y_continuous(breaks = seq(0, max(sens$n_flagged) + 1, by = 1)) +
    labs(x = expression("Threshold" ~ tau), y = "Number flagged",
         title = "Sensitivity of Classification to Threshold") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
```


## Cross-Clustering Comparison

The DER depends on the choice of clustering (PSU) definition. The
`der_compare()` function evaluates how DER changes across different
clustering schemes. Here we compare state-level clustering (51 clusters)
with PSU-level clustering (finer partitions within states):

```{r compare}
comp <- der_compare(
  nsece_demo$draws,
  clusters = list(
    state = nsece_demo$group,
    psu   = nsece_demo$psu
  ),
  y = nsece_demo$y, X = nsece_demo$X,
  group = nsece_demo$group, weights = nsece_demo$weights,
  family = "binomial", sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types
)
```

### Comparing DER across clustering

```{r compare-fe}
# Fixed effects comparison
fe_comp <- comp[comp$param %in% paste0("beta[", 1:3, "]"), ]
fe_wide <- reshape(fe_comp, idvar = "param", timevar = "cluster_name",
                   direction = "wide")
fe_wide
```

```{r fig-compare, fig.height = 4.5, fig.cap = "Cross-clustering comparison of DER values for the three fixed effects. PSU-level clustering (which captures finer-scale design effects) generally yields DER values equal to or higher than state-level clustering. Both clustering schemes flag the same parameter (poverty_cwc)."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  fe_comp$param_label <- c("Intercept", "Poverty (CWC)", "Tiered Reim.")[
    as.integer(gsub("beta\\[(\\d+)\\]", "\\1", fe_comp$param))
  ]

  ggplot(fe_comp, aes(x = param_label, y = der, fill = cluster_name)) +
    geom_col(position = position_dodge(width = 0.6), width = 0.5,
             colour = "white", linewidth = 0.3) +
    geom_hline(yintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.7) +
    scale_fill_manual(
      values = c(state = pal["tier_ib"], psu = pal["tier_ia"]),
      name = "Clustering"
    ) +
    labs(x = NULL, y = "Design Effect Ratio",
         title = "DER by Clustering Definition") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "bottom")
}
```

In general, finer clustering (PSU level) captures more of the design effect
and may produce higher DER values for within-cluster covariates. The
practical recommendation is to use the finest available clustering level in
the DER computation, as this provides the most conservative assessment of
design sensitivity.


## Theorem Verification

The `der_theorem_check()` function compares empirical DER values against the
closed-form predictions from Theorems 1 and 2.

```{r theorem-check}
thm <- der_theorem_check(result)
```

### Fixed effects: Theorem 1

```{r thm-fe}
fe_thm <- thm[thm$param_type != "re",
              c("param", "der_empirical", "der_theorem1",
                "relative_error", "theorem_used")]
fe_thm
```

### Random effects: Theorem 2

```{r thm-re}
re_thm <- thm[thm$param_type == "re",
              c("param", "der_empirical", "der_theorem2", "relative_error")]
head(re_thm, 10)

cat(sprintf("\nMean relative error (RE): %.4f\n",
            mean(re_thm$relative_error, na.rm = TRUE)))
```

The fixed-effect predictions are typically accurate within a few percent.
Random-effect predictions may show larger discrepancies in non-conjugate
models (e.g., logistic) or with substantial imbalance, but the qualitative
ordering is always preserved.


## Reporting DER Results for Publication

### Tidy summary table

The `tidy()` output provides a publication-ready table:

```{r tidy-output}
td_report <- tidy.svyder(result)
td_report_fe <- td_report[1:3, c("term", "estimate", "std.error",
                                  "der", "tier", "action")]
td_report_fe$estimate  <- round(td_report_fe$estimate, 3)
td_report_fe$std.error <- round(td_report_fe$std.error, 3)
td_report_fe$der       <- round(td_report_fe$der, 3)
td_report_fe
```

### Key numbers for the methods section

```{r key-numbers}
gl <- glance.svyder(result)

cat("Key statistics for reporting:\n")
cat(sprintf("  Mean DEFF:               %.2f\n", gl$mean_deff))
cat(sprintf("  Mean shrinkage (B):      %.3f\n", gl$mean_B))
cat(sprintf("  Parameters flagged:      %d / %d (%.1f%%)\n",
            gl$n_flagged, gl$n_params, gl$pct_flagged))
cat(sprintf("  Threshold (tau):         %.1f\n", gl$tau))
cat(sprintf("  DER range:               [%.3f, %.3f]\n",
            gl$der_min, gl$der_max))
cat(sprintf("  CI width ratio (beta2):  %.1f%%\n", width_ratio * 100))
```

### Template paragraph for the methods section

Based on the analysis above, a methods section might include language such
as:

> We applied the DER diagnostic framework (Lee, 2026) to assess
> design sensitivity for each model parameter. The mean Kish design effect
> was DEFF = [value], reflecting substantial variation in survey weights
> (CV_w = [value]). Of 54 parameters (3 fixed effects + 51 state random
> effects), only 1 exceeded the classification threshold tau = 1.2: the
> within-state poverty coefficient (DER = [value]). This parameter's
> 95% credible interval was widened by [value]% after selective
> correction. The remaining 53 parameters, including all state random
> effects (mean DER = [value]) and between-state fixed effects
> (DER < [value]), were left unchanged. The conservation law was
> verified with relative error [value].


## Full DER Profile

For completeness, the built-in profile plot shows all 54 parameters:

```{r fig-profile-full, fig.height = 5, fig.cap = "Full DER profile across all 54 parameters. Each point is one parameter, colored by tier classification. The dashed purple line marks the threshold tau = 1.2. The single outlier (Tier I-a, red) is the within-state poverty coefficient. All random effects (green, Tier II) and between-state fixed effects (blue, Tier I-b) fall well below the threshold."}
plot(result, type = "profile")
```


## Summary and Recommendations

This case study demonstrated the full DER diagnostic workflow on a realistic
survey dataset. The key findings are:

1. **Selective correction is precise.** Only 1 of 54 parameters required
   correction. Blanket correction would have inappropriately narrowed 53
   parameters, including all random effects.

2. **The information source determines sensitivity.** The within-cluster
   poverty covariate inherits the full design effect ($\text{DER} \approx
   \text{DEFF}$) because the random effects cannot absorb intra-cluster
   correlation at the individual level. Between-cluster parameters are
   protected by hierarchical shrinkage.

3. **The conservation law constrains the system.** Protecting the random
   effects through shrinkage necessarily exposes the grand mean to a larger
   design effect. The total design sensitivity budget is approximately
   conserved.

4. **The classification is robust.** The within-state poverty coefficient is
   flagged at every threshold from 0.5 to its DER value, and no other
   parameter is flagged at any reasonable threshold.

### General workflow recommendations

| Step | Function | Purpose |
|:-----|:---------|:--------|
| 1. Diagnose | `der_diagnose()` | Full pipeline in one call |
| 2. Inspect | `print()`, `summary()`, `glance()` | Assess overall results |
| 3. Visualize | `plot(type = "profile")` | Identify flagged parameters |
| 4. Explain | `der_decompose()` | Understand why each parameter has its DER |
| 5. Verify | `der_theorem_check()` | Confirm theoretical predictions |
| 6. Sensitivity | `der_sensitivity()` | Check robustness to threshold |
| 7. Report | `tidy()`, `as.matrix()` | Publication-ready output |

### When DER matters most

- Large Kish DEFF (> 2), indicating substantial weight variation.
- Within-cluster covariates in the model (these inherit the full DEFF).
- Policy analyses where both within-cluster and between-cluster covariates
  are of substantive interest.

### When DER matters least

- Balanced designs with equal weights (DEFF $\approx$ 1).
- Models with only between-cluster or random-effect parameters.
- Exploratory analyses where exact coverage is less critical.

For the mathematical theory underlying this analysis, see
`vignette("understanding-der", package = "svyder")`.
For a quick-start guide, see
`vignette("getting-started", package = "svyder")`.
