---
title: "Choosing the Variance Target"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  The one modeling decision the Design Effect Ratio forces you to make: what is
  the sandwich aggregation unit? This vignette contrasts the two canonical
  targets (design-PSU vs model-group) on the bundled NSECE-style demo, walks
  through der_compare(), and explains how strata, the finite-cluster correction,
  the cluster universe, and the weight convention all become part of the
  variance target you declare.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing the Variance Target}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", out.width = "85%",
                      warning = FALSE, message = FALSE)
library(svyder)
# Consistent colour palette across all svyder vignettes (colour-blind aware)
svyder_pal <- c(
  `I-a`  = "#D6604D",  # within-group fixed effects (survey-dominated)
  `I-b`  = "#4393C3",  # between-group fixed effects (shielded)  [brand blue]
  `II`   = "#5AAE61",  # random effects (shrinkage-protected)
  `III`  = "#9E9E9E",  # hyperparameters (excluded)
  flag   = "#B2182B",  # flagged
  target = "#1A1A1A"   # DER = 1 / threshold reference lines
)
```

## Overview

The Design Effect Ratio compares two variances: a design-based *sandwich*
variance in the numerator and the model-based MCMC posterior variance in the
denominator. The denominator is fixed the moment you fit the model. The
numerator is not: it depends on *how you aggregate the score contributions into
clusters*, and that choice is yours to make. This vignette is written for
practitioners who have run a first diagnostic (see
[Getting Started](getting-started.html)) and now need to decide **which variance
target to declare**.

The central message is that the aggregation unit is part of the *estimand*, not
an implementation detail. Two defensible targets --- the survey's design PSUs
and the model's grouping factor --- answer two different questions about the
same fitted model, and they can disagree materially about which parameters are
survey-sensitive. svyder makes you name the target explicitly so that the
diagnostic is reproducible and its meaning is unambiguous.

We cover, in order:

- **Why `cluster` has no default** and the two canonical choices.
- **Two targets on the demo**: computing and reading both.
- **`der_compare()`**: the recommended single-call comparison.
- **Strata and the finite-cluster correction** for real stratified designs.
- **The cluster universe** (`cluster_strata=`) for selected-but-empty PSUs.
- **Weight conventions** (`normalize=`) and why random-effect DERs are not
  invariant to them.
- **Recommended practice** for reporting.

This vignette is about the *decision* and its mechanics. It does not derive the
sandwich estimator or explain the classification tiers --- for the mathematics
of the bread and meat see `vignette("theory-sandwich")`, and for the
compute --> classify --> correct workflow see
[The Compute--Classify--Correct Pipeline](pipeline.html).

> **Key insight.** The DER is only interpretable relative to a *declared
> variance target*. Changing the aggregation unit does not "fix" or "break" the
> diagnostic; it changes the question being asked. Report the target you chose.

```{r fit-both}
d <- nsece_demo

# Design-based target: aggregate scores to the survey's sampling clusters (PSUs)
res_psu <- der_compute(
  d$draws,
  y           = d$y,
  X           = d$X,
  group       = d$group,
  weights     = d$weights,
  cluster     = d$psu,           # <- the aggregation unit is the design PSU
  family      = "binomial",
  sigma_theta = d$sigma_theta,
  param_types = d$param_types
)

# Model-aligned target: aggregate scores to the model's grouping factor (states)
res_group <- der_compute(
  d$draws,
  y           = d$y,
  X           = d$X,
  group       = d$group,
  weights     = d$weights,
  cluster     = d$group,         # <- the aggregation unit is the model group
  family      = "binomial",
  sigma_theta = d$sigma_theta,
  param_types = d$param_types
)
```

## Why there is no default `cluster`

Most survey-variance software has a natural default: aggregate to the design's
primary sampling units. svyder deliberately refuses to guess. The meat of the
sandwich is a sum of *cluster* score totals,

$$
J_c \;=\; \sum_{h}\frac{C_h}{C_h-1}\sum_{c\in h}
          (t_{hc}-\bar t_h)(t_{hc}-\bar t_h)^{\!\top},
\qquad
t_{hc}=\sum_{i\in c} s_i,
$$

so the value of $J_c$ --- and therefore every DER --- depends on what you call a
cluster $c$. There is no way to compute the numerator without committing to an
aggregation unit, and different units encode genuinely different variance
targets. Requiring the argument keeps that commitment visible in your code.

Concretely, omitting `cluster` is an error, not a silent fallback. A call like

```r
der_compute(d$draws, y = d$y, X = d$X, group = d$group,
            weights = d$weights, family = "binomial",
            sigma_theta = d$sigma_theta)
```

stops with a message that spells out the decision you must make:

```
'cluster' must be supplied explicitly -- the sandwich aggregation unit is part
of the variance-target definition.
  * design-based target: cluster = <design PSU ids>
  * model-aligned target: cluster = group
See der_compare() to report both.
```

The two canonical choices named in that message are:

- **Design-PSU target** --- `cluster = nsece_demo$psu`. Aggregate to the
  survey's sampling clusters. This asks: *given the actual sampling design, how
  much sampling variability does each parameter carry?* It is the target most
  faithful to the survey's variance structure and typically the more
  conservative of the two for within-group effects.
- **Model-group target** --- `cluster = nsece_demo$group`. Aggregate to the
  model's grouping factor (here, the 51 states). This asks: *how much residual
  design variability remains after the random effects have already absorbed
  between-group heterogeneity?* Because the model's own clustering has soaked up
  much of the between-cluster signal, this target is usually smaller for random
  effects.

> **Note.** These are not "right" and "wrong" targets. The design-PSU target
> answers a question about the sampling process; the model-group target answers
> a question about what the fitted hierarchy has left unmodeled. A careful
> analysis reports both.

For reference, the demo has `r length(unique(d$psu))` design PSUs nested within
`r d$J` model groups, so the two aggregation units are genuinely different
partitions of the same `r format(d$N, big.mark = ",")` observations.

## Two targets on the demo

With both diagnostics fitted above, we classify each at the default threshold
$\tau = 1.2$ and read the summaries. Start with the design-PSU target.

```{r classify-psu}
cls_psu <- der_classify(res_psu, tau = 1.2, verbose = FALSE)
cls_psu
```

Under the **design-PSU target** the DER range is
$[`r round(min(der_decompose(res_psu)$der), 3)`,\,
   `r round(max(der_decompose(res_psu)$der), 3)`]$
and **30 of 54** parameters are flagged (55.6%). The within-group poverty
coefficient `beta[2]` carries the headline design sensitivity: its DER is
**2.689** (Tier I-a, flagged), while the two between-group fixed effects sit well
below threshold (`beta[1]` at 0.263 and `beta[3]` at 0.343, both Tier I-b,
retained). Among the random effects, **29 of 51** states are flagged, with a
median random-effect DER of 1.311 --- under the design PSUs, per-state variation
is exposed to the sampling design rather than shielded by shrinkage.

Now the model-group target.

```{r classify-group}
cls_group <- der_classify(res_group, tau = 1.2, verbose = FALSE)
cls_group
```

Under the **model-group target** the DER range widens to
$[`r round(min(der_decompose(res_group)$der), 3)`,\,
   `r round(max(der_decompose(res_group)$der), 3)`]$
but **only 25 of 54** parameters are flagged. The poverty coefficient `beta[2]`
remains flagged --- its DER is now **1.893**, still above $\tau$ but noticeably
smaller than under the design PSUs --- while the between-group fixed effects stay
retained (0.424 and 0.447). The random-effect picture shifts the most: **24 of
51** states are flagged and the median random-effect DER falls to **0.956**,
below 1. Once the state random effects absorb between-state heterogeneity, most
states no longer read as design-sensitive.

> **Key insight.** The one parameter that is flagged under *both* targets ---
> the within-group poverty coefficient `beta[2]` --- is the robust finding. Its
> flag does not depend on how you aggregate, so it is the parameter you can most
> confidently say needs a design-aware interval. Parameters that flip between
> targets (many of the random effects here) are target-dependent and should be
> reported as such.

## `der_compare()`: one call, both targets

Rather than fit and read two objects by hand, `der_compare()` takes a single
fitted result plus a *named list* of cluster-id vectors and returns the DER for
every parameter under every target. This is the recommended way to make the
comparison, because it guarantees identical draws, data, weights, and threshold
across targets --- the only thing that varies is the aggregation unit.

```{r der-compare}
cmp <- der_compare(
  res_psu,
  clusters = list(
    design_psu  = nsece_demo$psu,
    model_group = nsece_demo$group
  )
)
head(cmp, 6)
```

The result is a long data frame with one row per parameter x target combination
(columns `param`, `cluster_name`, `der`). For a compact side-by-side view of the
fixed effects, reshape to wide:

```{r der-compare-wide}
fe <- cmp[grepl("^beta", cmp$param), ]
reshape(fe, idvar = "param", timevar = "cluster_name", direction = "wide")
```

Reading across the row for `beta[2]`, the poverty coefficient carries a DER of
**2.689** under the design PSUs versus **1.893** under the model group --- the
same qualitative conclusion (correct it) but a different magnitude of design
inflation. The between-group coefficients `beta[1]` and `beta[3]` stay far below
threshold under either target, exactly as Theorem 1 predicts for effects whose
identifying variation is almost entirely between groups.

A quick visual comparison of the full profile makes the target-dependence of the
random effects easy to see:

```{r fig-compare, echo = TRUE, fig.height = 4.5, fig.cap = "DER under the two targets. Points above the dashed threshold line are flagged. The within-group poverty coefficient (beta[2]) sits above the line under both targets; many random effects cross the line only under the design-PSU target."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  cmp$is_beta <- grepl("^beta", cmp$param)
  ggplot(cmp, aes(x = cluster_name, y = der, group = param,
                  colour = is_beta)) +
    geom_hline(yintercept = 1.2, linetype = "dashed",
               colour = svyder_pal[["target"]]) +
    geom_line(alpha = 0.4, colour = "grey70") +
    geom_point(size = 2) +
    scale_colour_manual(values = c(`TRUE` = svyder_pal[["I-a"]],
                                   `FALSE` = svyder_pal[["II"]]),
                        labels = c(`TRUE` = "fixed effect",
                                   `FALSE` = "random effect"),
                        name = NULL) +
    scale_y_continuous(trans = "log10") +
    labs(x = "Variance target (aggregation unit)",
         y = "DER (log scale)",
         caption = "Dashed line: threshold tau = 1.2.") +
    theme_minimal()
}
```

## Strata and the finite-cluster correction

Real stratified designs add two refinements to the meat, both surfaced as
arguments to `der_compute()`:

- **`strata=`** --- a stratum id per observation. When supplied, cluster score
  totals are centered *within stratum* before the outer products are summed, so
  that between-stratum differences (which are by design, not sampling error) do
  not inflate the variance.
- **`center_meat = TRUE`** (default) --- subtract the stratum mean score total
  $\bar t_h$ from each cluster total, as in the formula above.
- **`df_correct = TRUE`** (default) --- apply Binder's finite-cluster factor
  $C_h/(C_h-1)$ per stratum, correcting for the fact that a stratum with $C_h$
  sampled clusters estimates its own mean.

Setting `center_meat = FALSE, df_correct = FALSE, strata = NULL` reproduces the
version-1 (uncentered, uncorrected) meat, which is occasionally useful for
matching legacy output.

The bundled demo has **no strata slot** --- it is a single-stratum synthetic
design --- so these arguments have nothing to center or correct here and we leave
them at their defaults. They matter for genuine multi-stratum surveys such as the
real NSECE (415 PSUs in 30 strata; see the [Case Study](case-study-nsece.html)).
The call for a stratified design would read:

```{r strata-shape, eval = FALSE}
# Illustrative: a real design would supply a stratum id per observation.
der_compute(
  draws, y = y, X = X, group = group, weights = weights,
  cluster     = psu,          # sampling clusters
  strata      = stratum_id,   # <- stratum per observation
  center_meat = TRUE,         # subtract within-stratum mean cluster total
  df_correct  = TRUE,         # Binder C_h / (C_h - 1) per stratum
  family = "binomial", sigma_theta = sigma_theta
)
```

> **Note.** Stratum centering and the $C_h/(C_h-1)$ factor are properties of the
> *variance target*, not tuning knobs. They encode how the survey was stratified
> and are recorded in the fitted object's `$target` so that the target is fully
> self-describing.

## The cluster universe

A subtlety arises in multi-stage designs when a selected PSU contributes *no*
observations to the model's estimation sample --- for example, a second-stage
sample that came back empty. Such a PSU still carries design information: it is a
cluster whose score total is exactly zero, and dropping it silently would
understate the between-cluster variance.

The `cluster_strata=` argument declares the **full cluster universe** so that
selected-but-empty PSUs enter the meat as zero score-total clusters. It requires
integer cluster ids running over `1:length(cluster_strata)` and requires
`strata` (each entry gives the stratum of the corresponding cluster id).

The demo needs none of this: every realized cluster has at least one
observation, so the cluster universe is exactly the set of clusters that appear
in the data.

```{r cluster-universe-check}
# Every PSU in the demo contributes >= 1 observation, so no empty clusters exist.
all(table(d$psu) >= 1)
```

Accordingly we leave `cluster_strata = NULL` (its default). Declare it only when
your design's frame includes clusters that were selected but returned empty
samples, so the finite-cluster factor counts *all selected* clusters, not just
the non-empty ones.

## Weight conventions

The final component of the target is how survey weights are scaled before they
enter the score. The `normalize` argument offers three conventions:

- `"unit_mean"` (default) --- rescale weights to mean 1 across the whole sample.
- `"group_size"` --- rescale weights within each group to sum to the group's
  sample size.
- `"none"` --- use the raw weights as supplied.

The chosen convention is recorded in the fitted object's `$target`, because ---
unlike a classical design-based point estimate --- **random-effect DERs are not
invariant to weight rescaling**. Rescaling changes the relative contribution of
observations to each group's score total, which changes the meat, which changes
the DER. The convention is therefore part of the variance target you declare,
not a harmless normalization.

On *this* demo, however, all three conventions coincide, because the demo's
weights are already normalized within group to mean 1:

```{r normalize-invariance}
der_at <- function(nrm) {
  r <- der_compute(d$draws, y = d$y, X = d$X, group = d$group,
                   weights = d$weights, cluster = d$psu, family = "binomial",
                   sigma_theta = d$sigma_theta, param_types = d$param_types,
                   normalize = nrm)
  der_decompose(r)$der
}
der_um <- der_at("unit_mean")
der_gs <- der_at("group_size")
der_no <- der_at("none")

# The three DER vectors are identical up to floating-point noise on this demo.
max(abs(der_um - der_gs))
max(abs(der_um - der_no))
```

The differences are at the level of machine precision, so we cannot use the demo
to illustrate weight-convention sensitivity --- doing so would be misleading.

Instead, the effect is real and can be *dramatic* on genuine survey weights. In
the paper's analysis of the **real NSECE 2019 data** (not this demo), the weight
convention reverses the ordering of the two targets:

| Target (real NSECE) | Convention | $\beta_1$ poverty DER |
|:--------------------|:-----------|----------------------:|
| Model-group         | unit-mean    | **4.59** |
| Model-group         | within-group | **2.70** |
| Design-PSU          | unit-mean    | **3.55** |
| Design-PSU          | within-group | **4.70** |

Under the unit-mean convention the model-group target reports the larger DER
(4.59 vs 3.55); under the within-group convention the ordering *flips* and the
design-PSU target reports the larger DER (4.70 vs 2.70). The paper's poverty
coefficient stays flagged under every target x convention combination --- that
robustness is what makes it a credible headline --- but its *magnitude*, and even
which target looks more conservative, depends on the convention.

> **Key insight.** Report the weight convention alongside the aggregation unit.
> The two together define the variance target; the numbers above are the paper's
> real-data demonstration that leaving either implicit can silently reverse your
> conclusion about which target is more conservative. (These are real-NSECE
> figures for attribution only --- they are *not* reproducible from
> `nsece_demo`, whose conventions coincide.)

## Recommended practice

Because the DER is defined relative to a declared target, two habits keep the
diagnostic honest and reproducible.

1. **Report both canonical targets.** Run `der_compare()` with
   `list(design_psu = ..., model_group = ...)` and present the DER of your key
   parameters under both. Parameters flagged under both targets (like `beta[2]`
   here) are the robust findings; parameters that flip are target-dependent and
   should be flagged as such rather than reported as if the choice were settled.

2. **Declare the weight convention and the aggregation unit explicitly.** State
   in your methods which `normalize` convention and which `cluster` you used ---
   ideally by showing the code. The fitted object already records this for you:

```{r target-record}
res_psu$target[c("cluster_is_group", "strata_n", "center", "df_correct",
                 "normalize", "cluster_universe_declared")]
```

Everything needed to reconstruct the variance target --- whether the cluster was
the model group, how many strata, whether the meat was centered and
DF-corrected, the weight convention, and whether a full cluster universe was
declared --- travels with the object. A reader who has the code and the recorded
target can reproduce the diagnostic exactly.

> **Note.** "Which target?" has no universal answer, but it always has a
> *reportable* answer. The design-PSU target is the more faithful description of
> the sampling variance; the model-group target describes what the fitted
> hierarchy leaves unmodeled. Naming your choice --- and showing the other ---
> is what makes the DER interpretable.

## What's next?

- To fold the chosen target into the full workflow --- classify at a threshold,
  correct only the flagged parameters, and verify the unflagged draws are
  untouched --- read
  [The Compute--Classify--Correct Pipeline](pipeline.html).
- To see both targets applied end-to-end on the demo, with fixed- and
  random-effect interpretation and a publication-ready write-up, read the
  [Case Study: NSECE-Style Survey Data](case-study-nsece.html).
- For the mathematics of the bread and meat --- why the aggregation unit,
  stratum centering, the $C_h/(C_h-1)$ factor, the cluster universe, and the
  weight convention enter the sandwich the way they do --- see
  `vignette("theory-sandwich")`.
