---
title: "Decomposition Theorems: Why DER Differs Across Parameters"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A deep dive into the DER decomposition theorems: understand why different
  parameters have different design sensitivities, verify the closed-form
  approximations against empirical values, and explore the conservation law
  that links hierarchical shrinkage to survey design effects.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decomposition Theorems: Why DER Differs Across Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(svyder)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
set.seed(42)

# Project colour palette (consistent across all svyder vignettes)
pal <- c(
  tier_ia    = "#D6604D",
  tier_ib    = "#4393C3",
  tier_ii    = "#1B7837",
  threshold  = "#762A83"
)
```

## Overview

The Design Effect Ratio (DER) measures how much each parameter's posterior
variance needs to change when the complex survey design is accounted for.
In a hierarchical logistic regression fitted to the NSECE, DER values
range from near zero for well-shrunk random effects to over 2.5 for a
within-cluster covariate --- a difference of two orders of magnitude. Why
does a single survey design produce such heterogeneous effects across
parameters?

The answer lies in the **decomposition theorems**, which break each DER
into a product of interpretable factors. These theorems provide the
theoretical foundation for the three-tier classification and explain
precisely which parameters need correction.

By the end of this vignette you will be able to:

1. **Decompose** DER into four constituent factors: $\text{DEFF}$, $B$,
   $R_k$, and $\kappa(J)$.
2. **Apply Theorem 1** to predict DER for fixed effects based on the
   protection factor $R_k$.
3. **Apply Theorem 2** to predict DER for random effects based on the
   shrinkage factor $B$.
4. **Verify the conservation law**: $\text{DER}_\mu + \text{DER}_{\theta}^{\text{cond}} = \text{DEFF}$.
5. **Interpret** decomposition components to understand why each parameter
   has its observed DER value.


## The Big Picture: From One Number to Many

The classical Kish design effect provides a single summary of the survey
design's impact:

$$
\text{DEFF} = 1 + \text{CV}_w^2,
$$

where $\text{CV}_w$ is the coefficient of variation of the survey weights.
For the NSECE-like data, this gives $\text{DEFF} \approx 3.76$. This
number tells us that the survey carries the information content of an SRS
roughly one-quarter the nominal size.

But in a hierarchical model, the story is more nuanced. Let us compute
the DER for every parameter in the NSECE dataset and observe the range:

```{r big-picture}
data(nsece_demo)

result <- der_diagnose(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  psu         = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types
)

# DER range across all 54 parameters
td <- tidy.svyder(result)
cat(sprintf("DER range: [%.4f, %.4f]\n", min(td$der), max(td$der)))
cat(sprintf("DER ratio (max/min): %.0fx\n", max(td$der) / min(td$der)))
```

A single $\text{DEFF}$ cannot capture this heterogeneity. Parameters
differ by a factor of 100 or more in their design sensitivity, and this
variation is entirely predictable from their information structure. The
decomposition theorems explain how.


## The Four Components

The DER decomposition expresses each parameter's design sensitivity as a
product of four factors. Each factor captures a different mechanism by
which the survey design interacts with the hierarchical model:

| Factor | Symbol | Range | What it measures |
|:-------|:------:|:-----:|:-----------------|
| Kish design effect | $\text{DEFF}$ | $[1, \infty)$ | Overall weight variability |
| Shrinkage factor | $B$ | $[0, 1]$ | Prior strength relative to data |
| Protection factor | $R_k$ | $[0, B]$ | Between-cluster information share (FE only) |
| Finite-group correction | $\kappa(J)$ | $[0, 1)$ | Adjustment for finite $J$ (RE only) |

The key insight is that the raw design effect $\text{DEFF}$ is
*modulated* --- sometimes dramatically --- by the hierarchical model
structure. The decomposition reveals exactly how this modulation works.


## Theorem 1: Fixed Effects

**Theorem 1** governs fixed-effect parameters $\beta_k$ in the
hierarchical model:

$$
\text{DER}_{\beta_k} = \text{DEFF} \times (1 - R_k),
$$

where $R_k \in [0, B]$ is the **protection factor** --- the fraction of
identifying information for covariate $k$ that comes from between-cluster
variation. Two important special cases emerge:

- **Within-cluster covariates** (e.g., individual-level poverty centered
  within groups): $R_k \approx 0$, so $\text{DER} \approx \text{DEFF}$.
  The full design effect passes through because within-cluster variation
  is orthogonal to the random effects.

- **Between-cluster covariates** (e.g., state-level policy indicators):
  $R_k \approx B$, so $\text{DER} \approx \text{DEFF} \times (1 - B)$.
  The random effects absorb between-cluster correlation, shielding this
  parameter from much of the design effect.

### Worked example: NSECE fixed effects

Let us run `der_decompose()` on the NSECE result and examine the fixed
effects:

```{r thm1-decompose}
decomp <- der_decompose(result)

# Fixed effects only
fe_decomp <- decomp[decomp$param_type != "re", ]
fe_decomp[, c("param", "param_type", "der", "deff_mean", "R_k", "der_predicted")]
```

The decomposition reveals the mechanism behind the three-tier structure:

- **beta[1]** (intercept, `fe_between`): Identified from between-state
  differences in baseline rates. The protection factor $R_k$ is large,
  reducing the effective DER well below 1.0. Hierarchical shrinkage
  shields this parameter.

- **beta[2]** (poverty_cwc, `fe_within`): This group-mean-centered
  covariate has $R_k \approx 0$, meaning virtually all of its identifying
  information comes from within-cluster variation. The design effect
  passes through nearly unattenuated, yielding DER close to DEFF.

- **beta[3]** (tiered_reim, `fe_between`): A state-level binary policy
  indicator, identified entirely from between-state comparisons. Like
  the intercept, this parameter is shielded by hierarchical shrinkage.

### Visualizing the protection factor

The protection factor $R_k$ is the key quantity that differentiates
within-cluster from between-cluster covariates. Let us visualize how
$R_k$ scales the raw design effect:

```{r fig-thm1-bar, fig.height = 4.5, fig.cap = "Theorem 1 decomposition for fixed effects. Each bar shows the observed DER, decomposed as DEFF multiplied by (1 - R_k). The within-cluster covariate (poverty_cwc) has R_k near zero and inherits the full design effect, while between-cluster covariates are substantially shielded."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  fe_df <- fe_decomp
  fe_df$param_label <- c("Intercept\n(between)", "Poverty CWC\n(within)",
                         "Tiered Reim.\n(between)")
  fe_df$param_label <- factor(fe_df$param_label,
                              levels = fe_df$param_label)

  # Build stacked components: R_k portion (absorbed) and (1-R_k) portion (exposed)
  fe_df$absorbed <- fe_df$deff_mean * fe_df$R_k
  fe_df$exposed  <- fe_df$deff_mean * (1 - fe_df$R_k)

  fe_long <- data.frame(
    param_label = rep(fe_df$param_label, 2),
    component   = rep(c("Exposed: DEFF x (1 - R_k)", "Absorbed: DEFF x R_k"),
                      each = nrow(fe_df)),
    value       = c(fe_df$exposed, fe_df$absorbed),
    stringsAsFactors = FALSE
  )
  fe_long$component <- factor(fe_long$component,
                              levels = c("Absorbed: DEFF x R_k",
                                         "Exposed: DEFF x (1 - R_k)"))

  ggplot(fe_long, aes(x = param_label, y = value, fill = component)) +
    geom_col(width = 0.6, colour = "white", linewidth = 0.3) +
    geom_hline(yintercept = 1.0, linetype = "dashed",
               colour = "grey40", linewidth = 0.5) +
    annotate("text", x = 3.4, y = 1.0, label = "DER = 1",
             colour = "grey40", size = 3.5, hjust = 0, vjust = -0.5) +
    scale_fill_manual(
      values = c("Absorbed: DEFF x R_k"       = "grey75",
                 "Exposed: DEFF x (1 - R_k)"  = pal["tier_ia"]),
      name = NULL
    ) +
    labs(x = NULL, y = "DER component value",
         title = "Theorem 1: Fixed-Effect DER Decomposition",
         subtitle = sprintf("DEFF = %.2f across all parameters",
                            fe_df$deff_mean[1])) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())
}
```

The visualization makes the dichotomy clear: the within-cluster covariate
absorbs almost none of the design effect (the grey portion is negligible),
while the between-cluster covariates absorb most of it.


## Theorem 2: Random Effects

**Theorem 2** governs the random effects $\theta_j$:

$$
\text{DER}_j = B_j \times \text{DEFF}_j \times \kappa_j(J),
$$

where:

- $B_j = \sigma_\theta^2 / (\sigma_\theta^2 + \tilde{V}_j)$ is the
  **group-specific shrinkage factor**, with $\tilde{V}_j = 1 / \sum_{i \in j} w_i \cdot \text{wt}_i$
  the effective within-group variance.
- $\text{DEFF}_j$ is the **group-specific design effect**, computed from
  the weight distribution within group $j$.
- $\kappa_j(J)$ is the **finite-group correction factor**.

The critical factor is $B_j$. When shrinkage is strong ($B_j$ small),
the prior dominates and the design effect is sharply attenuated. When
shrinkage is weak ($B_j$ near 1), the data dominate and the design
effect passes through.

### Worked example: NSECE random effects

```{r thm2-decompose}
# Random effects decomposition
re_decomp <- decomp[decomp$param_type == "re", ]

# Summary statistics
cat(sprintf("Number of random effects: %d\n", nrow(re_decomp)))
cat(sprintf("DER range (RE): [%.4f, %.4f]\n",
            min(re_decomp$der), max(re_decomp$der)))
cat(sprintf("Mean B: %.4f\n", re_decomp$B_mean[1]))
cat(sprintf("Mean kappa: %.4f\n", re_decomp$kappa[1]))
cat(sprintf("Mean DEFF: %.4f\n", re_decomp$deff_mean[1]))
cat(sprintf("Predicted DER (B * DEFF * kappa): %.4f\n",
            re_decomp$der_predicted[1]))
```

With $B$ typically large in the NSECE (because group sizes are
substantial, so the data dominate the prior), the random effects still
experience some design sensitivity. However, the shrinkage factor keeps
DER well below 1.0 for all states, confirming that no random effects
need correction.

### Visualizing shrinkage and design sensitivity

The relationship between shrinkage and design sensitivity is the core
mechanism underlying Theorem 2. States with larger group sizes (and thus
weaker shrinkage --- higher $B_j$) should show higher DER:

```{r fig-thm2-scatter, fig.height = 5, fig.cap = "Relationship between the group-specific shrinkage factor B_j and DER for each of the 51 state random effects. The dashed curve shows the Theorem 2 prediction. States with weaker shrinkage (higher B, typically larger states) show higher design sensitivity."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  # Extract per-group quantities from the svyder object
  B_j    <- result$B_j
  deff_j <- result$deff_j
  J      <- result$n_groups

  # Compute kappa_j from the actual formula used in the package
  kappa_j <- (J - 1) * (1 - B_j) / (J * (1 - B_j) + B_j)

  # Predicted DER per group
  der_predicted_j <- B_j * deff_j * kappa_j

  # Empirical DER for random effects (columns p+1 to p+J)
  n_fe   <- sum(decomp$param_type != "re")
  der_re <- result$der[(n_fe + 1):(n_fe + J)]

  re_df <- data.frame(
    state         = seq_len(J),
    B_j           = B_j,
    deff_j        = deff_j,
    kappa_j       = kappa_j,
    der_empirical = der_re,
    der_predicted = der_predicted_j
  )

  # Smooth prediction curve
  B_seq <- seq(min(B_j) * 0.95, max(B_j) * 1.02, length.out = 100)
  kappa_seq <- (J - 1) * (1 - B_seq) / (J * (1 - B_seq) + B_seq)
  pred_curve <- data.frame(
    B = B_seq,
    DER = B_seq * mean(deff_j) * kappa_seq
  )

  ggplot(re_df, aes(x = B_j, y = der_empirical)) +
    geom_line(data = pred_curve, aes(x = B, y = DER),
              colour = pal["tier_ii"], linewidth = 0.8, linetype = "dashed") +
    geom_point(aes(colour = deff_j), size = 2.5, alpha = 0.8) +
    scale_colour_gradient(low = "#A6DBA0", high = "#1B7837",
                          name = expression(DEFF[j])) +
    geom_hline(yintercept = 1.0, linetype = "dotted",
               colour = pal["threshold"], linewidth = 0.5) +
    annotate("text", x = max(B_j), y = 1.02, label = "DER = 1",
             colour = pal["threshold"], size = 3.5,
             hjust = 1, vjust = -0.3) +
    labs(x = expression(paste("Shrinkage factor ", B[j])),
         y = expression(paste("DER"[j])),
         title = "Theorem 2: Random-Effect DER vs Shrinkage",
         subtitle = sprintf("J = %d states; dashed line = B * DEFF * kappa prediction",
                            J)) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "right",
          panel.grid.minor = element_blank())
}
```

The scatter confirms the theoretical prediction: DER increases
monotonically with the shrinkage factor $B_j$, and all random-effect
DER values remain well below the correction threshold. The colour
gradient shows that groups with higher within-group design effects also
tend to have somewhat higher DER, consistent with the multiplicative
structure of Theorem 2.


## The $\kappa(J)$ Factor: Finite-Group Correction

The finite-group correction factor $\kappa_j(J)$ accounts for the fact
that with a finite number of groups $J$, the random-effect posterior is
slightly less sensitive to the design than the asymptotic ($J \to \infty$)
formula would predict. The exact expression implemented in svyder is:

$$
\kappa_j(J) = \frac{(J - 1)(1 - B_j)}{J(1 - B_j) + B_j}.
$$

Two limiting cases are instructive:

- As $J \to \infty$: $\kappa \to 1 - B$, and the finite-group correction
  vanishes from the dominant balance.
- As $J \to 1$: $\kappa \to 0$, reflecting the fact that a single group
  cannot identify the random-effect distribution.

In the NSECE ($J = 51$), $\kappa$ is close to its large-$J$ limit and
has little practical impact. However, for studies with few groups
($J = 5$ or $J = 10$), the correction can be substantial.

```{r fig-kappa, fig.height = 4.5, fig.cap = "Finite-group correction factor kappa as a function of J for different values of the shrinkage factor B. The vertical dashed line marks J = 51 (NSECE). For large J, kappa approaches its asymptotic value; for small J, it can substantially reduce the predicted DER."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  J_seq <- 2:100
  B_vals <- c(0.3, 0.6, 0.9)
  kappa_df <- expand.grid(J = J_seq, B = B_vals)
  kappa_df$kappa <- (kappa_df$J - 1) * (1 - kappa_df$B) /
    (kappa_df$J * (1 - kappa_df$B) + kappa_df$B)
  kappa_df$B_label <- factor(sprintf("B = %.1f", kappa_df$B))

  ggplot(kappa_df, aes(x = J, y = kappa, colour = B_label)) +
    geom_line(linewidth = 0.9) +
    geom_vline(xintercept = 51, linetype = "dashed",
               colour = "grey50", linewidth = 0.5) +
    annotate("text", x = 53, y = 0.05, label = "J = 51\n(NSECE)",
             colour = "grey50", size = 3.5, hjust = 0) +
    scale_colour_manual(
      values = c("B = 0.3" = pal["tier_ib"],
                 "B = 0.6" = pal["threshold"],
                 "B = 0.9" = pal["tier_ii"]),
      name = NULL
    ) +
    labs(x = "Number of groups (J)",
         y = expression(kappa(J)),
         title = expression(paste("Finite-Group Correction Factor ", kappa, "(J)")),
         subtitle = "Higher B and larger J both push kappa toward its asymptote") +
    coord_cartesian(ylim = c(0, 1)) +
    theme_minimal(base_size = 12) +
    theme(legend.position = c(0.85, 0.3),
          legend.background = element_rect(fill = "white", colour = "grey80",
                                           linewidth = 0.3),
          panel.grid.minor = element_blank())
}
```

For the NSECE with $J = 51$ groups, the $\kappa$ values are:

```{r kappa-nsece}
# kappa values at J = 51 for the NSECE shrinkage factors
kappa_vals <- (J - 1) * (1 - B_j) / (J * (1 - B_j) + B_j)
cat(sprintf("kappa range at J = 51: [%.4f, %.4f]\n",
            min(kappa_vals), max(kappa_vals)))
cat(sprintf("kappa mean at J = 51:  %.4f\n", mean(kappa_vals)))
```

These values are close to 1, confirming that the finite-group correction
has minimal practical impact in this application.


## The Conservation Law

Perhaps the most elegant structural result of the DER framework is the
**conservation law**. Under balanced conditions with a common design
effect, the total design sensitivity is conserved across the hierarchy:

$$
\text{DER}_\mu + \text{DER}_\theta^{\text{cond}} = \text{DEFF}.
$$

This equation has a profound interpretation: the hierarchical prior does
not create or destroy design sensitivity --- it **redistributes** it
between the global parameters and the group-level parameters. Every unit
of design sensitivity removed from the random effects (via shrinkage)
reappears in the grand mean.

Think of it as a "conservation budget." The total design sensitivity
available is $\text{DEFF}$. The model must allocate this budget between
levels:

- Random effects absorb $\text{DER}_\theta$ of the budget through
  their dependence on group-specific data.
- The remainder, $\text{DEFF} - \text{DER}_\theta$, falls on the
  fixed effects (particularly the intercept).

### Verifying the conservation law

We can check the conservation law empirically using
`der_theorem_check()`:

```{r conservation-check}
thm_check <- der_theorem_check(result)

conservation <- attr(thm_check, "conservation_law")
if (!is.null(conservation)) {
  cat("Conservation Law Verification\n")
  cat("-----------------------------\n")
  cat(sprintf("  DER_mu (intercept):    %.4f\n", conservation$der_mu))
  cat(sprintf("  DER_theta (mean RE):   %.4f\n", conservation$der_theta_mean))
  cat(sprintf("  Sum:                   %.4f\n", conservation$conservation_sum))
  cat(sprintf("  Mean DEFF:             %.4f\n", conservation$deff_mean))
  cat(sprintf("  Relative error:        %.4f (%.1f%%)\n",
              conservation$relative_error,
              conservation$relative_error * 100))
}
```

The conservation law holds approximately because the NSECE-like data
has unbalanced group sizes and a non-conjugate (logistic) likelihood.
The relative error quantifies the approximation quality. In practice,
this error is typically small enough that the conservation law provides
a useful conceptual framework.

### Visualizing the conservation budget

The "budget" interpretation becomes vivid when we visualize how the total
$\text{DEFF}$ is split between the fixed-effect and random-effect
contributions:

```{r fig-conservation, fig.height = 4, fig.cap = "Conservation law: the total design sensitivity (DEFF) is split between the intercept DER and the mean random-effect DER. The prior redistributes design sensitivity across levels without creating or destroying it."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  if (!is.null(conservation)) {
    budget_df <- data.frame(
      component = factor(
        c("DER (intercept)", "DER (mean RE)"),
        levels = c("DER (mean RE)", "DER (intercept)")
      ),
      value = c(conservation$der_mu, conservation$der_theta_mean)
    )

    ggplot(budget_df, aes(x = "DEFF Budget", y = value, fill = component)) +
      geom_col(width = 0.5, colour = "white", linewidth = 0.4) +
      geom_hline(yintercept = conservation$deff_mean, linetype = "dashed",
                 colour = "grey30", linewidth = 0.7) +
      annotate("text", x = 1.35, y = conservation$deff_mean,
               label = sprintf("DEFF = %.2f", conservation$deff_mean),
               colour = "grey30", size = 4, hjust = 0, vjust = -0.3) +
      annotate("text", x = 1.35, y = conservation$conservation_sum,
               label = sprintf("Sum = %.2f (error: %.1f%%)",
                               conservation$conservation_sum,
                               conservation$relative_error * 100),
               colour = "grey50", size = 3.5, hjust = 0, vjust = -0.3) +
      scale_fill_manual(
        values = c("DER (intercept)" = pal["tier_ib"],
                   "DER (mean RE)"   = pal["tier_ii"]),
        name = NULL
      ) +
      labs(x = NULL, y = "Design sensitivity",
           title = "The Conservation Law: DER Budget",
           subtitle = expression(
             paste(DER[mu], " + ", DER[theta], " ", "" %~~% "", " DEFF"))) +
      theme_minimal(base_size = 12) +
      theme(legend.position = "bottom",
            panel.grid.minor = element_blank(),
            panel.grid.major.x = element_blank(),
            axis.text.x = element_blank())
  }
}
```

The budget interpretation explains a key trade-off in hierarchical
modelling: protecting random effects from the survey design comes at the
cost of exposing fixed effects. This trade-off is not a deficiency of the
method --- it reflects the fundamental structure of design-based
inference in hierarchical models.


## Verifying the Theorems: Empirical vs Theoretical

The decomposition theorems provide simplified closed-form predictions of
DER. How well do these approximations match the exact sandwich-based
computation? The `der_theorem_check()` function answers this question by
comparing the empirical DER from Algorithm 1 against the theoretical
predictions.

### Fixed effects: Theorem 1

```{r thm-check-fe}
fe_check <- thm_check[thm_check$param_type != "re", ]
fe_check[, c("param", "param_type", "der_empirical",
             "der_theorem1", "relative_error", "theorem_used")]
```

The simplified Theorem 1 predictions capture the qualitative pattern
perfectly: within-cluster covariates have DER near $\text{DEFF}$, and
between-cluster covariates have DER near $\text{DEFF}(1-B)$. The
relative errors indicate how well the idealized formulas approximate
the exact values for this non-conjugate, unbalanced model.

### Random effects: Theorem 2

```{r thm-check-re}
re_check <- thm_check[thm_check$param_type == "re", ]
head(re_check[, c("param", "der_empirical", "der_theorem2",
                   "relative_error")], 10)
```

```{r thm-check-re-summary}
cat(sprintf("Mean absolute relative error (RE): %.4f (%.1f%%)\n",
            mean(abs(re_check$relative_error), na.rm = TRUE),
            mean(abs(re_check$relative_error), na.rm = TRUE) * 100))
```

The Theorem 2 predictions use per-group values of $B_j$, $\text{DEFF}_j$,
and $\kappa_j$, and typically approximate the exact DER well. Larger
discrepancies may arise when there is strong coupling between fixed and
random effects in the observed information matrix, or when the model is
highly non-conjugate.

### When do the formulas diverge?

The simplified formulas are derived under idealized assumptions:
balanced groups, a conjugate normal model, and diagonal observed
information. In practice:

- The formulas correctly identify **which** parameters are
  design-sensitive and the **qualitative ordering** of DER values.
- They may not predict exact DER magnitudes when groups are severely
  unbalanced, the model is non-conjugate (logistic, beta-binomial), or
  there is substantial coupling between $\beta$ and $\theta$ in the
  observed information.

This is why svyder always computes exact DER values via the full sandwich
variance (Algorithm 1). The decomposition formulas serve as
**interpretive tools** --- they explain *why* a parameter has its DER
value, but the algorithm provides the definitive answer.


## Comparing Equal-Weight vs Complex Survey

To sharpen the contrast, let us compare the NSECE decomposition with a
balanced, equal-weight dataset where the design effect should be
negligible:

```{r equal-weight}
data(sim_hlr)

result_hlr <- der_diagnose(
  sim_hlr$draws,
  y           = sim_hlr$y,
  X           = sim_hlr$X,
  group       = sim_hlr$group,
  weights     = sim_hlr$weights,
  psu         = sim_hlr$psu,
  family      = sim_hlr$family,
  sigma_theta = sim_hlr$sigma_theta,
  sigma_e     = sim_hlr$sigma_e,
  param_types = sim_hlr$param_types
)
```

```{r equal-weight-decompose}
decomp_hlr <- der_decompose(result_hlr)
decomp_hlr[, c("param", "param_type", "der", "deff_mean",
               "B_mean", "der_predicted")]
```

With equal weights ($\text{DEFF} \approx 1$), the design effect is
negligible and all DER values are near or below 1.0. There is simply
nothing to correct.

### Side-by-side comparison

```{r fig-comparison, fig.height = 5, fig.cap = "DER profiles for the NSECE-like data (complex survey, top) versus balanced equal-weight data (bottom). The complex survey produces a 100-fold range of DER values; the equal-weight design produces DER values near 1.0 across all parameters. The dashed purple line marks the default threshold (tau = 1.2)."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  # Build combined data frame
  td_nsece <- tidy.svyder(result)
  td_hlr   <- tidy.svyder(result_hlr)

  td_nsece$dataset <- "NSECE (complex survey)"
  td_hlr$dataset   <- "Balanced HLR (equal weights)"

  td_nsece$param_idx <- seq_len(nrow(td_nsece))
  td_hlr$param_idx   <- seq_len(nrow(td_hlr))

  # Determine tier colours
  tier_colour <- function(tier) {
    ifelse(tier == "I-a", pal["tier_ia"],
    ifelse(tier == "I-b", pal["tier_ib"],
                          pal["tier_ii"]))
  }

  # NSECE panel
  p1 <- ggplot(td_nsece, aes(x = param_idx, y = der)) +
    geom_hline(yintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.5) +
    geom_point(aes(colour = tier), size = 1.8, alpha = 0.7) +
    scale_colour_manual(
      values = c("I-a" = pal["tier_ia"],
                 "I-b" = pal["tier_ib"],
                 "II"  = pal["tier_ii"]),
      labels = c("I-a" = "Tier I-a (within FE)",
                 "I-b" = "Tier I-b (between FE)",
                 "II"  = "Tier II (RE)"),
      name = "Tier"
    ) +
    labs(x = "Parameter index", y = "DER",
         subtitle = "NSECE (complex survey, DEFF > 3)") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())

  # HLR panel
  p2 <- ggplot(td_hlr, aes(x = param_idx, y = der)) +
    geom_hline(yintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.5) +
    geom_point(aes(colour = tier), size = 1.8, alpha = 0.7) +
    scale_colour_manual(
      values = c("I-a" = pal["tier_ia"],
                 "I-b" = pal["tier_ib"],
                 "II"  = pal["tier_ii"]),
      labels = c("I-a" = "Tier I-a (within FE)",
                 "I-b" = "Tier I-b (between FE)",
                 "II"  = "Tier II (RE)"),
      name = "Tier"
    ) +
    labs(x = "Parameter index", y = "DER",
         subtitle = "Balanced HLR (equal weights, DEFF = 1)") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())

  # Combine
  if (requireNamespace("patchwork", quietly = TRUE)) {
    library(patchwork)
    p1 / p2 +
      plot_annotation(
        title = "DER Profiles: Complex Survey vs Equal Weights",
        theme = theme(plot.title = element_text(size = 14, face = "bold"))
      ) +
      plot_layout(guides = "collect") &
      theme(legend.position = "bottom")
  } else {
    print(p1)
    print(p2)
  }
}
```

The contrast is stark. The complex survey produces a 100-fold range of
DER values driven by the interplay between weight variability and
hierarchical structure. The equal-weight design produces DER values
clustered near 1.0 --- the hierarchical model is already well calibrated
because there is no design effect to account for.


## Putting It All Together: A Decomposition Summary

The following table summarizes the decomposition framework:

| Theorem | Parameter type | DER formula | Key factor | NSECE result |
|:--------|:--------------|:------------|:-----------|:-------------|
| 1 (within) | $\beta_k$ (within-cluster) | $\text{DEFF} \times (1 - R_k)$ | $R_k \approx 0$ | DER $\approx$ DEFF |
| 1 (between) | $\beta_k$ (between-cluster) | $\text{DEFF} \times (1 - R_k)$ | $R_k \approx B$ | DER $\approx$ DEFF(1-B) $<$ 1 |
| 2 | $\theta_j$ (random effects) | $B_j \times \text{DEFF}_j \times \kappa_j$ | $B_j < 1$ | DER $<$ 1 |
| Conservation | All | $\text{DER}_\mu + \text{DER}_\theta = \text{DEFF}$ | Budget constraint | Verified |

The decomposition reveals that the three-tier classification is not
arbitrary --- it emerges naturally from the mathematical structure of
hierarchical models fitted to complex surveys. Each tier corresponds to
a distinct mechanism by which the survey design interacts with the
model's information flow:

- **Tier I-a** inherits the full design effect because within-cluster
  variation is orthogonal to the random-effect prior.
- **Tier I-b** is shielded because between-cluster variation is
  partially absorbed by the random effects.
- **Tier II** is protected because the prior directly attenuates the
  design effect through shrinkage.


## What's Next?

This vignette covered the theoretical foundations of the DER
decomposition. The remaining vignettes show how to use these insights
in practice:

| Vignette | What you will learn |
|:---------|:--------------------|
| *The Compute-Classify-Correct Pipeline* | Step-by-step pipeline walkthrough, sensitivity analysis, cross-clustering comparison, and working with corrected draws |
| *Getting Started* | Quick-start tutorial with the full svyder API |
| *Understanding Design Effect Ratios* | Mathematical foundations of DER, conservation law, and three-tier classification |
