---
title: "Getting Started with svyder"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  Run the full compute-classify-correct DER pipeline in about five minutes on
  bundled survey data, then read, tidy, plot, and extract the corrected draws.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with svyder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", out.width = "85%",
                      warning = FALSE, message = FALSE)
library(svyder)
# Consistent colour palette across all svyder vignettes (colour-blind aware)
svyder_pal <- c(
  `I-a`  = "#D6604D",  # within-group fixed effects (survey-dominated)
  `I-b`  = "#4393C3",  # between-group fixed effects (shielded)  [brand blue]
  `II`   = "#5AAE61",  # random effects (shrinkage-protected)
  `III`  = "#9E9E9E",  # hyperparameters (excluded)
  flag   = "#B2182B",  # flagged
  target = "#1A1A1A"   # DER = 1 / threshold reference lines
)
```

## Overview

This vignette takes you from a matrix of posterior draws to a set of
design-corrected draws in a single function call. Working on bundled data, you
will run the whole **compute -> classify -> correct** pipeline in about five
minutes: run the diagnostic, read its report, pull the results into a tidy
table, draw one picture, and extract corrected posterior draws you can hand
straight to downstream code. No Stan installation is required --- the demo
datasets ship with precomputed draws.

It is written for practitioners who fit Bayesian hierarchical models to complex
survey data and want to know, parameter by parameter, whether the survey design
has been properly accounted for. You do not need to understand the sandwich
variance or the classification theory to use the package; this vignette stays
on the happy path and links out to the details.

We cover, in order:

- **Why DER?** The problem the diagnostic solves.
- **Installation** and the **five-minute demo** (one call).
- **Reading** the printed report, the **tidy table**, and the **model-level glance**.
- **One plot**, then **extracting corrected draws**.
- **The one decision you must make**, and where to go next.

## The problem: why DER?

When you fit a Bayesian hierarchical model to survey data with a pseudo-posterior
(plugging survey weights into the likelihood), the resulting posterior does not
know about the sampling design. For *some* parameters this matters a great deal:
the posterior is too narrow, understating the true design uncertainty, and any
interval you report is overconfident. For *other* parameters --- especially the
random effects that hierarchical shrinkage already protects --- the posterior is
fine, and "correcting" it would wrongly *narrow* it and throw away the very
shrinkage that makes the model work.

So neither extreme is safe. Ignoring the design leaves some parameters
overconfident; a blanket correction applied to every parameter damages the ones
that were already calibrated.

> **Key insight.** The Design Effect Ratio (DER) is a per-parameter number that
> tells you *which* parameters are which. DER $> 1$ means the posterior
> understates design uncertainty and should be widened; DER $\approx 1$ means
> leave it alone; DER $< 1$ means a correction would harm it. svyder computes a
> DER for every parameter, flags only those above a threshold, and corrects
> only the flagged ones --- leaving the rest untouched.

Formally, the DER of parameter $k$ is the ratio of its design-based *sandwich*
variance (for a declared variance target) to its model-based MCMC posterior
variance,
$$\mathrm{DER}_k = \frac{[V_{\text{target}}]_{kk}}{[\Sigma_{\text{MCMC}}]_{kk}}.$$
You will not need that formula to follow along --- see `vignette("theory-der")`
for the full story --- but it is the quantity every number below refers to.

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("joonho112/svyder")
```

Then load it:

```{r load, eval = FALSE}
library(svyder)
```

## The 5-minute demo

Load the bundled dataset. `nsece_demo` is a synthetic list modelled on the
2019 National Survey of Early Care and Education, carrying everything the
diagnostic needs:

```{r data}
data(nsece_demo)
```

The pieces you will pass in are:

- `draws` --- a $4000 \times 54$ matrix of posterior draws (one column per
  parameter: 3 fixed effects, 51 random effects for the states, plus a
  hyperparameter).
- `y`, `X` --- the binary outcome ($N = 6785$ providers) and the design matrix
  with three columns: `intercept`, `poverty_cwc`, `tiered_reim`.
- `group` --- the grouping factor, $J = 51$ states.
- `weights` --- the survey weights.
- `psu` --- the design's primary sampling units: **644 PSUs**. This is the
  aggregation unit you will pass as `cluster`.

That is a moderately large, genuinely clustered survey: 6,785 observations
nested in 51 states, drawn from 644 PSUs, with 3 fixed effects. One call runs
the entire pipeline:

```{r diagnose}
result <- der_diagnose(
  nsece_demo$draws,
  y            = nsece_demo$y,
  X            = nsece_demo$X,
  group        = nsece_demo$group,
  weights      = nsece_demo$weights,
  cluster      = nsece_demo$psu,          # design PSUs = the aggregation unit
  family       = "binomial",
  sigma_theta  = nsece_demo$sigma_theta,
  param_types  = nsece_demo$param_types
)
```

`der_diagnose()` computes a DER for every parameter, classifies each into a
tier, and (by default) applies the selective correction --- all in one step. The
returned object is everything you need for the rest of this vignette.

## Reading the output

Print it to get the headline diagnostic:

```{r print}
print(result)
```

Read this top to bottom:

- **54 parameters**, `binomial` family, `N = 6785`, `J = 51` --- confirms the
  model svyder saw.
- **Target: 644 cluster(s)** --- the design-based target you declared with
  `cluster = nsece_demo$psu`. The meat is centered and DF-corrected and weights
  use the default `unit_mean` convention; these are the v0.2.0 defaults.
- **DER range: [0.235, 5.308]** --- across all parameters, the smallest DER is
  0.235 (well below 1: correcting it would narrow the posterior) and the largest
  is 5.308 (badly understated).
- **Tier III (DER undefined): sigma_theta** --- the variance hyperparameter is
  excluded by construction. It is reported but never certified as corrected.
- **Flagged: 30 / 54 (55.6%)** at the default threshold $\tau = 1.2$. These are
  the parameters whose DER exceeds the threshold; only these are corrected.

The flagged list previews the biggest offenders (`beta[2]` and a run of state
random effects) and confirms `Correction applied: 30 parameter(s) rescaled`.

> **Note.** Better than half the parameters here are flagged, but that is a
> property of *this design and this target*, not a universal fact. Switch the
> aggregation unit and the count changes --- see "The one decision you must
> make" below.

## The tidy table

For a one-row-per-parameter data frame, use `tidy()`:

```{r tidy}
td <- tidy(result)
head(td, 3)
```

The columns are the term, its posterior `estimate` and `std.error`, its `der`,
the assigned `tier`, the `action` taken, whether it was `flagged`, and the
`scale_factor` applied by the correction. The first three rows are the fixed
effects. Reading them:

```{r tidy-fixed, echo = FALSE}
knitr::kable(
  data.frame(
    term   = c("beta[1] (intercept)", "beta[2] (poverty_cwc)", "beta[3] (tiered_reim)"),
    der    = round(td$der[1:3], 3),
    tier   = td$tier[1:3],
    action = td$action[1:3]
  ),
  col.names = c("Parameter", "DER", "Tier", "Action")
)
```

- **`beta[2]` is `poverty_cwc`** (the second column of `X`, hence `beta[2]`):
  DER `r round(td$der[2], 3)`, classified **Tier I-a**, and **flagged for
  correction**. This is a within-group covariate --- it varies among providers
  inside each state --- so the survey design bears on it fully, and its posterior
  was understating uncertainty by a factor near the design effect.
- **`beta[1]` (intercept)**, DER `r round(td$der[1], 3)`, and **`beta[3]`
  (`tiered_reim`)**, DER `r round(td$der[3], 3)`, are both **Tier I-b** and
  **retained**. These are between-group quantities: hierarchical shrinkage
  already shields them, their DERs sit well below 1, and correcting them would
  wrongly narrow already-honest intervals.

So of the three fixed effects, only poverty is corrected. The remaining 29
flagged parameters are state random effects (Tier II).

## Model-level glance

For the aggregate picture in one row, use `glance()`:

```{r glance}
glance(result)
```

This reports the totals you would put in a methods paragraph: `n_params`,
`n_flagged` and `pct_flagged`, the threshold `tau`, the model family and
dimensions, and design summaries --- `mean_deff` around 2.6 (the Kish design
effect is elevated here), `mean_B` around 0.85, and the DER range
[0.235, 5.308].

## One picture

The built-in profile plot lays every parameter out by DER, grouped by tier, with
the DER $= 1$ and threshold reference lines drawn in:

```{r plot-profile, fig.alt = "DER profile: each parameter's Design Effect Ratio, grouped by tier, with reference lines at DER = 1 and the threshold tau = 1.2. Parameters above the threshold are flagged for correction."}
plot(result, type = "profile")
```

Read it at a glance: points to the right of the threshold line are flagged;
points to the left of DER $= 1$ are the shrinkage-protected parameters a blanket
correction would have harmed. `poverty_cwc` sits out to the right; the retained
fixed effects and many random effects sit to the left.

## Extracting corrected draws

To feed downstream code, pull out the corrected posterior draws with
`as.matrix()`:

```{r as-matrix}
corrected <- as.matrix(result)
dim(corrected)
```

You get a matrix the same shape as the input draws ($4000 \times 54$), but with
the flagged parameters' columns rescaled to their declared variance target. The
correction is selective in the strictest sense:

> **Key insight.** Unflagged parameters are left **bitwise unchanged** --- their
> columns in the corrected matrix are identical to the originals, not merely
> rescaled by a factor of one. Only the 30 flagged parameters are touched. You
> can use `corrected` anywhere you would have used the original draws.

You can confirm the guarantee directly on any retained parameter, for example
the intercept (`beta[1]`, column 1):

```{r verify-unchanged}
identical(corrected[, 1], nsece_demo$draws[, 1])
```

## The one decision you must make

There is exactly one choice you cannot skip, and you already made it above:
**`cluster` is required and has no default.** In the demo you set
`cluster = nsece_demo$psu`, declaring a *design-based* target built from the
survey's 644 PSUs.

That choice is not a formality --- it is part of the estimand. The aggregation
unit defines *which* variance target the DER is measured against, and the DER of
a parameter changes materially depending on it. Setting `cluster = group`
instead would declare a *model-aligned* target and give you different DERs, a
different flagged set, and a different correction. On this demo the design-PSU
target flags 30 of 54 parameters; the model-group target flags 25 of 54.

> **Note.** Always pass `cluster =` explicitly and choose it deliberately. Do
> **not** use the older `psu =` argument --- it is deprecated and emits a warning.
> Which target to declare is the subject of its own vignette.

Deciding between the design-PSU and model-group targets --- and the related
`normalize`, `strata`, and `cluster_strata` options --- is covered in
[Choosing the Variance Target](choosing-the-target.html).

## What's next?

You have run the whole pipeline in one call. From here:

- [Choosing the Variance Target](choosing-the-target.html) --- the one decision
  above, in depth: design-PSU vs model-group, weight conventions, stratification,
  and `der_compare()`.
- [The Compute–Classify–Correct Pipeline](pipeline.html) --- split
  `der_diagnose()` into `der_compute()`, `der_classify()`, and `der_correct()`
  for fine-grained control, tune the threshold with `der_sensitivity()`, and see
  why blanket correction is harmful.
- For the underlying concepts --- what a DER *is*, the sandwich target, and the
  tier theory --- see `vignette("theory-der")`.
