---
title: "Getting Started with svyder"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A hands-on introduction to Design Effect Ratio diagnostics with svyder:
  run the full pipeline in one call, interpret the three-tier classification,
  visualize DER profiles, extract corrected posterior draws, and build a
  pipe-friendly workflow --- all using bundled datasets.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with svyder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(svyder)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
set.seed(42)

# Consistent colour palette across all svyder vignettes
pal <- c(
  tier_ia    = "#D6604D",
  tier_ib    = "#4393C3",
  tier_ii    = "#1B7837",
  threshold  = "#762A83"
)
```

## Overview

Bayesian hierarchical models are the workhorse for analysing grouped survey
data.  But complex survey designs --- unequal weights, stratification,
clustering --- create a tension: the model-based posterior can be too narrow
for some parameters while perfectly calibrated (or even conservative) for
others.  Existing corrections either ignore the survey design entirely or
apply a blanket adjustment that destroys the hierarchical structure.

The **svyder** package resolves this tension through the Design Effect Ratio
(DER), a per-parameter diagnostic that identifies *which* parameters need
correction and applies the adjustment selectively.

In this vignette you will learn how to:

- **Understand** why Bayesian hierarchical models need parameter-specific
  design corrections.
- **Run** a complete DER diagnostic in a single function call.
- **Interpret** the three-tier classification (Tier I-a, I-b, II).
- **Visualize** DER results with three built-in plot types.
- **Extract** corrected posterior draws and tidy summaries.
- **Build** a pipe-friendly workflow with fine-grained control.


## The Problem: Why Do We Need DER?

Consider a hierarchical logistic regression with state-level random effects,
fitted to data from a complex survey.  The model has three types of parameters:

$$
y_i \mid \beta, \theta_{j[i]} \sim
\text{Bernoulli}\bigl(\text{logit}^{-1}(\beta_0 + \beta_1 x_i + \beta_2 z_j
+ \theta_j)\bigr), \quad \theta_j \sim N(0, \sigma_\theta^2),
$$

where $x_i$ varies *within* clusters (e.g., individual poverty) and $z_j$
varies only *between* clusters (e.g., state policy).

When the survey design is ignored, the posterior for the within-cluster
coefficient $\beta_1$ can be substantially too narrow because survey weights
and clustering inflate variance that the model does not account for.  The
classical fix --- a sandwich variance estimator --- tells us the "right"
variance for every parameter simultaneously, and a blanket Cholesky
correction rescales the entire posterior to match.

The problem with blanket correction is devastating: it treats all 54
parameters identically.  In a typical NSECE-like model:

- **1 parameter** (the within-state poverty coefficient) genuinely needs
  correction --- its DER is approximately 2.6.
- **53 parameters** (2 between-state fixed effects + 51 random effects)
  have DER well below 1.0 --- their posteriors are already well-calibrated
  or conservative.

Blanket correction forces the corrected covariance to match the sandwich
matrix for all parameters.  For random effects, this can shrink credible
intervals to as little as **4% of their original width**, destroying the
variance gains from hierarchical shrinkage that motivated the model in the
first place.

The DER framework provides a third option: **selective correction**.
Identify the one parameter that needs help, correct it, and leave the
other 53 untouched.

```{r fig-problem-concept, echo = FALSE, fig.height = 3.5, fig.cap = "The dilemma of blanket correction. In a 54-parameter model, only 1 parameter (1.9%) needs design-based correction. Blanket correction inappropriately modifies the remaining 53 parameters, while selective correction targets only the design-sensitive parameter."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  concept_df <- data.frame(
    approach = factor(
      rep(c("Ignore Design", "Blanket Correction", "Selective (svyder)"), each = 2),
      levels = c("Ignore Design", "Blanket Correction", "Selective (svyder)")
    ),
    outcome = rep(c("Correctly handled", "Mishandled"), 3),
    count = c(53, 1, 1, 53, 54, 0)
  )
  concept_df$outcome <- factor(concept_df$outcome,
                                levels = c("Correctly handled", "Mishandled"))

  ggplot(concept_df, aes(x = approach, y = count, fill = outcome)) +
    geom_col(position = "stack", width = 0.6, colour = "white",
             linewidth = 0.4) +
    geom_text(aes(label = count), position = position_stack(vjust = 0.5),
              colour = "white", fontface = "bold", size = 4.5) +
    scale_fill_manual(
      values = c("Correctly handled" = pal["tier_ii"],
                 "Mishandled" = pal["tier_ia"]),
      name = NULL
    ) +
    labs(x = NULL, y = "Number of parameters (out of 54)",
         title = "Why Selective Correction Matters") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())
}
```


## Installation

```{r install, eval = FALSE}
# Development version from GitHub
remotes::install_github("joonho112/svyder")
```

After installation, load the package:

```{r load}
library(svyder)
```


## The 5-Minute Demo

The svyder package ships with `nsece_demo`, a synthetic dataset modelled
after the 2019 National Survey of Early Care and Education (NSECE).  It
contains all the ingredients needed for a DER analysis --- posterior draws,
survey weights, and design structure --- so no external model fitting is
required.

### Data structure

```{r data-load}
data(nsece_demo)
```

Let us examine what the dataset contains:

```{r data-structure}
cat("Components:\n")
cat(paste(" ", names(nsece_demo), collapse = "\n"), "\n\n")

cat(sprintf("Observations:       N = %d\n", nsece_demo$N))
cat(sprintf("Groups (states):    J = %d\n", nsece_demo$J))
cat(sprintf("Fixed effects:      p = %d\n", nsece_demo$p))
cat(sprintf("Posterior draws:    S = %d\n", nrow(nsece_demo$draws)))
cat(sprintf("Total parameters:   d = %d (p + J)\n",
            ncol(nsece_demo$draws)))
cat(sprintf("Family:             %s\n", nsece_demo$family))
cat(sprintf("Random effect SD:   sigma_theta = %.2f\n",
            nsece_demo$sigma_theta))
```

The `param_types` vector tells svyder how each fixed effect draws its
identifying information:

```{r param-types}
data.frame(
  covariate  = c("Intercept", "Poverty (CWC)", "Tiered Reim."),
  param_type = nsece_demo$param_types,
  description = c(
    "Identified from between-state comparisons",
    "Identified from within-state variation",
    "Identified from between-state comparisons"
  )
)
```

### Running the diagnostic

The full DER pipeline --- compute, classify, correct --- runs in a single
call to `der_diagnose()`:

```{r diagnose}
result <- der_diagnose(
  x           = nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  psu         = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types,
  tau         = 1.2
)
```

### Reading the print output

```{r print-result}
result
```

The `print` method gives a concise diagnostic summary:

- **DER range**: the minimum and maximum DER across all parameters.
- **Threshold (tau)**: the classification cutoff (default 1.2).
- **Flagged**: how many parameters exceed the threshold.
- **Flagged parameters**: the parameter name, DER value, tier, and action.

In this example, exactly 1 of 54 parameters is flagged: `beta[2]` (the
within-state poverty coefficient) with DER $\approx$ 2.6.  This means its
model-based posterior variance is only about 38% of what the sandwich
variance indicates it should be.  The correction will widen its credible
interval by a factor of $\sqrt{2.6} \approx 1.6$.

### The summary view

```{r summary-result}
summary(result)
```

The `summary()` method displays the full classification table for all
parameters, including tier assignment, DER value, and whether each
parameter is flagged.


## Understanding the Output

The `der_diagnose()` result is an S3 object of class `svyder`.  Its main
components are:

| Component | Description |
|:----------|:------------|
| `der` | Named numeric vector of DER values (length $d = p + J$) |
| `classification` | Data frame with tier, flag status, and action for each parameter |
| `V_sand` | Sandwich variance matrix ($d \times d$) |
| `sigma_mcmc` | Posterior covariance from MCMC ($d \times d$) |
| `deff_j` | Per-group design effects (length $J$) |
| `B_j` | Per-group shrinkage factors (length $J$) |
| `corrected_draws` | Corrected posterior draws matrix ($S \times d$) |
| `scale_factors` | Correction scale factors (length $d$; 1.0 for unflagged) |
| `original_draws` | Original (uncorrected) posterior draws |

### DER values

The DER vector contains one value per parameter:

```{r der-values}
# Fixed-effect DER values
cat("Fixed effects:\n")
round(result$der[1:3], 4)

cat("\nRandom effects (first 10):\n")
round(result$der[4:13], 4)
```

Notice the striking contrast: `beta[2]` (within-state poverty) has DER
$\approx$ 2.6, while all other parameters have DER well below 1.0.  This
100-fold difference in design sensitivity across parameters in the *same
model* is exactly what motivates selective correction.

### Per-group diagnostics

The per-group design effects and shrinkage factors reveal the survey
structure:

```{r group-diagnostics}
cat("Design effects (DEFF) by state --- first 10:\n")
round(result$deff_j[1:10], 3)

cat(sprintf("\nMean DEFF across states: %.3f\n", mean(result$deff_j)))

cat("\nShrinkage factors (B) by state --- first 10:\n")
round(result$B_j[1:10], 3)

cat(sprintf("\nMean B across states: %.3f\n", mean(result$B_j)))
```

A mean DEFF of approximately 3.5 means the survey carries only about 29% of
the information that an equally sized simple random sample would provide.
A mean shrinkage factor $B$ near 0.96 indicates that most states have enough
data that their random effects rely primarily on their own observations rather
than borrowing from the grand mean.

### The glance summary

The `glance.svyder()` method provides a one-row overview of the diagnostic:

```{r glance}
glance.svyder(result)
```


## Visualizing Results

The `plot()` method provides three complementary diagnostic visualizations.
When ggplot2 is installed, all plots are returned as `ggplot` objects that
can be further customised.

### Profile plot

The default profile plot shows DER values for each parameter, coloured by
tier, with the threshold $\tau$ as a dashed line:

```{r fig-profile, fig.height = 5, fig.cap = "DER profile plot. Each dot represents one parameter, coloured by its tier classification. The dashed purple line marks the threshold tau = 1.2. Only beta[2] (the within-state poverty coefficient) exceeds the threshold, confirming that it is the sole parameter requiring design-based correction."}
plot(result, type = "profile")
```

The profile plot is the single most informative DER visualization.  At a
glance, you can see:

- **Which parameters are flagged** (above the dashed line).
- **The magnitude of design sensitivity** (distance from 1.0).
- **The tier structure** (colour coding distinguishes within-cluster FE,
  between-cluster FE, and random effects).

### Decomposition plot

The decomposition plot compares observed DER values against their
theoretical predictions from the decomposition formulas:

```{r fig-decomposition, fig.height = 5, fig.cap = "DER decomposition: observed versus predicted values. Points near the 1:1 line indicate good agreement between the closed-form decomposition and the full sandwich computation. Fixed effects (orange/blue) and random effects (green) cluster in distinct regions, reflecting their different design sensitivity."}
plot(result, type = "decomposition")
```

This plot verifies that the theoretical framework correctly predicts the
empirical DER values.  When points lie close to the 1:1 line, the
simplified decomposition formulas (Theorems 1 and 2) accurately capture
the survey design's effect on each parameter.

### Comparison plot

The comparison plot shows naive versus corrected credible intervals for
flagged parameters:

```{r fig-comparison, fig.height = 3.5, fig.cap = "Credible interval comparison for the flagged parameter beta[2]. Grey shows the naive model-based interval; the coloured interval shows the DER-corrected result. The correction widens the interval to properly account for the survey design, ensuring valid frequentist coverage."}
plot(result, type = "comparison")
```

For unflagged parameters the intervals are identical (scale factor = 1.0),
so the plot displays only flagged parameters by default.

### Customising with autoplot

For full ggplot2 control, use `autoplot()`:

```{r fig-autoplot, fig.height = 5, fig.cap = "Customised DER profile using autoplot(). The ggplot2 object can be modified with standard ggplot2 functions, here adding a custom title and theme adjustments."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  p <- autoplot(result, type = "profile") +
    labs(title = "DER Diagnostic Profile --- NSECE Demo",
         subtitle = "Selective correction targets only design-sensitive parameters") +
    theme(plot.title = element_text(face = "bold"))
  print(p)
}
```


## Extracting Results

### Tidy parameter table

The `tidy.svyder()` method returns a one-row-per-parameter data frame
following broom conventions:

```{r tidy}
td <- tidy.svyder(result)
head(td, 10)
```

The data frame includes:

| Column | Description |
|:-------|:------------|
| `term` | Parameter name |
| `estimate` | Posterior mean |
| `std.error` | Posterior standard deviation |
| `der` | Design Effect Ratio |
| `tier` | Three-tier classification |
| `action` | `"CORRECT"` or `"retain"` |
| `flagged` | Whether the parameter exceeds the threshold |
| `scale_factor` | Cholesky scale factor applied |

### Model-level summary

The `glance.svyder()` method returns a one-row data frame with aggregate
diagnostics:

```{r glance-table}
gl <- glance.svyder(result)
t(gl)
```

### Corrected posterior draws

The `as.matrix()` method extracts the posterior draws.  If correction has
been applied, corrected draws are returned; otherwise the originals:

```{r draws}
draws_corrected <- as.matrix(result)
cat(sprintf("Draws matrix: %d samples x %d parameters\n",
            nrow(draws_corrected), ncol(draws_corrected)))
```

These corrected draws are suitable for any downstream analysis ---
posterior summaries, credible intervals, posterior predictive checks, or
derived quantities.

### Comparing original vs corrected intervals

A practical check: compute 90% credible intervals from both the original
and corrected draws for the fixed effects.

```{r ci-comparison}
# Fixed effects are in columns 1:3
fe_original  <- result$original_draws[, 1:3]
fe_corrected <- draws_corrected[, 1:3]

ci_orig <- apply(fe_original, 2, quantile, probs = c(0.05, 0.50, 0.95))
ci_corr <- apply(fe_corrected, 2, quantile, probs = c(0.05, 0.50, 0.95))

colnames(ci_orig) <- colnames(ci_corr) <-
  c("Intercept", "Poverty (CWC)", "Tiered Reim.")

cat("Original 90% credible intervals:\n")
round(ci_orig, 3)

cat("\nCorrected 90% credible intervals:\n")
round(ci_corr, 3)
```

Observe that only the poverty coefficient interval widens (reflecting the
DER correction with scale factor $\sqrt{2.6} \approx 1.6$), while the
intercept and tiered reimbursement intervals remain unchanged.

### Visualising the CI comparison

```{r fig-ci-manual, fig.height = 4, fig.cap = "Manual credible interval comparison for the three fixed effects. The poverty coefficient (beta[2]) is the only parameter whose interval widens after correction. The intercept and tiered reimbursement coefficient are identical before and after correction because their DER values are below the threshold."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  fe_names <- c("Intercept", "Poverty (CWC)", "Tiered Reim.")
  ci_df <- data.frame(
    param   = rep(fe_names, 2),
    type    = rep(c("Original", "Corrected"), each = 3),
    lower   = c(ci_orig[1, ], ci_corr[1, ]),
    median  = c(ci_orig[2, ], ci_corr[2, ]),
    upper   = c(ci_orig[3, ], ci_corr[3, ])
  )
  ci_df$param <- factor(ci_df$param, levels = rev(fe_names))
  ci_df$type  <- factor(ci_df$type, levels = c("Original", "Corrected"))

  ggplot(ci_df, aes(x = median, y = param, colour = type)) +
    geom_pointrange(aes(xmin = lower, xmax = upper),
                    position = position_dodge(width = 0.4),
                    size = 0.5, linewidth = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    scale_colour_manual(
      values = c(Original = "grey50", Corrected = pal["tier_ia"]),
      name = NULL
    ) +
    labs(x = "Coefficient estimate (90% CI)", y = NULL,
         title = "Fixed-Effect Credible Intervals: Original vs Corrected") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())
}
```


## The Pipe-Friendly Pipeline

The `der_diagnose()` function is a convenience wrapper.  For more control,
you can call each step individually using the pipe operator:

```{r pipeline}
# Step-by-step pipeline
obj <- der_compute(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  psu         = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types
)
```

At this point the DER values have been computed but no classification or
correction has been applied:

```{r pipeline-inspect}
obj
```

Now classify with a threshold.  You can try different values without
recomputing the expensive sandwich matrices:

```{r pipeline-classify}
obj <- der_classify(obj, tau = 1.2)
```

Finally, apply selective correction to flagged parameters:

```{r pipeline-correct}
obj <- der_correct(obj)
obj
```

The step-by-step approach is useful when you want to:

- **Inspect intermediate results** (e.g., check the sandwich matrix
  before classification).
- **Try different thresholds** without recomputing the DER values.
- **Skip correction** if you only need the diagnostic.

### Exploring alternative thresholds

```{r alt-threshold}
# What if we use a stricter threshold?
obj_strict <- der_classify(obj, tau = 2.0)

# What about a looser one?
obj_loose <- der_classify(obj, tau = 1.0)
```

The classification output immediately tells you how many parameters would
be flagged under each threshold.  With this NSECE-like data, the result is
remarkably stable: the same parameter (`beta[2]`) is the only one flagged
across a wide range of thresholds, because its DER of 2.6 stands well above
all other parameters.


## DER with Equal Weights: A Sanity Check

To verify that svyder behaves correctly when there are no design effects,
we can run the diagnostic on `sim_hlr`, a balanced Gaussian dataset with
equal weights (DEFF = 1):

```{r sim-hlr}
data(sim_hlr)

result_hlr <- der_diagnose(
  x           = sim_hlr$draws,
  y           = sim_hlr$y,
  X           = sim_hlr$X,
  group       = sim_hlr$group,
  weights     = sim_hlr$weights,
  psu         = sim_hlr$psu,
  family      = sim_hlr$family,
  sigma_theta = sim_hlr$sigma_theta,
  sigma_e     = sim_hlr$sigma_e,
  param_types = sim_hlr$param_types,
  tau         = 1.2
)

result_hlr
```

With equal weights, no parameters are flagged.  All DER values are near 1.0
or below, confirming that the hierarchical model is already well-calibrated
when there is no survey design effect.  This is the expected behaviour:
svyder is conservative and only flags parameters when the design genuinely
distorts the posterior.

```{r fig-hlr-profile, fig.height = 4, fig.cap = "DER profile for the equal-weight sim_hlr dataset. All parameters are near or below the threshold, confirming the absence of design effects. The hierarchical model is well-calibrated without any correction."}
plot(result_hlr, type = "profile")
```


## Key Takeaway

Selective correction is always preferable to blanket correction when the
goal is valid marginal coverage for each parameter.  In this NSECE-like
example:

- Only **1 of 54 parameters** (1.9%) needs correction.
- Blanket correction would inappropriately modify 53 parameters.
- For random effects, blanket correction can shrink credible intervals
  to as little as **4% of their original width**.

The svyder package makes selectivity practical: one function call identifies
the design-sensitive parameters and corrects only those, leaving the
well-calibrated hierarchical structure undisturbed.


## What's Next?

| Vignette | What you will learn |
|:---------|:--------------------|
| *Understanding Design Effect Ratios* | DER definition, the three regimes, decomposition theorems, conservation law, and sensitivity analysis |
| *Advanced Workflows* | Custom sandwich matrices, brms/rstanarm integration, multi-level clustering comparisons |
| *Simulation Studies* | Verifying DER coverage properties through Monte Carlo simulation |

For the mathematical details behind the DER framework, continue to
`vignette("understanding-der", package = "svyder")`.
