---
title: "The Compute-Classify-Correct Pipeline"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A step-by-step tour of the svyder workflow: der_compute() builds the
  declared variance target and per-parameter DERs, der_classify() flags the
  parameters that need correction at a threshold tau, and der_correct()
  rescales only those parameters while leaving everything else bitwise
  untouched. Covers der_diagnose() as the one-call shortcut, threshold
  sensitivity, and why selective correction beats a blanket adjustment.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Compute-Classify-Correct Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", out.width = "85%",
                      warning = FALSE, message = FALSE)
library(svyder)
# Consistent colour palette across all svyder vignettes (colour-blind aware)
svyder_pal <- c(
  `I-a`  = "#D6604D",  # within-group fixed effects (survey-dominated)
  `I-b`  = "#4393C3",  # between-group fixed effects (shielded)  [brand blue]
  `II`   = "#5AAE61",  # random effects (shrinkage-protected)
  `III`  = "#9E9E9E",  # hyperparameters (excluded)
  flag   = "#B2182B",  # flagged
  target = "#1A1A1A"   # DER = 1 / threshold reference lines
)
```

## Overview

svyder implements a three-step workflow for correcting Bayesian survey models:
**compute** the parameter-level Design Effect Ratios (DER), **classify** each
parameter into a correction tier, and **correct** only the parameters that need
it. This vignette is for practitioners who want to understand what happens at
each step and when to run the steps separately instead of all at once. You will
see how the correction is *selective by construction* --- flagged parameters are
rescaled and everything else is left exactly as it was.

The single call `der_diagnose()` runs all three steps for you. But separating
them gives you control: you can inspect the DERs before choosing a threshold,
re-run classification at several thresholds, or correct with an alternative
method. We work with the built-in `nsece_demo` dataset throughout, so every
chunk runs with no Stan installation.

We cover, in order:

- **Step 1 --- Compute.** `der_compute()`: what it needs and what it returns.
- **Step 2 --- Classify.** `der_classify()`: the tiers and the threshold `tau`.
- **Step 3 --- Correct.** `der_correct()`: block-Cholesky vs marginal, and the
  bitwise-identity guarantee for unflagged parameters.
- **All in one.** `der_diagnose()` as the equivalent shortcut.
- **Threshold sensitivity.** `der_sensitivity()` and choosing a stable `tau`.
- **Selective vs blanket** correction, and why blanket adjustment is harmful.
- **Working with corrected draws** downstream.

> **Key insight.** The three functions form a pipeline where each stage annotates
> the same `svyder` object. `der_compute()` produces the raw diagnostic;
> `der_classify()` adds tier labels and a flag column; `der_correct()` adds a
> corrected draws matrix. Nothing is thrown away, so you can always step back
> and re-classify at a different threshold.

This vignette is scoped to the *workflow* and the *threshold*. It does not cover
how to choose the aggregation unit or variance target --- see
[Choosing the Variance Target](choosing-the-target.html) for that decision --- and
it does not derive the tier and correction mathematics, which live in
`vignette("theory-classification-correction")`.

## Step 1 --- Compute

`der_compute()` takes posterior draws plus the design ingredients and returns a
`svyder` object carrying the DERs and the matrices behind them. The canonical
call passes the design PSU ids as the aggregation unit:

```{r compute}
result <- der_compute(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  cluster     = nsece_demo$psu,     # aggregation unit (design PSU ids)
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta
)
```

The required inputs are:

- **`x`** --- the posterior draws (an `n_draws x n_params` matrix here; S3 methods
  also accept `brmsfit`, `CmdStanMCMC`, and `stanreg` fits, see
  [Backends](backends.html)).
- **`y`, `X`, `group`, `weights`** --- the response, design matrix, group index,
  and survey weights used to fit the model.
- **`cluster`** --- the aggregation unit for the sandwich meat. This is
  **required and has no default**: the DER changes materially depending on
  whether you target the design PSUs (`cluster = nsece_demo$psu`) or the model
  groups (`cluster = nsece_demo$group`). Always pass it explicitly.
- **`family`** --- `"binomial"` here, or `"gaussian"`.
- **`sigma_theta`** --- the random-effect scale (and `sigma_e` for gaussian
  models), used to form the penalized bread.

`param_types` is optional. When it is `NULL` the types are inferred from the
design matrix --- a column that is constant within every group becomes a
between-group fixed effect --- and a message reports the inference. Passing
`param_types` explicitly suppresses that message. On the demo the inferred types
are `intercept = fe_between`, `poverty_cwc = fe_within`, `tiered_reim = fe_between`.

> **Note.** In the package output the poverty (within-group) covariate is
> `beta[2]` --- the second design-matrix column --- not `beta[1]`. The intercept
> is `beta[1]`. Keep this in mind when reading the flagged list below.

Printing the object *before* classification shows the diagnostic summary:

```{r print-unclassified}
result
```

Read the header line by line. The **Target** describes the declared variance
target: 644 clusters, a meat matrix that is centered and degrees-of-freedom
corrected (the v2 defaults), and `unit_mean` weight normalization. The **DER
range** `[0.235, 5.308]` already tells you there is real per-parameter spread ---
some parameters have DER well below 1 (shielded) and some well above 1
(survey-sensitive). The random-effect scale `sigma_theta` is reported as **Tier
III**: hyperparameters have no direct data score, so their DER is undefined by
construction and they are never certified as corrected.

The returned object carries everything the later steps need:

```{r compute-contents}
# The parameter-level DERs
head(round(result$der, 3))

# Design metadata for the declared target
result$target[c("cluster_n", "normalize", "center", "df_correct")]

# Tier III parameters, excluded by construction
result$excluded
```

It also stores `result$V_sand` and `result$sigma_mcmc` (the sandwich and MCMC
variances whose ratio is the DER), and `result$data` --- the design ingredients ---
so you can *re-target* to a different aggregation unit without re-passing the
inputs. Re-targeting is the subject of
[Choosing the Variance Target](choosing-the-target.html).

## Step 2 --- Classify

`der_classify()` labels each parameter with a tier and flags the ones whose DER
exceeds the threshold `tau`. The default threshold is `tau = 1.2`:

```{r classify}
result <- der_classify(result, tau = 1.2)
```

The four tiers partition the parameters by *how the design touches them*:

- **Tier I-a --- within-group fixed effects.** Fully exposed to the survey design;
  DER equals the Kish design effect. These are the primary correction candidates.
- **Tier I-b --- between-group fixed effects.** Shielded by hierarchical
  shrinkage; DER equals the design effect discounted by the reliability, so these
  usually sit below the threshold.
- **Tier II --- random effects.** Shrinkage-protected; DER is the design effect
  scaled by the reliability and a finite-group correction.
- **Tier III --- hyperparameters.** Excluded by construction (reported in
  `$excluded`, above), never flagged.

The flag rule is a **strict** inequality: a parameter is flagged when
$\mathrm{DER}_k > \tau$, so a DER exactly equal to `tau` is *not* flagged. On the
demo this flags **30 of 54** parameters (55.6%). The classification print shows
the flagged set:

```{r classify-print}
result
```

You can also work with the classification table directly. Here is the flagged
subset:

```{r classify-flagged}
cls <- result$classification
flagged <- cls[cls$flagged, c("param_name", "param_type", "der", "tier", "action")]
head(flagged, 8)
nrow(flagged)
```

The single flagged fixed effect is `beta[2]` (poverty, Tier I-a, DER `r round(result$der["beta[2]"], 3)`) ---
a within-group covariate fully exposed to the design. The other 29 flags are
random effects (`theta[j]`). The intercept `beta[1]` (DER `r round(result$der["beta[1]"], 3)`)
and the between-group policy covariate `beta[3]` (DER `r round(result$der["beta[3]"], 3)`)
are Tier I-b and stay *below* the threshold --- shrinkage shields them, so
correcting them would wrongly narrow their intervals. The tier and correction
mathematics are derived in `vignette("theory-classification-correction")`.

## Step 3 --- Correct

`der_correct()` rescales the flagged block toward the declared variance target
and leaves every unflagged parameter exactly as it was. The default method is
`"block_cholesky"`, the paper's Algorithm 1:

```{r correct}
result <- der_correct(result)
result
```

The two methods differ in how they treat the *joint* structure of the flagged
block:

- **`method = "block_cholesky"`** (default, Algorithm 1) applies a joint
  Cholesky transform to the whole flagged block, matching the target
  covariance --- correlations and marginals alike.
- **`method = "marginal"`** rescales each flagged parameter independently by
  $\sqrt{\mathrm{DER}_k}$. It matches the target *marginals* but not the
  cross-parameter covariance. When exactly one parameter is flagged the two
  methods coincide (a scalar $\sqrt{\mathrm{DER}}$ rescaling).

A third value, `method = "cholesky"`, is retired and **errors** if requested ---
it was ambiguous between the block and per-parameter transforms.

We can confirm that block-Cholesky and marginal agree when a single parameter is
flagged, using the `sim_hlr` reference dataset (which flags exactly one
parameter):

```{r one-flagged-equiv}
r_bc <- der_correct(
  der_classify(
    der_compute(sim_hlr$draws, y = sim_hlr$y, X = sim_hlr$X, group = sim_hlr$group,
                weights = sim_hlr$weights, cluster = sim_hlr$group,
                family = sim_hlr$family, sigma_theta = sim_hlr$sigma_theta,
                sigma_e = sim_hlr$sigma_e),
    tau = 1.2, verbose = FALSE),
  method = "block_cholesky")

r_mg <- der_correct(
  der_classify(
    der_compute(sim_hlr$draws, y = sim_hlr$y, X = sim_hlr$X, group = sim_hlr$group,
                weights = sim_hlr$weights, cluster = sim_hlr$group,
                family = sim_hlr$family, sigma_theta = sim_hlr$sigma_theta,
                sigma_e = sim_hlr$sigma_e),
    tau = 1.2, verbose = FALSE),
  method = "marginal")

sum(r_bc$classification$flagged)                       # exactly one flagged
isTRUE(all.equal(r_bc$corrected_draws, r_mg$corrected_draws))
```

The per-parameter scale factors are recorded in `result$scale_factors` (an entry
of `1` means the parameter was left alone):

```{r scale-factors}
sf <- result$scale_factors
head(round(sf, 3), 12)
```

### The unflagged parameters are bitwise unchanged

This is the guarantee that makes selective correction safe: parameters that were
not flagged are copied through *bit for bit*. Every column of the corrected
matrix that corresponds to an unflagged parameter is identical to the original
draws:

```{r bitwise-check}
flagged <- result$classification$flagged
corrected <- as.matrix(result)          # corrected draws (see below)
original  <- result$original_draws

# Unflagged columns are byte-for-byte identical
identical(corrected[, !flagged], original[, !flagged])

# Flagged columns did change
!identical(corrected[, flagged], original[, flagged])
```

The first check returns `TRUE`: nothing about the shrinkage-protected parameters
was disturbed. Only the flagged block was rescaled.

## All in one --- `der_diagnose()`

`der_diagnose()` runs compute, classify, and correct in a single call with the
same defaults (`tau = 1.2`, `method = "block_cholesky"`, `correct = TRUE`). It is
the shortcut you would use once you have settled on a threshold:

```{r diagnose}
diag_result <- der_diagnose(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  cluster     = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  tau         = 1.2
)
```

It reaches the same classification as running the three steps by hand --- the same
30 flagged parameters:

```{r diagnose-equiv}
sum(diag_result$classification$flagged)   # der_diagnose
sum(result$classification$flagged)        # step-by-step
```

Use `der_diagnose()` for a quick end-to-end pass; step through
`der_compute()` -> `der_classify()` -> `der_correct()` when you want to inspect
the DERs before choosing `tau`, sweep several thresholds, or correct with the
`"marginal"` method.

## Threshold sensitivity

The threshold `tau` is a decision, not a law of nature, so it is worth checking
how the flagged set responds to it. `der_sensitivity()` re-classifies over a grid
of thresholds (default `0.8` to `2.0`) without recomputing the DERs:

```{r sensitivity}
sens <- der_sensitivity(result)
sens[, c("tau", "n_flagged", "pct_flagged")]
```

The flag count declines smoothly as the threshold rises: `40, 35, 34, 34, 30`
(at the default `tau = 1.2`), then `27, 26, 26, 22, 21`, and so on. There is no
cliff --- the count moves by a handful of random effects at a time --- which means
the classification is not perched on a knife-edge. The default of `1.2` sits in a
stable interior region rather than at a boundary where a small change in `tau`
would flip many parameters.

> **Note.** Look for a *plateau* rather than a single "best" threshold. The paper
> reports that the correction decision is robust across $\tau \in [0.8, 2.0]$,
> with 96.9% per-replication decision stability across the tighter band
> $[1.1, 1.6]$ and roughly a 2% false-positive rate at the `1.2` default. If your
> flag count is stable across a band around your chosen `tau`, the exact value
> matters little.

## Selective vs blanket correction

Why correct only the flagged parameters instead of rescaling everything by
$\sqrt{\mathrm{DER}}$? Because a blanket adjustment applies the design inflation
to parameters that were never over-confident to begin with, and *narrows* the
ones that shrinkage was protecting.

The distinction is stark in the paper's real-data and simulation results (these
are paper findings on the real NSECE 2019 data and its simulation study, not
outputs of the synthetic demo):

- **Real data (NSECE 2019).** A blanket correction would rescale **53 of 54**
  parameters toward a rank-deficient target, with a mean interval-width ratio of
  **0.26** and a worst case of **0.04** --- intervals collapsing to a
  twenty-fifth of their honest width. Selective correction instead widens only
  the parameter that needs it: the poverty coefficient's 90% interval widens
  from **[-0.166, -0.072]** to **[-0.207, -0.030]**, a width ratio of **1.88**,
  while every shrinkage-protected parameter is left intact.
- **Simulation.** Under a blanket adjustment the coverage of the unflagged random
  effects collapses from about **90%** to **11--37%**. Selective correction keeps
  their coverage at the nominal level because it never touches them.

> **Key insight.** DER < 1 is a warning, not a target. For a shrinkage-protected
> parameter, "correcting" toward the sandwich variance *narrows* the interval and
> destroys the hierarchical protection. The flagged/unflagged split exists
> precisely so the correction is applied where DER > 1 and withheld where it
> would do harm.

The tiers and the block-Cholesky algorithm that make this split principled are
derived in `vignette("theory-classification-correction")`.

## Working with corrected draws

After correction, `as.matrix()` returns the corrected draws if a correction has
been applied, and the original draws otherwise. This lets you write downstream
code that is agnostic to whether correction happened:

```{r as-matrix}
draws <- as.matrix(result)
dim(draws)

# Corrected posterior summary for the flagged poverty coefficient
b2 <- draws[, "beta[2]"]
c(mean = mean(b2),
  q05  = unname(quantile(b2, 0.05)),
  q95  = unname(quantile(b2, 0.95)))
```

Feed `draws` into whatever comes next --- posterior intervals, predictions,
figures, or a publication table. Because the unflagged columns are the original
draws untouched, any downstream quantity that depends only on protected
parameters is exactly what your model produced; only the survey-sensitive
parameters carry the design correction. The `tidy()` and `glance()` methods give
tabular summaries of the same result and are demonstrated in the
[Case Study](case-study-nsece.html).

## What's next?

You now have the full workflow: `der_compute()` -> `der_classify()` ->
`der_correct()`, the `der_diagnose()` shortcut, and the tools to check the
threshold. Where to go from here:

- **[Choosing the Variance Target](choosing-the-target.html)** --- the upstream
  decision this vignette took for granted: design-PSU vs model-group aggregation,
  weight normalization, and stratification, with `der_compare()` to hold the two
  targets side by side.
- **[Case Study: NSECE-Style Survey Data](case-study-nsece.html)** --- the whole
  pipeline applied end to end on the demo, with interpretation of the fixed and
  random effects, the DER decomposition, and a publication-ready report.
- **`vignette("theory-classification-correction")`** --- the mathematics behind
  the four tiers, the threshold rationale, and the block-Cholesky correction
  (Algorithm 1), including a formal account of why blanket correction is harmful.
