---
title: "The Compute-Classify-Correct Pipeline"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A comprehensive walkthrough of the three-step DER diagnostic pipeline:
  compute sandwich-based DER values, classify parameters into tiers with
  threshold-based flagging, and apply selective Cholesky correction to flagged
  parameters only. Includes sensitivity analysis, cross-clustering comparison,
  and practical guidance for working with corrected posterior draws.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Compute-Classify-Correct Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(svyder)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
set.seed(42)

# Project colour palette (consistent across all svyder vignettes)
pal <- c(
  tier_ia    = "#D6604D",
  tier_ib    = "#4393C3",
  tier_ii    = "#1B7837",
  threshold  = "#762A83"
)
```

## Overview

The svyder package implements a three-step diagnostic pipeline for
Bayesian hierarchical models fitted to complex survey data. Each step
addresses a specific question:

1. **Compute**: How much does each parameter's posterior variance differ
   from the design-consistent target?
2. **Classify**: Which parameters exceed the tolerance threshold and need
   correction?
3. **Correct**: How should we adjust the posterior draws for flagged
   parameters?

This pipeline can be executed as a single call to `der_diagnose()` or as
three modular steps that can be inspected and customized individually.

By the end of this vignette you will be able to:

1. **Execute** the full pipeline using both the all-in-one and step-by-step
   approaches.
2. **Understand** each step's inputs, outputs, and computational details.
3. **Perform sensitivity analysis** across threshold values to assess the
   robustness of classification.
4. **Compare DER** across different clustering definitions using
   `der_compare()`.
5. **Extract and use** corrected posterior draws for downstream analysis.


## The Three-Step Pipeline: Conceptual Overview

The pipeline processes posterior draws from a hierarchical Bayesian model
through three sequential stages:

```
                      +------------------+
  Posterior draws --> | Step 1: COMPUTE  | --> DER values
  Survey metadata     | der_compute()    |     Sandwich matrices
                      +------------------+     Per-group diagnostics
                              |
                              v
                      +------------------+
  Threshold tau ----> | Step 2: CLASSIFY | --> Tier assignments
                      | der_classify()   |     Flagged parameters
                      +------------------+
                              |
                              v
                      +------------------+
                      | Step 3: CORRECT  | --> Corrected draws
                      | der_correct()    |     Scale factors
                      +------------------+
```

Each step produces an updated `svyder` object that carries forward all
previous results. This design allows inspection at any intermediate stage
and makes the pipeline pipe-friendly.


## Step 1: Compute --- `der_compute()`

The compute step is the computational core of the pipeline. It constructs
the sandwich variance matrix $\mathbf{V}_{\text{sand}}$ that respects the
survey design, computes the MCMC posterior covariance
$\boldsymbol{\Sigma}_{\text{MCMC}}$, and calculates the per-parameter
DER.

### Required inputs

| Argument | Description |
|:---------|:------------|
| `x` | Posterior draws matrix ($S \times d$), where $S$ is the number of draws and $d$ is the total number of parameters |
| `y` | Response vector (length $N$) |
| `X` | Design matrix ($N \times p$) |
| `group` | Integer group indicator ($1$ to $J$) |
| `weights` | Survey weights (positive, length $N$) |
| `family` | Model family: `"binomial"` or `"gaussian"` |
| `sigma_theta` | Estimated random-effect standard deviation |

### Optional inputs

| Argument | Description | Default |
|:---------|:------------|:--------|
| `psu` | PSU indicators for clustering | Same as `group` |
| `sigma_e` | Residual SD (Gaussian only) | --- |
| `param_types` | Character vector identifying each covariate as `"fe_within"` or `"fe_between"` | Automatic detection |
| `beta_prior_sd` | Prior SD for fixed effects | 5 |

### Running the compute step

```{r step1-compute}
data(nsece_demo)

obj <- der_compute(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  psu         = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types
)

obj
```

### What the compute step produces

The resulting `svyder` object contains several components:

```{r step1-inspect}
# Per-parameter DER values
cat("DER values (first 5):\n")
print(round(obj$der[1:5], 4))

# Per-group design effects
cat("\nPer-group DEFF (first 5 states):\n")
print(round(obj$deff_j[1:5], 4))

# Per-group shrinkage factors
cat("\nPer-group shrinkage B (first 5 states):\n")
print(round(obj$B_j[1:5], 4))

# Dimensions
cat("\nSandwich matrix dimension:", dim(obj$V_sand), "\n")
cat("MCMC covariance dimension:", dim(obj$sigma_mcmc), "\n")
```

### Under the hood

The compute step performs the following operations:

1. **Observed information**: Computes the Hessian $\mathbf{H}_{\text{obs}}$
   of the weighted log-posterior at the posterior mean.
2. **Clustered score matrix**: Aggregates individual score contributions
   by PSU to form the $J_{\text{cluster}} \times d$ score matrix
   $\mathbf{J}_{\text{cluster}}$.
3. **Sandwich variance**: Forms
   $\mathbf{V}_{\text{sand}} = \mathbf{H}_{\text{obs}}^{-1} \, \mathbf{J}_{\text{cluster}}' \mathbf{J}_{\text{cluster}} \, \mathbf{H}_{\text{obs}}^{-1}$.
4. **DER**: Takes the elementwise ratio of diagonals:
   $\text{DER}_p = [\mathbf{V}_{\text{sand}}]_{pp} / [\boldsymbol{\Sigma}_{\text{MCMC}}]_{pp}$.

The computational cost is dominated by the matrix operations and scales
as $O(d^2)$, where $d = p + J$ is the total number of parameters.


## Step 2: Classify --- `der_classify()`

The classify step assigns each parameter to a design-sensitivity tier and
flags those whose DER exceeds the threshold $\tau$.

### The threshold $\tau$

The threshold determines the boundary between "acceptable" and
"needs correction." The default is $\tau = 1.2$, meaning a parameter is
flagged if its posterior variance would need to increase by more than
20% when the survey design is accounted for.

### Running the classify step

```{r step2-classify}
obj <- der_classify(obj, tau = 1.2)
```

### Examining the classification

```{r step2-inspect}
# Classification table for fixed effects
cls <- obj$classification
cls[cls$param_type != "re",
    c("param_name", "param_type", "der", "tier", "tier_label",
      "flagged", "action")]
```

The three-tier structure emerges directly from the parameter types:

| Tier | Code | Condition | Typical DER |
|:-----|:-----|:----------|:------------|
| I-a | `fe_within` | Within-cluster covariate | $\approx \text{DEFF}$ |
| I-b | `fe_between` | Between-cluster covariate | $\approx \text{DEFF}(1-B)$ |
| II | `re` | Random effect | $\approx B \times \text{DEFF}$ |

The flagging uses strict inequality: $\text{DER}_p > \tau$ triggers
correction. A parameter with $\text{DER}$ exactly at $\tau$ is **not**
flagged.

```{r step2-flag-summary}
cat(sprintf("Total parameters: %d\n", nrow(cls)))
cat(sprintf("Flagged (DER > %.1f): %d\n", obj$tau, sum(cls$flagged)))
cat(sprintf("Retained: %d\n", sum(!cls$flagged)))
```


## Step 3: Correct --- `der_correct()`

The correct step applies selective Cholesky correction to the flagged
parameters, rescaling their posterior draws so that the marginal
variance matches the sandwich variance. Unflagged parameters are left
**bitwise identical** to the originals.

### How the correction works

For each flagged parameter $p$:

1. Compute the scale factor:
   $s_p = \sqrt{\text{DER}_p} = \sqrt{[\mathbf{V}_{\text{sand}}]_{pp} / [\boldsymbol{\Sigma}_{\text{MCMC}}]_{pp}}$.
2. Center the draws at the posterior mean.
3. Multiply by $s_p$.
4. Re-center at the posterior mean.

This preserves the posterior mean while widening the credible interval
by a factor of $\sqrt{\text{DER}_p}$.

### Running the correct step

```{r step3-correct}
obj <- der_correct(obj)
```

### Verifying the correction

```{r step3-verify}
# Scale factors (1.0 for unflagged, > 1 for flagged)
cat("Scale factors for fixed effects:\n")
cat(sprintf("  beta[1] (intercept):    %.4f %s\n",
            obj$scale_factors[1],
            ifelse(obj$scale_factors[1] > 1, "(corrected)", "(unchanged)")))
cat(sprintf("  beta[2] (poverty_cwc):  %.4f %s\n",
            obj$scale_factors[2],
            ifelse(obj$scale_factors[2] > 1, "(corrected)", "(unchanged)")))
cat(sprintf("  beta[3] (tiered_reim):  %.4f %s\n",
            obj$scale_factors[3],
            ifelse(obj$scale_factors[3] > 1, "(corrected)", "(unchanged)")))

# Verify unflagged draws are identical
original   <- obj$original_draws
corrected  <- obj$corrected_draws
unchanged  <- !obj$classification$flagged

cat(sprintf("\nUnflagged parameters unchanged: %s\n",
            all(original[, unchanged] == corrected[, unchanged])))
```

### Computational cost

The selective correction operates on $|S|$ flagged parameters at cost
$O(|S| \times N_{\text{draws}})$. This is vastly cheaper than a blanket
correction that would require a full $O(d^3)$ Cholesky decomposition of
the $d \times d$ sandwich matrix, particularly when $d = p + J$ is
large.


## Putting It Together: Two Approaches

### Approach A: All-in-one with `der_diagnose()`

For most analyses, the all-in-one wrapper is the simplest entry point:

```{r approach-a}
result_a <- der_diagnose(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  psu         = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types,
  tau         = 1.2
)
```

### Approach B: Step-by-step pipeline

For more control, call each function individually. The pipe-friendly
design lets you inspect results at each stage:

```{r approach-b}
result_b <- der_compute(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  psu         = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types
)
result_b <- der_classify(result_b, tau = 1.2, verbose = FALSE)
result_b <- der_correct(result_b)
```

### Verifying equivalence

Both approaches produce identical results:

```{r verify-equivalent}
cat("DER values match:", all.equal(result_a$der, result_b$der), "\n")
cat("Classifications match:",
    all.equal(result_a$classification$flagged,
              result_b$classification$flagged), "\n")
cat("Corrected draws match:",
    all.equal(as.matrix(result_a), as.matrix(result_b)), "\n")
```

The step-by-step approach is preferred when you want to:

- Inspect intermediate sandwich matrices or per-group diagnostics.
- Try multiple threshold values without recomputing the sandwich.
- Skip correction and use DER only as a diagnostic.


## Sensitivity Analysis: How Robust Is the Classification?

The classification depends on the choice of threshold $\tau$. A natural
question is: how sensitive are the results to this choice? The
`der_sensitivity()` function evaluates the number of flagged parameters
across a range of $\tau$ values.

```{r sensitivity}
sens <- der_sensitivity(result_a, tau_range = seq(0.5, 3.0, by = 0.1))
sens[, c("tau", "n_flagged", "pct_flagged")]
```

### Visualizing threshold sensitivity

```{r fig-sensitivity, fig.height = 4.5, fig.cap = "Number of flagged parameters as a function of the threshold tau. The step pattern reveals that the classification is highly robust: only one parameter (poverty_cwc) is flagged across a wide range of thresholds. The vertical dashed lines mark commonly used threshold values."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  ggplot(sens, aes(x = tau, y = n_flagged)) +
    geom_step(colour = pal["tier_ia"], linewidth = 0.9) +
    geom_point(colour = pal["tier_ia"], size = 2) +
    geom_vline(xintercept = c(1.0, 1.2, 1.5, 2.0),
               linetype = "dashed", colour = "grey60", linewidth = 0.4) +
    annotate("text", x = c(1.0, 1.2, 1.5, 2.0),
             y = rep(max(sens$n_flagged) * 0.9, 4),
             label = c("1.0", "1.2\n(default)", "1.5", "2.0"),
             colour = "grey50", size = 3, vjust = -0.5) +
    scale_y_continuous(breaks = seq(0, max(sens$n_flagged) + 1, by = 1)) +
    labs(x = expression(paste("Threshold ", tau)),
         y = "Number of flagged parameters",
         title = "Sensitivity Analysis: Flagged Parameters vs Threshold",
         subtitle = sprintf("Total parameters: %d", length(result_a$der))) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
```

### Identifying the stable region

The "stable region" is the range of $\tau$ values over which the set of
flagged parameters does not change. In this example, the poverty
coefficient is flagged at every threshold below its DER value, and no
other parameters enter the flagged set. This stability reflects the sharp
separation between within-cluster and other parameter types in the NSECE.

```{r sensitivity-detail}
# At which thresholds does poverty_cwc enter/leave?
for (i in seq_len(nrow(sens))) {
  params_i <- sens$flagged_params[[i]]
  if (length(params_i) > 0) {
    cat(sprintf("tau = %.1f: flagged = {%s}\n",
                sens$tau[i], paste(params_i, collapse = ", ")))
  }
}
```

### When should you change $\tau$?

Guidelines for threshold selection:

- **$\tau = 1.0$**: Flag any parameter with DER above 1. Conservative ---
  corrects even small design effects. Appropriate when exact coverage
  calibration is paramount.
- **$\tau = 1.2$** (default): The posterior variance would need to
  increase by at least 20%. Balances sensitivity and specificity.
- **$\tau = 1.5$--$2.0$**: Only large design effects trigger correction.
  Appropriate when you want to preserve model-based inference and
  correct only the most egregious cases.

The sensitivity analysis helps you verify that your substantive
conclusions do not change across a reasonable range of thresholds.


## Cross-Clustering Comparison: `der_compare()`

The survey design may involve multiple levels of clustering. In the
NSECE, observations are clustered within PSUs (geographic areas) that
are themselves nested within states. The choice of clustering level for
the sandwich variance can affect the DER values.

The `der_compare()` function computes DER under different clustering
definitions and presents the results side by side:

```{r compare}
comp <- der_compare(
  nsece_demo$draws,
  clusters = list(
    state = nsece_demo$group,
    psu   = nsece_demo$psu
  ),
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types
)

# Show fixed effects comparison
fe_comp <- comp[grepl("^beta", comp$param), ]
fe_comp
```

### Visualizing cross-clustering results

```{r fig-compare, fig.height = 5, fig.cap = "DER comparison across two clustering definitions: state-level and PSU-level. Each facet shows one fixed-effect parameter. Finer clustering (PSU) captures more of the design effect, often producing larger DER values for within-cluster covariates."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  # Focus on fixed effects for clarity
  fe_comp$param_label <- factor(
    fe_comp$param,
    levels = unique(fe_comp$param),
    labels = c("Intercept", "Poverty (CWC)", "Tiered Reim.")
  )

  ggplot(fe_comp, aes(x = cluster_name, y = der, fill = cluster_name)) +
    geom_col(width = 0.6, colour = "white", linewidth = 0.3) +
    geom_hline(yintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.5) +
    annotate("text", x = 2.4, y = 1.25,
             label = expression(tau == 1.2),
             colour = pal["threshold"], size = 3.5, hjust = 0) +
    facet_wrap(~ param_label, scales = "free_y", nrow = 1) +
    scale_fill_manual(
      values = c("state" = pal["tier_ib"], "psu" = pal["tier_ia"]),
      name = "Clustering level"
    ) +
    labs(x = NULL, y = "DER",
         title = "DER by Clustering Level",
         subtitle = "State-level vs PSU-level clustering") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          strip.text = element_text(face = "bold"))
}
```

### Interpretation

The comparison reveals how the choice of clustering level affects
DER diagnostics:

- **Finer clustering** (PSU level): Captures more of the survey design's
  correlation structure, typically producing larger DER values for
  within-cluster covariates. This is the more conservative approach.
- **Coarser clustering** (state level): May underestimate the design
  effect if there is substantial within-state, between-PSU correlation.

**Practical recommendation**: Use the finest available clustering level
(PSU) for the primary analysis. The state-level comparison serves as a
diagnostic to assess sensitivity.


## Choosing the Right Threshold: Practical Guidance

The threshold $\tau$ trades off between two risks:

- **Too low** ($\tau < 1.0$): Flags parameters that are already
  adequately calibrated, leading to unnecessary corrections that
  may degrade coverage for well-estimated parameters.
- **Too high** ($\tau > 2.0$): Misses parameters with meaningful
  design effects, leaving posteriors that are too narrow.

### Comparing thresholds on NSECE data

Let us examine how different thresholds affect the classification and
resulting credible intervals:

```{r threshold-comparison}
tau_vals <- c(1.0, 1.2, 1.5, 2.0)

for (tau_i in tau_vals) {
  obj_i <- der_compute(
    nsece_demo$draws,
    y           = nsece_demo$y,
    X           = nsece_demo$X,
    group       = nsece_demo$group,
    weights     = nsece_demo$weights,
    psu         = nsece_demo$psu,
    family      = nsece_demo$family,
    sigma_theta = nsece_demo$sigma_theta,
    param_types = nsece_demo$param_types
  )
  obj_i <- der_classify(obj_i, tau = tau_i, verbose = FALSE)

  n_flag <- sum(obj_i$classification$flagged)
  flagged_names <- obj_i$classification$param_name[obj_i$classification$flagged]
  cat(sprintf("tau = %.1f: %d flagged", tau_i, n_flag))
  if (n_flag > 0 && n_flag <= 5) {
    cat(sprintf(" (%s)", paste(flagged_names, collapse = ", ")))
  }
  cat("\n")
}
```

### Coverage implications

From the simulation study in the accompanying paper, the default
$\tau = 1.2$ achieves the best balance between:

- **Marginal coverage for flagged parameters**: Corrected credible
  intervals achieve near-nominal (90%) frequentist coverage.
- **Marginal coverage for unflagged parameters**: Retained intervals
  maintain the variance gains from hierarchical shrinkage.
- **Overall calibration**: The average coverage across all parameters
  is close to nominal.

Using $\tau = 1.0$ would also achieve good coverage for flagged
parameters but risks over-correcting parameters with marginal design
effects.


## Working with Corrected Draws

The corrected posterior draws are the primary output of the pipeline.
They can be used for any downstream analysis that would normally use
MCMC draws.

### Extracting the draws matrix

```{r extract-draws}
draws_corrected <- as.matrix(result_a)
dim(draws_corrected)
```

The matrix has $S$ rows (draws) and $d$ columns (parameters), identical
in dimension to the original input. For unflagged parameters, the
columns are identical to the original draws.

### Credible interval comparison

The most immediate application is comparing credible intervals before
and after correction. For flagged parameters, the corrected interval is
wider by a factor of $\sqrt{\text{DER}}$:

```{r ci-comparison}
# Fixed-effect columns
p <- nsece_demo$p

original_draws  <- nsece_demo$draws[, 1:p]
corrected_draws <- draws_corrected[, 1:p]

# 90% credible intervals
ci_original  <- apply(original_draws, 2,
                      quantile, probs = c(0.05, 0.50, 0.95))
ci_corrected <- apply(corrected_draws, 2,
                      quantile, probs = c(0.05, 0.50, 0.95))

fe_names <- c("Intercept", "Poverty (CWC)", "Tiered Reim.")

cat("90% Credible Intervals: Original vs Corrected\n")
cat("----------------------------------------------\n")
for (k in seq_len(p)) {
  width_orig <- ci_original[3, k] - ci_original[1, k]
  width_corr <- ci_corrected[3, k] - ci_corrected[1, k]
  ratio <- width_corr / width_orig
  cat(sprintf("  %s:\n", fe_names[k]))
  cat(sprintf("    Original:  [%6.3f, %6.3f]  width = %.3f\n",
              ci_original[1, k], ci_original[3, k], width_orig))
  cat(sprintf("    Corrected: [%6.3f, %6.3f]  width = %.3f\n",
              ci_corrected[1, k], ci_corrected[3, k], width_corr))
  cat(sprintf("    Width ratio: %.3f\n\n", ratio))
}
```

Note that the width ratio for the flagged parameter equals
$\sqrt{\text{DER}} \approx \sqrt{2.6} \approx 1.61$, while unflagged
parameters have a ratio of exactly 1.000.

### Visualizing the correction effect

```{r fig-ci-comparison, fig.height = 4, fig.cap = "Naive (grey) versus corrected (coloured) 90% credible intervals for fixed effects. Only the poverty coefficient (Tier I-a, within-cluster) is widened by the DER correction. The intercept and tiered reimbursement (Tier I-b, between-cluster) are unchanged."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  ci_df <- data.frame(
    param    = rep(fe_names, 2),
    source   = rep(c("Original (naive)", "Corrected (DER)"), each = p),
    lower    = c(ci_original[1, ], ci_corrected[1, ]),
    median   = c(ci_original[2, ], ci_corrected[2, ]),
    upper    = c(ci_original[3, ], ci_corrected[3, ]),
    flagged  = rep(c(FALSE, TRUE, FALSE), 2),
    stringsAsFactors = FALSE
  )
  ci_df$param  <- factor(ci_df$param, levels = rev(fe_names))
  ci_df$source <- factor(ci_df$source,
                          levels = c("Corrected (DER)", "Original (naive)"))

  # Offset for visual separation
  ci_df$y_pos <- as.numeric(ci_df$param) +
    ifelse(ci_df$source == "Original (naive)", 0.12, -0.12)

  ggplot(ci_df, aes(y = y_pos)) +
    geom_vline(xintercept = 0, linetype = "dotted",
               colour = "grey60", linewidth = 0.4) +
    geom_segment(aes(x = lower, xend = upper,
                     colour = source),
                 linewidth = 1.5, lineend = "round") +
    geom_point(aes(x = median, colour = source), size = 2.5) +
    scale_colour_manual(
      values = c("Original (naive)" = "grey65",
                 "Corrected (DER)"  = pal["tier_ia"]),
      name = NULL
    ) +
    scale_y_continuous(breaks = seq_len(p), labels = rev(fe_names)) +
    labs(x = "Parameter estimate", y = NULL,
         title = "Credible Interval Comparison: Naive vs DER-Corrected",
         subtitle = "90% intervals; only the flagged parameter is widened") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.y = element_blank())
}
```

### Feeding corrected draws into downstream analysis

The corrected draws can be used anywhere that original MCMC draws would
be used. For example, to compute posterior predictive distributions or
derived quantities:

```{r downstream-example}
# Example: probability that the poverty effect is positive
prob_positive_orig <- mean(original_draws[, 2] > 0)
prob_positive_corr <- mean(corrected_draws[, 2] > 0)

cat(sprintf("P(beta_poverty > 0):\n"))
cat(sprintf("  Original:  %.4f\n", prob_positive_orig))
cat(sprintf("  Corrected: %.4f\n", prob_positive_corr))
cat("\nNote: the wider corrected posterior slightly changes tail\n")
cat("probabilities, reflecting the additional uncertainty from\n")
cat("the survey design.\n")
```

### Important: only flagged parameters change

This is a critical design principle of the selective correction. Random
effects and between-cluster fixed effects retain their original
posterior draws with no modification:

```{r verify-unchanged}
# Random effects are in columns (p+1) to (p+J)
J <- nsece_demo$J
re_original  <- nsece_demo$draws[, (p + 1):(p + J)]
re_corrected <- draws_corrected[, (p + 1):(p + J)]

# These should be bitwise identical
cat("Random effects unchanged:", identical(re_original, re_corrected), "\n")

# Fixed effects: only beta[2] changes
cat("beta[1] unchanged:", identical(nsece_demo$draws[, 1],
                                     draws_corrected[, 1]), "\n")
cat("beta[2] changed:",  !identical(nsece_demo$draws[, 2],
                                     draws_corrected[, 2]), "\n")
cat("beta[3] unchanged:", identical(nsece_demo$draws[, 3],
                                     draws_corrected[, 3]), "\n")
```

This is the key advantage of selective correction over blanket
correction: the hierarchical structure is preserved for the 53 of 54
parameters that are already well calibrated.


## The Built-In Diagnostic Plots

The `plot()` method provides three complementary views of the DER
diagnostics. Each uses `ggplot2` when available and falls back to base
R graphics otherwise.

### Profile plot

The profile plot displays DER values for each parameter, coloured by
tier, with the threshold as a reference line:

```{r fig-profile, fig.height = 4.5, fig.cap = "DER profile plot. Each dot represents one parameter, coloured by its tier classification. The dashed purple line marks the threshold (tau = 1.2). Only beta[2] (poverty_cwc, Tier I-a) exceeds the threshold."}
plot(result_a, type = "profile")
```

### Decomposition plot

The decomposition plot compares observed DER values against their
theoretical predictions from the decomposition theorems:

```{r fig-decomposition, fig.height = 4.5, fig.cap = "Decomposition diagnostic: observed DER values versus theoretical predictions from Theorems 1 and 2. Points near the 1:1 line indicate that the simplified decomposition formulas accurately approximate the exact sandwich computation."}
plot(result_a, type = "decomposition")
```

### Comparison plot

The comparison plot shows naive versus corrected credible intervals
for flagged parameters:

```{r fig-comparison-builtin, fig.height = 3.5, fig.cap = "Credible interval comparison for flagged parameters. The grey interval shows the naive (model-based) posterior, and the coloured interval shows the DER-corrected posterior. The correction widens the interval by a factor of sqrt(DER)."}
plot(result_a, type = "comparison")
```


## Advanced: Iterating Over Thresholds

For research applications, you may want to examine the full set of
corrected posteriors under different threshold values. Because the
compute step is the most expensive, you can reuse a single `svyder`
object and reclassify/correct with different thresholds:

```{r iterate-thresholds}
# Compute once
base_obj <- der_compute(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  psu         = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types
)

# Classify and correct at multiple thresholds
thresholds <- c(1.0, 1.2, 1.5, 2.0)
results_list <- lapply(thresholds, function(tau_i) {
  obj_i <- der_classify(base_obj, tau = tau_i, verbose = FALSE)
  obj_i <- der_correct(obj_i)
  obj_i
})
names(results_list) <- paste0("tau_", thresholds)

# Compare the 90% CI width for beta[2] across thresholds
cat("Poverty coefficient 90% CI width by threshold:\n")
for (i in seq_along(thresholds)) {
  draws_i <- as.matrix(results_list[[i]])[, 2]
  width_i <- diff(quantile(draws_i, c(0.05, 0.95)))
  cat(sprintf("  tau = %.1f: width = %.4f\n", thresholds[i], width_i))
}
```

This confirms that the correction for the poverty coefficient is
consistent across thresholds --- the parameter is flagged (and thus
corrected) at all four threshold values. The CI width is identical
across thresholds because the scale factor is determined by the DER
value, not the threshold.


## Broom-Style Output: `tidy()` and `glance()`

The svyder package provides tidy output methods following the conventions
of the broom package:

### tidy(): One row per parameter

```{r tidy-output}
td <- tidy.svyder(result_a)
head(td, 6)
```

The tidy output includes the posterior mean, standard deviation, DER
value, tier classification, action label, flagged indicator, and the
scale factor applied during correction.

### glance(): One-row model summary

```{r glance-output}
gl <- glance.svyder(result_a)
gl
```

The `glance()` summary provides a quick overview of the diagnostic
results: the number of parameters, number flagged, threshold used,
mean DER, and the DER range.


## Summary of the Pipeline

The compute-classify-correct pipeline provides a principled workflow for
adjusting Bayesian posteriors to respect complex survey designs:

| Step | Function | Input | Output | Key parameter |
|:-----|:---------|:------|:-------|:--------------|
| 1. Compute | `der_compute()` | Draws + survey metadata | DER values, sandwich matrices | `family`, `sigma_theta` |
| 2. Classify | `der_classify()` | svyder object | Tier assignments, flags | `tau` (threshold) |
| 3. Correct | `der_correct()` | svyder object (classified) | Corrected draws | `method` |
| All-in-one | `der_diagnose()` | Draws + survey metadata | Fully processed result | `tau`, `correct` |

The pipeline is designed to be:

- **Selective**: Only flagged parameters are corrected, preserving the
  hierarchical structure for the majority of parameters.
- **Transparent**: Every intermediate result is accessible for
  inspection.
- **Modular**: Each step can be run independently, and the threshold
  can be varied without recomputing the sandwich matrices.
- **Exact**: The DER values come from the full sandwich computation
  (Algorithm 1), not from the simplified decomposition formulas.


## What's Next?

This vignette covered the full diagnostic pipeline in detail. The
remaining vignettes provide additional context and theory:

| Vignette | What you will learn |
|:---------|:--------------------|
| *Getting Started* | Quick-start tutorial with the full svyder API |
| *Understanding Design Effect Ratios* | Mathematical foundations of DER, conservation law, and three-tier classification |
| *Decomposition Theorems* | Why DER differs across parameters: the four-factor decomposition and conservation law |
