---
title: "Decomposition Theorems and the Conservation Law"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  The mathematical centerpiece of svyder: under a balanced conjugate Gaussian
  model the Design Effect Ratio admits exact closed forms. This article states
  Theorems 1-2 and the conservation law, and verifies them numerically.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decomposition Theorems and the Conservation Law}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", out.width = "85%",
                      warning = FALSE, message = FALSE)
library(svyder)
# Consistent colour palette across all svyder vignettes (colour-blind aware)
svyder_pal <- c(
  `I-a`  = "#D6604D",  # within-group fixed effects (survey-dominated)
  `I-b`  = "#4393C3",  # between-group fixed effects (shielded)  [brand blue]
  `II`   = "#5AAE61",  # random effects (shrinkage-protected)
  `III`  = "#9E9E9E",  # hyperparameters (excluded)
  flag   = "#B2182B",  # flagged
  target = "#1A1A1A"   # DER = 1 / threshold reference lines
)
```

## Overview

The Design Effect Ratio is defined for any weighted pseudo-posterior, but for
one model class it stops being a black box. Under a **balanced conjugate
Gaussian hierarchical model**, every parameter's DER admits an **exact
finite-sample closed form** in terms of three quantities you can read off the
design and the fit: the Kish design effect $\mathrm{DEFF}$, the shrinkage factor
$B$, and, for fixed effects, the between-group share $R_k$ of that coefficient's
identifying variation. This article states those closed forms --- Theorem 1
(fixed effects), Theorem 2 (random effects), and the conservation law (Corollary
1) --- and then verifies them numerically on the bundled `nsece_demo` data.

This vignette is written for methodologists. The DER *definition* and the Kish
design effect it builds on are established (see
`vignette("theory-der")` and `vignette("theory-sandwich")`); the **closed-form
decomposition of the DER into a design-effect term modulated by hierarchical
shrinkage is the paper's novel contribution**. Where earlier work gives an
asymptotic sandwich, the results below are algebraic identities that hold at any
finite $J$ and $n$.

We cover, in order:

- **The conditions** (C1)-(C3) and the common-$\mathrm{DEFF}$ clause under which
  the closed forms hold.
- **Theorem 1** --- fixed-effect DERs live in a band $[\mathrm{DEFF}(1-B),\ \mathrm{DEFF}]$.
- **Theorem 2** --- random-effect DERs factor into design, shrinkage, and a
  finite-group correction.
- **The conservation law** --- design sensitivity is redistributed across the
  hierarchy, never eliminated.
- **Empirical verification** with `der_decompose()` and `der_theorem_check()`.
- **When the formulas diverge** from the software's exact computation.

## Conditions (C1)-(C3) and the common-DEFF clause

The theorems are stated for a specific, deliberately clean model. Let unit $i$
in group $j$ have outcome
$$y_{ij} \mid \theta_j, \boldsymbol{\beta} \sim \mathcal{N}(\mathbf{x}_{ij}^{\mathsf{T}}\boldsymbol{\beta} + \theta_j,\ \sigma_e^2), \qquad \theta_j \sim \mathcal{N}(0, \sigma_\theta^2).$$
The intercept sits in $\mathbf{x}_{ij}$; its coefficient, the grand mean, is
written $\mu$. The three conditions are:

- **(C1) Model class.** The conjugate Gaussian hierarchical model above, with
  **balanced groups** of common size $n$, a **flat prior** on
  $\boldsymbol{\beta}$, known variances, and non-collinear covariates.
- **(C2) Weights.** Strictly positive weights under the **global unit-mean
  convention** $\tilde{w}_{ij} = N\,w_{ij}/\sum w_{i'j'}$, with finite
  within-group design effects $\mathrm{DEFF}_j$.
- **(C3) Target.** The declared sandwich target of `vignette("theory-sandwich")`
  with the **model groups as the aggregation unit**.

Balance makes the shrinkage factor common across groups,
$$B_j \equiv B = \frac{\sigma_\theta^2}{\sigma_\theta^2 + \sigma_e^2/n},$$
but it does **not** make the design effects common. The clean $\mathrm{DEFF}$ in
Theorems 1 and 2 therefore requires one further hypothesis.

> **Note.** Theorems 1 and 2 state their headline results under the additional
> **common-$\mathrm{DEFF}$ clause** $\mathrm{DEFF}_j \equiv \mathrm{DEFF}$. This
> is a genuine assumption, separate from balance. Without it, the same algebra
> yields general **effective-meat forms** (distillation S-B.1, S-B.3) in which
> $\mathrm{DEFF}$ is replaced by an ensemble over $\{(B_j, \mathrm{DEFF}_j)\}$;
> we return to this in "When the formulas diverge."

> **Key insight.** These are **finite-sample algebraic identities**, not
> asymptotics in $J$ or $n$. Nothing below relies on a limit; the equalities hold
> exactly at, say, $J = 3$ groups of $n = 5$. That is what makes the DER
> decomposition a *design* statement rather than a large-sample approximation.

The shrinkage factor $B$ is the empirical-Bayes reliability ratio of Fay and
Herriot (1979): the weight a group's own data receives relative to the prior. As
$n \to \infty$, $B \to 1$ (data-rich groups trust themselves); as $n \to 0$,
$B \to 0$ (data-poor groups are pulled to the grand mean). Every result in this
article is ultimately a statement about how this one number reallocates the
design effect across the parameter vector.

## Theorem 1 (fixed effects)

**Statement.** Under (C1)-(C3) and $\mathrm{DEFF}_j \equiv \mathrm{DEFF}$,
$$\mathrm{DER}_{\beta_k} = \mathrm{DEFF}\,(1 - R_k), \qquad R_k \in [0, B],$$
where $R_k$ is the fraction of covariate $k$'s identifying variation that comes
from **between-group** differences,
$$R_k = \frac{[\mathbf{S}_\beta^{-1}\, \boldsymbol{\Delta}\, \mathbf{S}_\beta^{-1}]_{kk}}{[\mathbf{S}_\beta^{-1}]_{kk}}, \qquad \boldsymbol{\Delta} = \sum_j \frac{\tau}{(a_j + \tau)^2}\,\mathbf{b}_j\mathbf{b}_j^{\mathsf{T}} \succeq 0,$$
and $B = \sigma_\theta^2/(\sigma_\theta^2 + \sigma_e^2/n)$ is the common
shrinkage factor. ($\mathbf{S}_\beta$ is the Schur-complement precision for
$\boldsymbol{\beta}$; $\mathbf{b}_j$, $a_j$, and $\tau = 1/\sigma_\theta^2$ are
as in `vignette("theory-sandwich")`.)

Read $R_k$ as a dial from 0 to $B$. The design effect $\mathrm{DEFF}$ is the raw
inflation from unequal weighting; the factor $(1 - R_k)$ is how much of it a
given coefficient actually inherits, and that inheritance shrinks precisely to
the extent the coefficient leans on between-group contrasts. Two endpoints make
the mechanism concrete.

**Special case: pure within-group covariate ($R_k = 0$).**
$$\mathrm{DER}_{\beta_k} = \mathrm{DEFF}.$$
A covariate that varies only *within* groups is identified entirely from
same-group comparisons. Those comparisons sit inside the very clusters the design
correlated, and the prior on $\boldsymbol{\theta}$ acts at the group level, so it
**cannot absorb** the intra-group design-induced correlation. Such a coefficient
is **fully exposed** and inherits the whole design effect. These are the
**Tier I-a** parameters --- the primary correction candidates.

**Special case: pure between-group covariate ($R_k = B$).**
$$\mathrm{DER}_{\beta_k} = \mathrm{DEFF}\,(1 - B).$$
A covariate that varies only *between* groups (or is group-level) is identified
from inter-group contrasts, and that between-group information is exactly what
hierarchical shrinkage diffuses into the prior. The coefficient is **shielded**
by the factor $(1 - B)$: heavily shrunk models (large $B$) protect it most.
These are the **Tier I-b** parameters, typically well below any sensible
threshold even when $\mathrm{DEFF}$ is large.

> **Key insight.** Fixed-effect DERs span the **band $[\mathrm{DEFF}(1-B),\ \mathrm{DEFF}]$**,
> and where a coefficient lands is a property of *its covariate*, not of the
> design alone. Two coefficients in the same model, fit to the same data with the
> same $\mathrm{DEFF}$, can differ by a factor of $1/(1-B)$ purely because one is
> identified within groups and the other between them.

For `nsece_demo` the band endpoints are easy to compute. With the Kish global
design effect $\mathrm{DEFF} = `r 2.728`$ and shrinkage
$B = `r 0.854`$ (the fitted `mean_B`), the band runs from

```{r band-endpoints}
DEFF <- 2.728   # Kish global design effect for nsece_demo (deff_used)
B    <- 0.854   # fitted shrinkage factor (glance: mean_B)
c(lower_DEFF_1minusB = round(DEFF * (1 - B), 3),
  upper_DEFF         = round(DEFF, 3))
```

so a within-group coefficient should sit near `r round(2.728, 2)` and a
between-group coefficient near `r round(2.728 * (1 - 0.854), 2)`. We confirm both
in the verification section.

## Theorem 2 (random effects)

Random effects are identified from *both* layers --- within groups through the
error variance, between groups through the prior --- so their DER carries the
shrinkage factor explicitly.

**Statement (a): conditional on the grand mean.** Under (C1)-(C3) and
$\mathrm{DEFF}_j \equiv \mathrm{DEFF}$,
$$\mathrm{DER}_\theta^{\mathrm{cond}} = B \cdot \mathrm{DEFF}.$$

**Statement (b): marginal, at finite $J$.**
$$\mathrm{DER}_\theta = B \cdot \mathrm{DEFF} \cdot \kappa(J), \qquad \kappa(J) = 1 - \frac{1}{J(1 - B) + B}.$$

The conditional form is the clean one: a group effect inherits the design effect
scaled by $B$. When groups are data-rich ($B \to 1$) they behave like ordinary
fixed effects and $\mathrm{DER}_\theta^{\mathrm{cond}} \to \mathrm{DEFF}$; when
groups are data-poor ($B \to 0$) shrinkage dominates and
$\mathrm{DER}_\theta^{\mathrm{cond}} \to 0$. The marginal form multiplies this by
the **finite-group correction** $\kappa(J)$, which arises because the marginal
posterior of $\theta_j$ integrates over the grand mean $\mu$, coupling the $J$
groups together.

The correction $\kappa(J)$ is a pure function of $J$ and $B$. It rises
monotonically to 1 as the number of groups grows, and is materially below 1 only
when $J$ is tiny:

```{r kappa-values}
kappa <- function(J, B) 1 - 1 / (J * (1 - B) + B)
data.frame(
  J          = c(3, 5, 10, 20, 50, 100),
  kappa_B0.5 = round(kappa(c(3, 5, 10, 20, 50, 100), B = 0.5), 3)
)
```

With $J = 3$ groups and $B = 0.5$, $\kappa = `r round(1 - 1/(3*(1-0.5)+0.5), 2)`$
--- a large downward correction; by $J = 50$ it is
$`r round(1 - 1/(50*(1-0.5)+0.5), 3)`$, essentially negligible. For the 51-group
demo, $\kappa$ is close enough to 1 that the conditional and marginal forms
nearly coincide.

> **Key insight.** The **shrinkage-design tradeoff**: the parameters most
> protected by shrinkage (small $B$) are the *least* exposed to the design. This
> is the opposite of the usual worry. Applying a design correction to a
> heavily-shrunk random effect does not fix an understated interval --- there is
> little design sensitivity there to begin with --- it overwrites the shrinkage
> that was doing the work. Under a model-group target, random effects are
> **Tier II** and rarely need correction.

## The conservation law (Corollary 1)

The two theorems are not independent. Combining Theorem 1 at the grand mean
(which is pure between-group, so $R_\mu = B$ and
$\mathrm{DER}_\mu = \mathrm{DEFF}(1 - B)$) with the conditional random-effect
result yields an exact identity.

**Statement.** Under (C1)-(C3),
$$\mathrm{DER}_\mu + \mathrm{DER}_\theta^{\mathrm{cond}} = \mathrm{DEFF}.$$

Read $\mathrm{DEFF}$ as a fixed **design-sensitivity budget** --- the total
variance inflation that unequal weighting injects into the hierarchy. The
conservation law says that budget is **redistributed, not eliminated**: shrinkage
moves it off the group effects and onto the grand mean rather than making it
disappear. The share $\mathrm{DEFF}(1 - B)$ accumulates at the grand mean; the
share $B \cdot \mathrm{DEFF}$ is absorbed by the group effects. As $B$ swings from
0 to 1, the budget slides continuously from the grand mean to the group effects,
but the two shares always sum to $\mathrm{DEFF}$.

```{r conservation-plot, echo = TRUE, fig.cap = "How the fixed design-sensitivity budget DEFF splits between the grand mean (blue) and the group effects (green) as the shrinkage factor B varies. The two shares always sum to DEFF (dashed line); shrinkage redistributes design sensitivity across the hierarchy rather than removing it."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  DEFF <- 2.728
  budget <- data.frame(B = seq(0, 1, by = 0.01))
  budget$grand_mean   <- DEFF * (1 - budget$B)   # DER_mu
  budget$group_effect <- DEFF * budget$B         # DER_theta^cond

  ggplot(budget, aes(x = B)) +
    geom_ribbon(aes(ymin = 0, ymax = group_effect),
                fill = svyder_pal[["II"]], alpha = 0.5) +
    geom_ribbon(aes(ymin = group_effect, ymax = DEFF),
                fill = svyder_pal[["I-b"]], alpha = 0.5) +
    geom_line(aes(y = group_effect), colour = svyder_pal[["II"]], linewidth = 1) +
    geom_hline(yintercept = DEFF, linetype = "dashed",
               colour = svyder_pal[["target"]]) +
    annotate("text", x = 0.5, y = DEFF * 0.20, label = "group effects\nB . DEFF",
             colour = svyder_pal[["II"]], fontface = "bold", size = 3.4) +
    annotate("text", x = 0.5, y = DEFF * 0.80, label = "grand mean\nDEFF(1 - B)",
             colour = svyder_pal[["I-b"]], fontface = "bold", size = 3.4) +
    annotate("text", x = 0.02, y = DEFF + 0.12, label = "DEFF (total budget)",
             hjust = 0, colour = svyder_pal[["target"]], size = 3.2) +
    labs(x = "shrinkage factor  B", y = expression(DER),
         title = "Corollary 1: the design-sensitivity budget is redistributed, not eliminated") +
    coord_cartesian(ylim = c(0, DEFF + 0.35)) +
    theme_minimal(base_size = 11)
}
```

At $B = 0$ the whole budget sits on the grand mean; at $B = 1$ it has all moved
to the group effects; everywhere in between the split is exactly
$\mathrm{DEFF}(1-B) : B\,\mathrm{DEFF}$.

> **Note.** The conservation law is proven for the balanced conjugate Gaussian
> class under common $\mathrm{DEFF}$. Whether an analogous identity survives to
> unbalanced, non-Gaussian, or deeper hierarchies is **open** (paper, Discussion).

## Empirical verification

We now check the closed forms against the software's exact sandwich computation
on `nsece_demo` --- binomial, $N = 6785$, $J = 51$ states, $\mathrm{DEFF} = 2.728$
--- with the **design-PSU target** (`cluster = nsece_demo$psu`). The demo is
binomial rather than Gaussian, so the theorems apply only as a *linearized
approximation*; the point is to see the mechanism reproduce, not to claim
machine-precision equality.

```{r compute}
result <- der_compute(
  nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  cluster     = nsece_demo$psu,          # design-based target; NEVER psu =
  family      = "binomial",
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types
)
```

### `der_decompose()`: the closed-form ingredients

`der_decompose()` returns, for every parameter, the design effect used, the
shrinkage $B$, the between-group share $R_k$, the finite-group factor $\kappa$,
and the resulting **predicted** DER --- alongside the empirically computed `der`.

```{r decompose}
dc <- der_decompose(result)
head(dc, 6)
```

Look first at the three fixed effects. The within-group covariate is
**`beta[2]`** (`poverty_cwc`; in package output the poverty covariate is column
2, not column 1):

```{r decompose-fe}
dc[dc$param_type != "re",
   c("param", "param_type", "der", "deff_used", "R_k", "der_predicted")]
```

This is Theorem 1 working exactly as advertised:

- **`beta[2]` (within, Tier I-a):** $R_k \approx `r round(0.0143, 3)`$ --- almost
  no between-group content, so $\mathrm{DER}_{\beta_k} \approx \mathrm{DEFF}$.
  The empirical `der` is **2.689** and the predicted value is **2.689**: a
  within-group coefficient inherits essentially the whole design effect.
- **`beta[1]` and `beta[3]` (between, Tier I-b):** $R_k \approx 0.93$ --- almost
  all between-group content, so both are pulled down into the shielded end of the
  band, with `der` 0.263 and 0.343 respectively. These are *below 1*: correcting
  them would wrongly narrow already-calibrated intervals.

The `beta[2]` agreement is the cleanest demonstration in the package that a pure
within-group coefficient sits at $R_k \approx 0$ and inherits the full
$\mathrm{DEFF}$.

> **Note.** `der_decompose()` reports `der_predicted` using each parameter's own
> Kish `deff_used` (here 2.728). The companion `der_theorem_check()` below scores
> Theorem 1 against the *mean* design effect (2.595, the fitted `mean_deff`), so
> its `der_theorem1` for `beta[2]` reads about 2.595 rather than 2.689. Both are
> legitimate; they differ only in which $\mathrm{DEFF}$ estimate stands in for the
> common value the theorem assumes.

For the random effects, `der_decompose()` populates $B$ and $\kappa$ (and leaves
$R_k$ as `NA`, since $R_k$ is a fixed-effect concept):

```{r decompose-re}
re <- dc[dc$param_type == "re", ]
head(re[, c("param", "der", "deff_used", "B_used", "kappa", "der_predicted")], 4)
```

Here $B \approx 0.58$-$0.64$ and $\kappa \approx 0.95$, matching Theorem 2:
`theta[1]`, for instance, predicts $B \cdot \mathrm{DEFF} \cdot \kappa \approx 2.12$.
But note the empirical `der` for `theta[1]` is 3.386 against a predicted 2.122 ---
a large gap. That is expected, and it is a *data* limitation, not a theory
failure.

> **Key insight.** This example is using the **design-PSU target**, so the
> random-effect gaps are best read as linearized-theory approximation error:
> the demo is binomial, groups are unbalanced, and the design effects are
> heterogeneous across states. Read the random-effect results **tier-wise** ---
> the median RE DER is `r round(1.311, 3)`, comfortably in the Tier II range ---
> not state-by-state.
>
> Under a **model-group** target there is an additional estimator limitation:
> each state contributes only one centered score-total contrast to its own meat
> entry. That one-contrast issue belongs to the model-group target, not to the
> design-PSU computation shown here.

### `der_theorem_check()`: scoring the closed forms

`der_theorem_check()` puts the empirical DER next to the theorem-predicted value
and reports which theorem was used and the relative error.

```{r theorem-check}
tc <- der_theorem_check(result)
head(tc, 6)
```

The fixed-effect rows confirm the pattern: `beta[2]` (within) is scored against
Theorem 1's within branch with small relative error, while the between effects
`beta[1]`/`beta[3]` are scored against the between branch. The random-effect rows
use Theorem 2 and, per the caveat above, show larger per-state relative errors
because the design-PSU demo is a finite, unbalanced, binomial approximation to
the balanced Gaussian theorem --- `theta[2]` is the worst offender in the first
few rows.

The **conservation law** is attached as an attribute, evaluated on the fitted
quantities:

```{r conservation-check}
cl <- attr(tc, "conservation_law")
cl
```

The identity to check is
$\mathrm{DER}_\mu + \mathrm{DER}_\theta^{\mathrm{cond}} = \mathrm{DEFF}$. The
attribute reports `der_mu` (the grand-mean DER, here the `beta[1]` intercept),
`der_theta_cond` (the conditional theorem term, computed from $B_j
\mathrm{DEFF}_j$), their `conservation_sum`, and `deff_mean` for comparison. It
also reports `der_theta_marginal_mean`, the empirical mean of the fitted
random-effect DERs. The marginal mean is useful as an empirical diagnostic, but
it is **not** the conditional term in the conservation identity.

On this binomial demo the conditional sum and $\mathrm{DEFF}$ agree only
approximately --- the `relative_error` field quantifies the gap --- because the
model is not Gaussian, the groups are unbalanced, and the design effects are
heterogeneous. The identity is *exact* only in the balanced Gaussian world the
theorem inhabits; the demo shows the mechanism, not machine-precision equality.

## When the formulas diverge

The closed forms are exact under (C1)-(C3) **and** the common-$\mathrm{DEFF}$
clause. Two departures, both present in real applications and partly in the demo,
move the software's exact sandwich away from the simple predictions:

- **Heterogeneous design effects ($\mathrm{DEFF}_j$ not common).** Balance forces
  a common $B$ but never a common $\mathrm{DEFF}$. When the within-group design
  effects differ, Theorem 1's $\mathrm{DEFF}$ is replaced by
  $\mathbf{M} = \sum_j \mathrm{DEFF}_j\, \mathbf{C}_j^{(d)}$ with
  $\mathbf{C}_j^{(d)} = \mathbf{W}_j + \tfrac{(1 - B_j)^2}{a_j}\mathbf{b}_j\mathbf{b}_j^{\mathsf{T}}$
  (distillation S-B.1), and the random-effect DER depends on the whole ensemble
  $\{(B_j, \mathrm{DEFF}_j)\}$ rather than a naive per-group product
  (distillation S-B.3).
- **Unbalanced groups ($n_j$ not common).** Then $B_j$ varies too, and the single
  band $[\mathrm{DEFF}(1-B),\ \mathrm{DEFF}]$ becomes a family of bands. The
  ensemble-exact expressions still hold; the two-number summary does not.

The software **always computes the exact sandwich**, so `der_decompose()` and
`der_theorem_check()` remain correct diagnostics regardless. The closed forms are
best read as the *interpretive skeleton* --- they tell you which direction each
tier moves and why --- with the ensemble forms of distillation S-B.1-S-B.3 as the
exact statement when balance or common $\mathrm{DEFF}$ fails.

## What's next?

- To see how these tiers turn into a **classification and a selective
  correction** --- the four-tier table, the threshold $c_0$, and block-Cholesky
  CCC --- read `vignette("theory-classification-correction")`.
- For the **bread and meat** behind the sandwich the theorems decompose ---
  penalized Hessian, cluster-aggregated meat, weight conventions --- see
  `vignette("theory-sandwich")`.
- For the DER definition and the three regimes that motivate all of this, see
  `vignette("theory-der")`.
- To watch the decomposition play out on a full worked example, see the applied
  [Case Study](case-study-nsece.html).

### References

- Lee, J., Williams, M. R., & Savitsky, T. D. (2026). *Design Effect Ratios for
  Bayesian Survey Models: A Diagnostic Framework for Identifying Survey-Sensitive
  Parameters.* Journal of Survey Statistics and Methodology (submitted).
- Fay, R. E., & Herriot, R. A. (1979). Estimates of income for small places: An
  application of James-Stein procedures to census data. *Journal of the American
  Statistical Association, 74*(366), 269-277.
- Kish, L. (1965). *Survey Sampling.* Wiley.
