---
title: "The Design Effect Ratio: Definition and Regimes"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A methodological introduction to the parameter-level Design Effect Ratio (DER):
  how it generalises the Kish design effect to Bayesian hierarchical models, the
  three regimes it distinguishes, and the information-source intuition behind why
  parameters inherit design effects so differently.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Design Effect Ratio: Definition and Regimes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", out.width = "85%",
                      warning = FALSE, message = FALSE)
library(svyder)
# Consistent colour palette across all svyder vignettes (colour-blind aware)
svyder_pal <- c(
  `I-a`  = "#D6604D",  # within-group fixed effects (survey-dominated)
  `I-b`  = "#4393C3",  # between-group fixed effects (shielded)  [brand blue]
  `II`   = "#5AAE61",  # random effects (shrinkage-protected)
  `III`  = "#9E9E9E",  # hyperparameters (excluded)
  flag   = "#B2182B",  # flagged
  target = "#1A1A1A"   # DER = 1 / threshold reference lines
)
```

## Overview

This vignette is written for methodologists. It defines the **Design Effect
Ratio (DER)**, the per-parameter diagnostic that `svyder` computes, and explains
the three regimes it distinguishes and the information-source reasoning that
governs them. We stay at the level of *definition and intuition*: we motivate the
quantity, state what it measures, and interpret its regimes. The explicit
construction of its numerator (the declared variance target) is deferred to
`vignette("theory-sandwich")`, and the closed-form decomposition theorems that
predict its value are deferred to `vignette("theory-decomposition")`.

It is worth being precise about what is old and what is new here. The **Kish
design effect** (Kish, 1965), the **sandwich / linearisation variance** for
complex surveys (Binder, 1983), the **pseudo-posterior** for informative
sampling (Savitsky & Williams, 2022), and the **design-based covariance
calibration** of pseudo-Bayesian draws (Williams & Savitsky, 2021) are all
established. What is
*novel* in this work is (i) turning the design effect into a **per-parameter
ratio** against the model posterior variance, (ii) showing that this ratio
decomposes into a design-effect factor modulated by hierarchical shrinkage, and
(iii) using it to correct *selectively* --- only the parameters that need it.

We cover, in order:

- **From Kish DEFF to DER** --- why one global scalar is not enough for a
  hierarchical model, and the formal definition of the ratio.
- **The three regimes** --- what $\mathrm{DER}_k > 1$, $\approx 1$, and $< 1$ each mean,
  and why the third one is the reason to correct selectively.
- **Why parameters differ** --- the information-source argument: within-group
  variation is exposed to the design; between-group variation is shielded by
  shrinkage.
- **The declared variance target, in brief** --- one paragraph on the numerator,
  with the construction deferred.
- **A first look at the numbers** --- DER computed live on the bundled data.
- **Notation** --- a compact table matching the paper.

## From Kish DEFF to DER

The classical **Kish design effect** summarises, in a single number, how much
unequal weighting inflates the variance of an estimator relative to a
self-weighting sample of the same size:

$$
\mathrm{DEFF} \;=\; \frac{N \sum_{ij} w_{ij}^{2}}{\left(\sum_{ij} w_{ij}\right)^{2}},
$$

where $w_{ij}$ is the design weight for unit $i$ in group $j$ and $N$ is the
sample size (Kish, 1965). By construction $\mathrm{DEFF} \ge 1$, with equality
under equal weights. It is a property of the *weights*, not of any particular
parameter.

That scalar view is exactly what breaks down for a Bayesian hierarchical model.
A hierarchical fit produces many parameters at once --- a grand mean, regression
coefficients that may vary within or between groups, and a whole vector of
random effects --- and these do **not** inherit the design effect uniformly. A
coefficient identified purely from within-group contrasts feels the full force of
the design-induced correlation among same-group units; a coefficient identified
from between-group contrasts is largely shielded by the shrinkage prior; the
random effects sit somewhere in between. A single $\mathrm{DEFF}$ cannot express
this --- it has one value, but the model has $d = p + J$ directions of uncertainty
that respond differently.

> **Key insight.** The Kish $\mathrm{DEFF}$ is a scalar summary of the *weights*.
> A hierarchical model has many parameters that inherit design uncertainty at
> different rates. The DER promotes the design effect to a *vector* --- one value
> per parameter --- so the diagnostic can distinguish them.

The DER makes this per-parameter. For each parameter $k$, compare the variance
its posterior *should* carry if it properly reflected the design (the declared
variance target $V_{\mathrm{target}}$) against the variance its MCMC posterior
*does* carry ($\Sigma_{\mathrm{MCMC}}$):

**Definition 1 (Design Effect Ratio).**
$$
\mathrm{DER}_k \;=\; \frac{[V_{\mathrm{target}}]_{kk}}{[\Sigma_{\mathrm{MCMC}}]_{kk}}.
$$

The numerator $V_{\mathrm{target}}$ is a design-based *sandwich* covariance
evaluated at the posterior mean $\hat{\boldsymbol{\phi}}$; the denominator is the
empirical covariance of the pseudo-posterior draws. Both are $d \times d$
matrices, and the DER reads off their *diagonal* entries, parameter by
parameter. The ratio is population-indexed and post-fit: it diagnoses a fitted
posterior, it does not change the fit.

> **Note.** The DER uses only diagonal entries of the two covariance matrices,
> so it is well defined even when $V_{\mathrm{target}}$ is rank-deficient (which
> happens whenever the number of clusters is smaller than the number of
> parameters). The off-diagonal structure matters only at the *correction* step;
> see `vignette("theory-classification-correction")`.

## The three regimes

Reading the ratio against $1$ partitions the parameters into three regimes. The
number $1$ is the natural reference: it is the point at which the posterior
already carries exactly the variance the design demands.

- **$\mathrm{DER}_k > 1$ --- the posterior understates the target.** The model-based
  interval is too *narrow* relative to the design-based target; the design
  injects uncertainty the pseudo-posterior has not absorbed. The corrective
  action is to **widen** this parameter's interval. These are the parameters a
  correction is *meant* to touch.

- **$\mathrm{DER}_k \approx 1$ --- agreement.** The posterior variance already
  matches the target to within a tolerance. **Leave it alone**; correcting it
  would only add noise.

- **$\mathrm{DER}_k < 1$ --- the posterior is already wider than the target.** Here a
  blanket correction would *narrow* an interval that is legitimately wide. That
  width is not an error to be removed --- it is **hierarchical-shrinkage
  protection**, the pooling that stabilises weakly-identified group effects. A
  correction that "fixed" it would destroy exactly the protection the
  hierarchical prior was there to provide.

> **Key insight.** The whole case for *selective* correction lives in the third
> regime. A correction applied to every parameter cannot tell $\mathrm{DER} > 1$
> (needs widening) from $\mathrm{DER} < 1$ (needs to be left alone) --- it rescales
> both, and in doing so it collapses the intervals of the shrinkage-protected
> parameters. The DER is the switch that turns correction on only where
> $\mathrm{DER}_k$ exceeds a threshold $c_0$.

The threshold is a tolerance around the "$\approx 1$" region: `svyder` uses
$c_0 = 1.2$ by default (a 20% margin), so a parameter is flagged for correction
only when $\mathrm{DER}_k > 1.2$. The rationale for that value, and its
robustness, belong to `vignette("theory-classification-correction")`.

The schematic below places a few representative DERs on the number line. Two
sit below $1$ (protected --- leave them), one straddles the tolerance band, and
one sits well above the threshold (flag and widen).

```{r regime-schematic, echo = FALSE, fig.height = 2.6, fig.cap = "The three DER regimes on a number line. The solid line marks DER = 1 (posterior matches target); the dashed line marks the default threshold c0 = 1.2. Points left of 1 are shrinkage-protected (leave alone); points above the threshold are flagged for widening."}
op <- par(mar = c(4, 1, 2, 1))
xr <- c(0, 6)
plot(NA, xlim = xr, ylim = c(-1, 1), axes = FALSE, xlab = "", ylab = "")
axis(1, at = 0:6)
mtext(expression(DER[k]), side = 1, line = 2.4)
# reference lines
abline(v = 1,   col = svyder_pal["target"], lwd = 2)
abline(v = 1.2, col = svyder_pal["flag"],   lwd = 1.5, lty = 2)
text(1,   0.9, "DER = 1", pos = 2, col = svyder_pal["target"], cex = 0.85, xpd = NA)
text(1.2, 0.9, expression(c[0] == 1.2), pos = 4, col = svyder_pal["flag"], cex = 0.85, xpd = NA)
# example points (approx. nsece_demo values: intercept, tiered_reim, an RE, poverty)
ex <- c(0.26, 0.34, 1.31, 2.69)
cols <- c(svyder_pal["I-b"], svyder_pal["I-b"], svyder_pal["II"], svyder_pal["I-a"])
labs <- c("between FE", "between FE", "random effect", "within FE (flagged)")
points(ex, rep(0, 4), pch = 19, cex = 1.8, col = cols)
text(ex, rep(-0.55, 4), labs, cex = 0.72, srt = 0, col = cols, xpd = NA)
par(op)
```

## Why parameters differ: information source

Why should two coefficients in the *same* model inherit the design effect at such
different rates? The answer is where their identifying information comes from.

Consider a design that oversamples some clusters and undersamples others, inducing
correlation among units in the same cluster. A **within-group covariate** --- one
that varies among the units of a cluster --- is identified entirely from
*intra-cluster* contrasts. Those contrasts are precisely the ones the design
distorts, and the random-effect prior on $\theta_j$ operates at the *group* level,
so it cannot absorb within-group correlation. The within-group coefficient is
therefore **fully exposed** to the design effect.

A **between-group covariate** --- one that is constant within each cluster and
varies only across clusters --- is identified from *inter-cluster* contrasts. But
those between-group differences are exactly what the shrinkage prior pools: the
prior treats each group's mean as a draw around a common centre, so between-group
information is *diffuse* in the group-level prior. The between-group coefficient is
therefore **shielded** by shrinkage, and its DER comes in well below the naive
design effect.

> **Key insight.** Within-group variation is fully exposed to design-induced
> intra-cluster correlation --- the $\theta_j$ prior cannot reach it. Between-group
> variation is shielded, because that is the variation the shrinkage prior pools.
> Same design, same $\mathrm{DEFF}$; opposite exposure.

This split is captured by a single quantity, the **between-group share** $R_k$ of
covariate $k$'s identifying variation. A pure within-group covariate has
$R_k = 0$ (nothing shielded); a pure between-group covariate has $R_k = B$, the
maximum the shrinkage factor allows. `svyder` reports $R_k$ in `der_decompose()`,
and `vignette("theory-decomposition")` proves that the fixed-effect DER is
exactly $\mathrm{DEFF}\,(1 - R_k)$ --- so $R_k$ is the dial that slides a
coefficient's DER between $\mathrm{DEFF}(1-B)$ (fully shielded) and $\mathrm{DEFF}$
(fully exposed). For now, read $R_k$ as *the fraction of a covariate's signal
that lives between groups, and is therefore protected*.

## The declared variance target, in brief

The numerator of Definition 1 is the **declared variance target**, a design-based
sandwich (Binder, 1983; Savitsky & Williams, 2022):

$$
V_{\mathrm{target}} \;=\; H^{-1}\, J_c\, H^{-1},
$$

with a **bread** $H$ (the penalised posterior Hessian --- the curvature of the
objective the sampler actually explores, including the prior precision on the
random effects) and a **meat** $J_c$ (a cluster-aggregated, stratum-centred outer
product of the weighted score totals, with a finite-cluster correction). The word
*declared* is deliberate: the analyst declares the **aggregation unit** (design
PSUs versus model groups) and the **weight convention**, and those choices are
part of the estimand --- the DER changes materially between them. The full
construction, the reasons the *posterior* Hessian is the right bread, and the
gaussian score fix are the subject of `vignette("theory-sandwich")`; we do not
reconstruct them here.

## A first look at the numbers

The cleanest way to see that the DER is *not* one global number is to compute it
where the design effect is switched off. The bundled `sim_hlr` dataset is a
balanced gaussian hierarchical model with **equal weights**, so its Kish design
effect is exactly $\mathrm{DEFF} = 1$. If the DER were merely a relabelled
$\mathrm{DEFF}$, every parameter would sit at $1$. It does not: with the design
effect removed, the DERs isolate the *within/between + shrinkage* structure alone.

```{r sim-hlr, message = TRUE}
der_sim <- der_compute(
  sim_hlr$draws,
  y            = sim_hlr$y,
  X            = sim_hlr$X,
  group        = sim_hlr$group,
  weights      = sim_hlr$weights,
  cluster      = sim_hlr$group,          # model-group target
  family       = "gaussian",
  sigma_theta  = 0.5,
  sigma_e      = 1.0,
  param_types  = c("fe_between", "fe_within")
)
der_sim <- der_classify(der_sim, verbose = FALSE)

td  <- tidy(der_sim)
gl  <- glance(der_sim)

# the two fixed effects
td[td$term %in% c("beta[1]", "beta[2]"), c("term", "der", "tier", "flagged")]
```

```{r sim-hlr-summary}
# whole-model summary: DEFF = 1, yet the DERs span a wide range
c(deff       = round(gl$mean_deff, 3),
  der_min    = round(gl$der_min, 3),
  der_max    = round(gl$der_max, 3),
  n_flagged  = gl$n_flagged,
  n_params   = gl$n_params)
```

The mean design effect is `r round(gl$mean_deff, 3)`, exactly $1$, yet the DERs
run from about `r round(gl$der_min, 3)` to about `r round(gl$der_max, 3)`. The
structure is precisely what the theory predicts with $\mathrm{DEFF} = 1$:

- the **within-group** coefficient `beta[2]` sits at $\mathrm{DER} \approx 1.31$
  --- near $\mathrm{DEFF} = 1$, the fully-exposed value (it is a small sample, so it
  overshoots slightly);
- the **between-group** intercept `beta[1]` sits at $\mathrm{DER} \approx 0.10$
  --- near $\mathrm{DEFF}\,(1 - B) \approx 0.17$, the fully-shielded value ($B \approx 0.83$ here);
- the **random effects** all sit **below $1$** (their DERs run roughly $[0.05, 0.85]$),
  reflecting shrinkage protection.

Only `r gl$n_flagged` of `r gl$n_params` parameters is flagged. This is the honest
reading: with $\mathrm{DEFF} = 1$ there is *no design inflation to inherit*, so the
DERs $\ne 1$ trace out the shrinkage geometry alone, up to finite-sample noise
($J = 10$ groups is small). One should **not** say "all DERs equal $1$" --- they do
not, and the point of the example is exactly that they do not.

> **Note.** Contrast this with a real design. The bundled `nsece_demo`
> (a synthetic NSECE-style survey, $\mathrm{DEFF} = 2.728$, computed under the
> design-PSU target `cluster = nsece_demo$psu`) produces a DER range of
> $[0.235, 5.308]$ with 30 of 54 parameters flagged: the within-state poverty
> coefficient `beta[2]` reaches $2.689$ while the two between-state coefficients
> sit at $0.263$ and $0.343$. Now the design effect *is* present, and the DERs
> spread across both regimes. The applied
> [Getting Started](getting-started.html) vignette walks through that diagnostic
> end to end.

## Notation

The symbols used throughout the method vignettes, matching the paper:

| Symbol | Meaning |
|:---|:---|
| $\mathrm{DER}_k$ | Design Effect Ratio for parameter $k$: $[V_{\mathrm{target}}]_{kk} / [\Sigma_{\mathrm{MCMC}}]_{kk}$ |
| $V_{\mathrm{target}}$ | Declared design-based sandwich covariance target, $H^{-1} J_c H^{-1}$ ($d \times d$) |
| $\Sigma_{\mathrm{MCMC}}$ | Posterior covariance of the pseudo-posterior MCMC draws ($d \times d$) |
| $H$ | Penalised (posterior) Hessian: weighted pseudo-likelihood curvature $+\,\mathrm{diag}(\mathbf 0_p, \tau, \dots, \tau)$ |
| $J_c$ | Cluster-aggregated, stratum-centred, DF-corrected meat (rank $\le$ number of clusters) |
| $\mathrm{DEFF}$ | Kish global design effect, $N \sum w^2 / (\sum w)^2$ ($\ge 1$) |
| $B$ | Shrinkage / reliability factor, $\sigma_\theta^2 / (\sigma_\theta^2 + \sigma_e^2 / n)$; $B \in (0,1)$ |
| $R_k$ | Between-group share of covariate $k$'s identifying variation; $R_k \in [0, B]$ |
| $\kappa(J)$ | Finite-group correction, $1 - 1/[J(1-B) + B]$; $\to 1$ as $J \to \infty$ |
| $\sigma_\theta$ | Random-effect standard deviation |
| $\tau$ | Random-effect prior precision, $\tau = 1/\sigma_\theta^2$ |
| $c_0$ | DER threshold (default $c_0 = 1.2$): parameter flagged when $\mathrm{DER}_k > c_0$ |

The paper stacks the parameters as $\boldsymbol{\phi} = (\boldsymbol{\beta}^{\mathsf T},
\boldsymbol{\theta}^{\mathsf T})^{\mathsf T}$ of dimension $d = p + J$, with $p$
fixed effects and $J$ random effects; the grand mean $\mu$ is the intercept
coefficient. The full symbol table --- including the bread's block structure
($\mathbf A$, $\mathbf G$, $\mathbf D_H$, the Schur complement $\mathbf S_\beta$)
and the decomposition quantities --- is given in `vignette("theory-decomposition")`.

## What's next?

- **`vignette("theory-sandwich")`** --- *The Declared Variance Target: Bread and
  Meat.* Constructs the numerator $V_{\mathrm{target}} = H^{-1} J_c H^{-1}$: why
  the *posterior* Hessian is the right bread, the cluster-aggregated meat with
  stratum centring and the $C_h/(C_h - 1)$ correction, weight conventions, the
  cluster universe (empty PSUs), and the gaussian score fix.
- **`vignette("theory-decomposition")`** --- *Decomposition Theorems and the
  Conservation Law.* Proves $\mathrm{DER}_{\beta_k} = \mathrm{DEFF}\,(1 - R_k)$
  (Theorem 1), the random-effect factorisation $\mathrm{DER}_\theta = B \cdot
  \mathrm{DEFF} \cdot \kappa(J)$ (Theorem 2), and the conservation identity
  $\mathrm{DER}_\mu + \mathrm{DER}_\theta^{\mathrm{cond}} = \mathrm{DEFF}$, verified
  live with `der_theorem_check()` and `der_decompose()`.
- **`vignette("theory-classification-correction")`** --- the four-tier
  classification, the threshold rationale, and the selective block-Cholesky
  correction that acts only on flagged parameters.
- **Applied:** the [Getting Started](getting-started.html) vignette runs a full
  diagnostic on `nsece_demo` in one call; the other applied vignettes cover
  target choice, the pipeline, and the case study.

## References

Binder, D. A. (1983). On the variances of asymptotically normal estimators from
complex surveys. *International Statistical Review*, *51*(3), 279--292.

Kish, L. (1965). *Survey Sampling*. Wiley.

Lee, J., Williams, M. R., & Savitsky, T. D. (2026). *Design Effect Ratios for
Bayesian Survey Models: A Diagnostic Framework for Identifying Survey-Sensitive
Parameters.* Journal of Survey Statistics and Methodology (submitted).

Savitsky, T. D., & Williams, M. R. (2022). Pseudo Bayesian mixed models under informative sampling. *Journal of Official Statistics*, *38*(4), 901–928. <https://doi.org/10.2478/jos-2022-0039>

Williams, M. R., & Savitsky, T. D. (2021). Uncertainty estimation for pseudo-Bayesian inference under complex sampling. *International Statistical Review*, *89*(1), 72--107. <https://doi.org/10.1111/insr.12376>
