---
title: "Understanding Design Effect Ratios"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A deep dive into the Design Effect Ratio (DER) framework: definition and
  three regimes, why parameters differ in design sensitivity, the decomposition
  theorems linking DER to Kish DEFF and hierarchical shrinkage, the conservation
  law, three-tier classification, threshold sensitivity, and comparison between
  complex survey and equal-weight data.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding Design Effect Ratios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(svyder)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
set.seed(42)

# Consistent colour palette across all svyder vignettes
pal <- c(
  tier_ia    = "#D6604D",
  tier_ib    = "#4393C3",
  tier_ii    = "#1B7837",
  threshold  = "#762A83"
)
```

## Overview

The Design Effect Ratio (DER) is the central diagnostic in svyder.
Understanding *what* it measures, *why* it differs across parameters, and
*how* its decomposition connects survey design theory to Bayesian hierarchical
modelling is essential for making principled decisions about posterior
correction.

In this vignette you will learn how to:

- **Define** the DER and interpret its three regimes.
- **Explain** why parameters in the same model have vastly different DER
  values.
- **Grasp** the intuition behind hierarchical shielding.
- **Apply** the three-tier classification in practice.
- **Verify** the decomposition theorems and conservation law.
- **Assess** sensitivity of the classification to the threshold $\tau$.
- **Compare** DER behaviour under complex survey versus equal-weight designs.

For a hands-on quickstart with the package, see
`vignette("getting-started", package = "svyder")`.


## From DEFF to DER: Building Intuition

### The Kish design effect

Complex surveys use stratification, clustering, and unequal weighting to
efficiently cover a target population.  These design features mean that
observations are no longer independent and identically distributed, so the
effective sample size for estimating a population mean is smaller than the
nominal sample size.

The classical **Kish design effect** (DEFF) quantifies this loss:

$$
\text{DEFF} = \frac{\text{Var}_{\text{design}}(\bar{y})}
                    {\text{Var}_{\text{SRS}}(\bar{y})}
= 1 + \text{CV}_w^2,
$$

where $\text{CV}_w$ is the coefficient of variation of the survey weights.
The effective sample size is $N_{\text{eff}} = N / \text{DEFF}$.

For many national surveys, DEFF ranges from 1.5 to 5 or higher.  A DEFF of
3 means the survey carries the information content of a simple random sample
one-third the nominal size.

### DEFF's limitation for hierarchical models

The Kish DEFF is a useful scalar summary of design complexity, but it has a
fundamental limitation: **it is a single number that applies uniformly to all
parameters**.  In a hierarchical model, different parameters draw their
information from qualitatively different sources.  Applying the same DEFF
correction to a within-cluster coefficient and a random effect is like
prescribing the same medicine for two entirely different conditions.

### DER: parameter-specific design sensitivity

The **Design Effect Ratio** (DER) generalises DEFF to the parameter level.
For the $p$th parameter, DER is defined as:

$$
\text{DER}_p = \frac{[\mathbf{V}_{\text{sand}}]_{pp}}
                     {[\boldsymbol{\Sigma}_{\text{MCMC}}]_{pp}},
$$

where:

- $\mathbf{V}_{\text{sand}}$ is the sandwich variance matrix, the
  design-consistent estimator of the posterior covariance.
- $\boldsymbol{\Sigma}_{\text{MCMC}}$ is the posterior covariance from
  standard MCMC sampling (which ignores the survey design).

The DER asks: *by what factor does the design-consistent variance differ
from the model-based variance for this particular parameter?*

A natural interpretation in terms of effective sample size follows:

$$
N_{\text{eff}}^{\text{design}}(\beta_k) = N / \text{DER}_{\beta_k}.
$$

```{r load-data}
data(nsece_demo)
data(sim_hlr)
```


## Three Regimes of DER

The DER divides parameters into three regimes based on their value relative
to 1.0:

| Regime | Condition | Meaning | Action |
|:-------|:----------|:--------|:-------|
| Under-dispersed | DER $> 1$ | Posterior too narrow | Widen (correct) |
| Calibrated | DER $\approx 1$ | Posterior correct | None needed |
| Over-dispersed | DER $< 1$ | Posterior conservative | Do **not** narrow |

The critical insight is the third regime: when DER $< 1$, the model-based
posterior is already wider than the design-consistent target.  Correcting
such a parameter would *narrow* its credible interval, reducing coverage
below the nominal level.  This is why blanket correction is harmful ---
it forces DER = 1 for every parameter, including those where DER $< 1$
(which describes the vast majority of parameters in hierarchical models).

### Visualising the three regimes

Let us compute DER for both the NSECE-like data (complex survey) and the
balanced HLR data (equal weights) to see all three regimes:

```{r compute-both}
result_nsece <- der_diagnose(
  x           = nsece_demo$draws,
  y           = nsece_demo$y,
  X           = nsece_demo$X,
  group       = nsece_demo$group,
  weights     = nsece_demo$weights,
  psu         = nsece_demo$psu,
  family      = nsece_demo$family,
  sigma_theta = nsece_demo$sigma_theta,
  param_types = nsece_demo$param_types,
  tau         = 1.2
)

result_hlr <- der_diagnose(
  x           = sim_hlr$draws,
  y           = sim_hlr$y,
  X           = sim_hlr$X,
  group       = sim_hlr$group,
  weights     = sim_hlr$weights,
  psu         = sim_hlr$psu,
  family      = sim_hlr$family,
  sigma_theta = sim_hlr$sigma_theta,
  sigma_e     = sim_hlr$sigma_e,
  param_types = sim_hlr$param_types,
  tau         = 1.2
)
```

```{r fig-regimes, fig.height = 5.5, fig.cap = "The three DER regimes illustrated across two datasets. Top: NSECE-like survey data with complex design (DEFF approx 3.5). The within-cluster poverty coefficient (beta[2]) is under-dispersed (DER > 1), while between-cluster fixed effects and random effects are over-dispersed (DER < 1). Bottom: balanced Gaussian HLR with equal weights (DEFF = 1). All parameters are near 1.0, confirming no design effects."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  td_nsece <- tidy.svyder(result_nsece)
  td_hlr   <- tidy.svyder(result_hlr)

  td_nsece$dataset <- "NSECE (complex survey, DEFF ~ 3.5)"
  td_hlr$dataset   <- "Simulated HLR (equal weights, DEFF = 1)"

  td_nsece$param_type <- result_nsece$classification$param_type
  td_hlr$param_type   <- result_hlr$classification$param_type

  # Assign regime labels
  regime_label <- function(der) {
    ifelse(der > 1.2, "Under-dispersed (DER > 1)",
    ifelse(der > 0.8, "Calibrated (DER ~ 1)",
                       "Over-dispersed (DER < 1)"))
  }

  td_nsece$regime <- regime_label(td_nsece$der)
  td_hlr$regime   <- regime_label(td_hlr$der)

  td_all <- rbind(td_nsece, td_hlr)
  td_all$idx <- ave(seq_len(nrow(td_all)),
                    td_all$dataset, FUN = seq_along)

  ggplot(td_all, aes(x = as.numeric(idx), y = der, colour = regime)) +
    geom_hline(yintercept = 1, linetype = "solid", colour = "grey70",
               linewidth = 0.5) +
    geom_hline(yintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.6) +
    geom_point(alpha = 0.7, size = 1.8) +
    scale_colour_manual(
      values = c("Under-dispersed (DER > 1)" = pal["tier_ia"],
                 "Calibrated (DER ~ 1)"      = pal["tier_ib"],
                 "Over-dispersed (DER < 1)"   = pal["tier_ii"]),
      name = "Regime"
    ) +
    scale_y_log10() +
    facet_wrap(~ dataset, ncol = 1, scales = "free_x") +
    labs(x = "Parameter index", y = "DER (log scale)",
         title = "DER Across Two Datasets: Three Regimes") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          strip.text = element_text(face = "bold", size = 11))
}
```

The contrast between the two panels illustrates the DER's discriminating
power.  The NSECE data shows all three regimes clearly separated, while
the equal-weight data clusters entirely in the calibrated regime.


## Why Parameters Differ: The Role of Information Source

The key insight of the DER framework is that design sensitivity depends on
*where a parameter gets its identifying information*.  In a hierarchical
model, this varies dramatically across parameter types.

### Within-cluster variation: fully exposed

A covariate like individual-level poverty varies *within* states.  Its
coefficient $\beta_{\text{poverty}}$ is identified from comparisons among
individuals in the same cluster.  The hierarchical prior on the random
effects $\theta_j$ cannot absorb within-cluster correlation, so the full
survey design effect passes through to this parameter.  Its DER is
approximately equal to DEFF.

### Between-cluster variation: shielded by shrinkage

The intercept and state-level policy indicators are identified from
*between-state* comparisons.  The random effects $\theta_j$ absorb a
fraction of the design-induced between-state correlation through the
shrinkage mechanism.  The DER is attenuated: $\text{DER} \approx
\text{DEFF} \times (1 - B)$, where $B$ is the shrinkage factor.

### Concrete example from NSECE data

```{r param-contrast}
td <- tidy.svyder(result_nsece)
fe_td <- td[1:3, ]

fe_display <- data.frame(
  Parameter     = c("Intercept", "Poverty (CWC)", "Tiered Reim."),
  Type          = nsece_demo$param_types,
  DER           = round(fe_td$der, 4),
  Scale_Factor  = round(fe_td$scale_factor, 4),
  Action        = fe_td$action
)
fe_display
```

```{r contrast-magnitude}
der_within  <- fe_td$der[2]   # poverty_cwc
der_between <- fe_td$der[1]   # intercept

cat(sprintf("Within-cluster DER  (poverty):   %.4f\n", der_within))
cat(sprintf("Between-cluster DER (intercept): %.4f\n", der_between))
cat(sprintf("Ratio:                           %.0fx\n",
            der_within / der_between))
```

The poverty coefficient has a DER roughly 100 times larger than the
intercept --- in the *same model*, from the *same data*.  This dramatic
contrast arises entirely from the information source:  within-cluster
variation is fully exposed to design effects, while between-cluster
variation is shielded by hierarchical shrinkage.

### Visualising parameter-level DER

```{r fig-param-bar, fig.height = 4.5, fig.cap = "DER values for the three fixed effects in the NSECE-like model. The within-cluster poverty coefficient (DER approx 2.6) stands in stark contrast to the between-cluster parameters (DER approx 0.03). The dashed line at DER = 1 marks the calibration reference. Parameters above this line need widening; those below are already conservative."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  fe_bar <- data.frame(
    param = factor(c("Intercept\n(between)", "Poverty CWC\n(within)",
                     "Tiered Reim.\n(between)"),
                   levels = c("Intercept\n(between)", "Poverty CWC\n(within)",
                              "Tiered Reim.\n(between)")),
    der   = fe_td$der,
    type  = nsece_demo$param_types
  )

  ggplot(fe_bar, aes(x = param, y = der, fill = type)) +
    geom_col(width = 0.6, colour = "white", linewidth = 0.4) +
    geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40",
               linewidth = 0.6) +
    geom_hline(yintercept = 1.2, linetype = "dotted",
               colour = pal["threshold"], linewidth = 0.6) +
    geom_text(aes(label = sprintf("%.3f", der)),
              vjust = -0.5, size = 3.8, fontface = "bold") +
    scale_fill_manual(
      values = c(fe_within = pal["tier_ia"],
                 fe_between = pal["tier_ib"]),
      labels = c(fe_within = "Within-cluster",
                 fe_between = "Between-cluster"),
      name = "Information source"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = NULL, y = "Design Effect Ratio",
         title = "Fixed-Effect DER: Within vs Between Cluster") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())
}
```


## The Decomposition Theorems

The DER's value is not arbitrary --- it decomposes into recognisable
components from survey sampling theory and Bayesian statistics.

### Theorem 1: Fixed effects

For the $k$th fixed-effect coefficient $\beta_k$:

$$
\text{DER}_{\beta_k} = \text{DEFF} \times (1 - R_k),
$$

where $R_k \in [0, B]$ is the **protection factor** measuring the fraction
of identifying variation for covariate $k$ that comes from between-group
differences, and $B = \sigma_\theta^2 / (\sigma_\theta^2 + \sigma_e^2 / n)$
is the hierarchical shrinkage factor.

Two important special cases emerge:

| Case | $R_k$ | DER | Interpretation |
|:-----|:-------|:----|:---------------|
| Pure within-cluster | $\approx 0$ | $\approx$ DEFF | Fully exposed |
| Pure between-cluster | $\approx B$ | $\approx$ DEFF$(1 - B)$ | Shielded |

### Theorem 2: Random effects

For the $j$th random effect $\theta_j$:

$$
\text{DER}_j = B_j \times \text{DEFF}_j \times \kappa_j(J),
$$

where:

- $B_j = \sigma_\theta^2 / (\sigma_\theta^2 + \sigma_e^2 / n_j)$ is
  the group-specific shrinkage factor.
- $\text{DEFF}_j$ is the group-specific design effect from unequal weights
  within group $j$.
- $\kappa_j(J) = 1 - (1 - B_j) / [J(1 - B_j) + B_j]$ is the
  finite-group coupling factor (close to 1 for moderate $J$).

The crucial term is $B_j$: the shrinkage factor directly attenuates the
design effect before it reaches the posterior.  With strong shrinkage
($B_j$ small), the random effect borrows heavily from the grand mean,
leaving little room for design effects to distort the posterior.

### Verifying the decomposition

The `der_decompose()` function breaks each parameter's DER into its
constituent factors:

```{r decompose}
decomp <- der_decompose(result_nsece)

# Fixed effects
cat("=== Fixed Effects ===\n")
fe_decomp <- decomp[decomp$param_type != "re", ]
fe_decomp[, c("param", "param_type", "der", "deff_mean",
               "B_mean", "R_k", "der_predicted")]
```

For the within-cluster covariate (`beta[2]`), $R_k \approx 0$, consistent
with its role as a pure within-cluster variable.  For the between-cluster
parameters, $R_k$ is close to $B$, confirming that hierarchical shrinkage
absorbs most of their design sensitivity.

```{r decompose-re}
# Random effects (first 10)
cat("\n=== Random Effects (first 10) ===\n")
re_decomp <- decomp[decomp$param_type == "re", ]
head(re_decomp[, c("param", "der", "deff_mean", "B_mean",
                    "kappa", "der_predicted")], 10)
```

### Checking theoretical predictions

The `der_theorem_check()` function formally compares empirical DER values
against their theoretical predictions:

```{r theorem-check}
check <- der_theorem_check(result_nsece)

# Fixed effects
cat("=== Fixed-Effect Theorem Check ===\n")
fe_check <- check[check$param_type != "re", ]
fe_check[, c("param", "der_empirical", "der_theorem1",
              "relative_error", "theorem_used")]
```

```{r theorem-check-re}
# Random effects (first 10)
cat("\n=== Random-Effect Theorem Check (first 10) ===\n")
re_check <- check[check$param_type == "re", ]
head(re_check[, c("param", "der_empirical", "der_theorem2",
                    "relative_error")], 10)
```

The fixed-effect predictions are typically accurate within a few percent.
Random-effect predictions may show larger discrepancies in non-conjugate
models (e.g., logistic) with substantial coupling between $\beta$ and
$\theta$ in the observed information matrix.  The key point is that the
decomposition correctly predicts the *qualitative ordering* and the
*magnitude* of DER values across parameter types.

### Visualising observed vs predicted DER

```{r fig-decomp-plot, fig.height = 5, fig.cap = "DER decomposition plot: observed DER (from the full sandwich computation) versus predicted DER (from the closed-form decomposition). Points near the 1:1 line indicate good agreement. The two clusters correspond to fixed effects (upper right, blue/red) and random effects (lower left, green)."}
plot(result_nsece, type = "decomposition")
```


## The Conservation Law

An elegant structural result emerges from the decomposition:

$$
\text{DER}_\mu + \text{DER}_\theta^{\text{cond}} = \text{DEFF}.
$$

The total design sensitivity is **conserved**.  The hierarchical prior does
not create or destroy design effects --- it *redistributes* them between
the global mean and the group-level parameters.

### Intuition

Think of the total design effect DEFF as a fixed budget of design
sensitivity.  The hierarchical prior redistributes this budget:

- **Random effects** ($\theta_j$) are protected by shrinkage: the prior
  absorbs a fraction $B$ of their design sensitivity.
- **The grand mean** ($\mu$) inherits the released sensitivity: it becomes
  *more* exposed to design effects.

Protecting random effects has a cost --- increased exposure for fixed
effects.  There is no free lunch.

### Empirical verification

```{r conservation}
thm_check <- der_theorem_check(result_nsece)

conservation <- attr(thm_check, "conservation_law")
if (!is.null(conservation)) {
  cat("Conservation Law Verification\n")
  cat("-----------------------------\n")
  cat(sprintf("DER(mu)          = %.4f\n", conservation$der_mu))
  cat(sprintf("DER(theta) mean  = %.4f\n", conservation$der_theta_mean))
  cat(sprintf("Sum              = %.4f\n", conservation$conservation_sum))
  cat(sprintf("Mean DEFF        = %.4f\n", conservation$deff_mean))
  cat(sprintf("Relative error   = %.4f\n", conservation$relative_error))
}
```

The conservation law holds approximately because the NSECE-like data is
unbalanced.  With perfectly balanced groups and a conjugate normal model,
the identity would be exact.

### Visualising the conservation law

```{r fig-conservation, fig.height = 4, fig.cap = "The conservation law visualised as a stacked bar. The total design effect (DEFF) is partitioned into the design sensitivity absorbed by the grand mean (DER_mu) and the sensitivity absorbed by random effects (DER_theta). The prior redistributes but does not eliminate design sensitivity."}
if (requireNamespace("ggplot2", quietly = TRUE) && !is.null(conservation)) {
  library(ggplot2)

  cons_df <- data.frame(
    component = factor(c("DER(mu)", "DER(theta)"),
                        levels = c("DER(theta)", "DER(mu)")),
    value     = c(conservation$der_mu, conservation$der_theta_mean)
  )

  ggplot(cons_df, aes(x = "DEFF Budget", y = value, fill = component)) +
    geom_col(width = 0.5, colour = "white", linewidth = 0.5) +
    geom_hline(yintercept = conservation$deff_mean, linetype = "dashed",
               colour = "grey30", linewidth = 0.7) +
    annotate("text", x = 1.4, y = conservation$deff_mean,
             label = sprintf("DEFF = %.2f", conservation$deff_mean),
             hjust = 0, size = 4, fontface = "italic") +
    geom_text(aes(label = sprintf("%.3f", value)),
              position = position_stack(vjust = 0.5),
              colour = "white", fontface = "bold", size = 4.5) +
    scale_fill_manual(
      values = c("DER(mu)" = pal["tier_ib"],
                 "DER(theta)" = pal["tier_ii"]),
      name = NULL
    ) +
    labs(x = NULL, y = "Design Sensitivity",
         title = "Conservation Law: DER(mu) + DER(theta) = DEFF") +
    coord_flip() +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.y = element_blank(),
          axis.text.y = element_blank())
}
```


## The Three-Tier Classification

The decomposition results naturally motivate a classification of parameters
into three tiers:

| Tier | Parameter type | Approx. DER | Action |
|:-----|:---------------|:------------|:-------|
| **I-a** | Fixed effects (within-cluster) | DEFF | Sandwich correction |
| **I-b** | Fixed effects (between-cluster) | DEFF$(1 - B)$ | Monitor; often $< 1$ |
| **II** | Random effects ($\theta_j$) | $B \times$ DEFF $\times \kappa$ | Do not correct |

### Tier I-a: Survey-dominated parameters

Within-cluster fixed effects are the primary candidates for correction.
Their DER is approximately equal to DEFF because the hierarchical prior
cannot absorb within-cluster design effects.  In the NSECE example, the
poverty coefficient falls squarely in this tier.

### Tier I-b: Partially protected parameters

Between-cluster fixed effects benefit from partial protection through the
shrinkage mechanism.  Their DER is typically well below 1.0 when shrinkage
is strong ($B$ near 1), making correction unnecessary.  However, in designs
with very large DEFF or weak shrinkage, these parameters can occasionally
exceed the threshold.

### Tier II: Protected random effects

Random effects are always partially protected by the prior.  Their DER
is proportional to the shrinkage factor $B$ times DEFF, which is typically
well below 1.0.  Correcting these parameters would narrow their posteriors
and degrade coverage --- they should *never* be corrected.

### Classification output from NSECE data

```{r classify-output}
td <- tidy.svyder(result_nsece)

# Show tier distribution
cat("Tier distribution:\n")
print(table(td$tier))

cat("\nFixed effects with classification:\n")
td[1:3, c("term", "der", "tier", "action", "flagged", "scale_factor")]
```

```{r classify-re}
cat("\nRandom effects (first 10):\n")
td[4:13, c("term", "der", "tier", "action", "flagged")]
```

### Why $\tau = 1.2$?

The default threshold $\tau = 1.2$ means a parameter is flagged only when
its model-based posterior variance is more than 20% below the
design-consistent target.  This choice balances two concerns:

1. **Coverage**: Parameters with DER substantially above 1 have poor
   frequentist coverage.  A 20% discrepancy ($\tau = 1.2$) corresponds
   to interval widths that are about 10% too narrow
   ($\sqrt{1.2} \approx 1.095$), which begins to matter for applications
   requiring valid marginal coverage.

2. **Stability**: Setting $\tau$ too close to 1.0 risks flagging
   parameters due to Monte Carlo noise rather than genuine design effects.
   The 20% buffer provides protection against spurious flagging.

### Sensitivity analysis

The `der_sensitivity()` function evaluates how the classification changes
across a range of thresholds:

```{r sensitivity}
sens <- der_sensitivity(result_nsece, tau_range = seq(0.5, 3.0, by = 0.1))
sens_display <- sens[, c("tau", "n_flagged", "pct_flagged")]
sens_display$pct_flagged <- round(sens_display$pct_flagged * 100, 1)
colnames(sens_display) <- c("tau", "n_flagged", "pct_flagged (%)")
sens_display
```

```{r fig-sensitivity, fig.height = 4.5, fig.cap = "Threshold sensitivity analysis. The number of flagged parameters (y-axis) is plotted against the classification threshold tau (x-axis). The classification is remarkably stable: only 1 parameter (poverty coefficient) is flagged across a wide range of thresholds from 0.5 to 2.5. This stability reflects the clean separation between the within-cluster DER (approx 2.6) and all other DER values (below 1.0)."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  ggplot(sens, aes(x = tau, y = n_flagged)) +
    geom_step(colour = pal["tier_ia"], linewidth = 0.9) +
    geom_point(colour = pal["tier_ia"], size = 2) +
    geom_vline(xintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.6) +
    annotate("text", x = 1.25, y = max(sens$n_flagged) * 0.85,
             label = expression(paste("default ", tau, " = 1.2")),
             hjust = 0, colour = pal["threshold"], size = 3.5) +
    scale_x_continuous(breaks = seq(0.5, 3.0, by = 0.5)) +
    scale_y_continuous(breaks = seq(0, max(sens$n_flagged) + 1, by = 1)) +
    labs(x = expression("Threshold " * tau),
         y = "Number of flagged parameters",
         title = "Sensitivity of Classification to Threshold") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
```

In this example, the classification is highly robust: the poverty
coefficient is the only parameter flagged for all thresholds up to
approximately 2.5, and no other parameters enter the flagged set.  This
stability is characteristic of models with clear separation between
within-cluster and between-cluster parameters.


## Comparing Equal-Weight vs Complex Survey

To build further intuition, let us compare the DER diagnostic on the two
bundled datasets side by side.

### Equal-weight data (sim_hlr)

```{r hlr-detail}
result_hlr

cat("\nFixed-effect DER values:\n")
round(result_hlr$der[1:2], 4)

cat("\nRandom-effect DER values (first 5):\n")
round(result_hlr$der[3:7], 4)
```

With equal weights ($\text{DEFF} = 1$), all DER values cluster near 1.0
or below.  The hierarchical model is already well-calibrated without any
design correction.

### Summary comparison

```{r comparison-table}
gl_nsece <- glance.svyder(result_nsece)
gl_hlr   <- glance.svyder(result_hlr)

comp_df <- data.frame(
  Metric = c("N", "J", "d (total params)", "Mean DEFF",
             "Mean B (shrinkage)", "DER min", "DER max",
             "Parameters flagged"),
  NSECE = c(gl_nsece$n_obs, gl_nsece$n_groups, gl_nsece$n_params,
            round(gl_nsece$mean_deff, 3), round(gl_nsece$mean_B, 3),
            round(gl_nsece$der_min, 4), round(gl_nsece$der_max, 4),
            gl_nsece$n_flagged),
  sim_hlr = c(gl_hlr$n_obs, gl_hlr$n_groups, gl_hlr$n_params,
              round(gl_hlr$mean_deff, 3), round(gl_hlr$mean_B, 3),
              round(gl_hlr$der_min, 4), round(gl_hlr$der_max, 4),
              gl_hlr$n_flagged)
)
comp_df
```

### Side-by-side profile plots

```{r fig-side-by-side, fig.height = 5, fig.cap = "Side-by-side DER profiles for two datasets. Top: the NSECE-like data shows one prominent spike (beta[2], DER approx 2.6) above the threshold, with all other parameters well below 1.0. Bottom: the equal-weight sim_hlr data shows all DER values near or below 1.0, confirming that the DER correctly detects the absence of design effects."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  td_nsece <- tidy.svyder(result_nsece)
  td_hlr   <- tidy.svyder(result_hlr)

  td_nsece$dataset <- "NSECE (complex survey)"
  td_nsece$idx     <- seq_len(nrow(td_nsece))
  td_nsece$tier_colour <- result_nsece$classification$tier

  td_hlr$dataset <- "sim_hlr (equal weights)"
  td_hlr$idx     <- seq_len(nrow(td_hlr))
  td_hlr$tier_colour <- result_hlr$classification$tier

  td_both <- rbind(td_nsece, td_hlr)

  ggplot(td_both, aes(x = idx, y = der, colour = tier_colour)) +
    geom_hline(yintercept = 1, linetype = "solid", colour = "grey70",
               linewidth = 0.4) +
    geom_hline(yintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.6) +
    geom_point(alpha = 0.7, size = 1.8) +
    scale_colour_manual(
      values = c("I-a" = pal["tier_ia"],
                 "I-b" = pal["tier_ib"],
                 "II"  = pal["tier_ii"]),
      labels = c("I-a" = "Tier I-a (within FE)",
                 "I-b" = "Tier I-b (between FE)",
                 "II"  = "Tier II (random effects)"),
      name = "Tier"
    ) +
    facet_wrap(~ dataset, ncol = 1, scales = "free") +
    labs(x = "Parameter index", y = "DER",
         title = "DER Profiles: Complex Survey vs Equal Weights") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          strip.text = element_text(face = "bold", size = 11))
}
```


## Effective Sample Size Interpretation

The DER admits a natural interpretation through the design-effective
sample size:

$$
N_{\text{eff}}^{\text{design}}(\beta_k) = \frac{N}{\text{DER}_{\beta_k}}.
$$

```{r neff}
N  <- nsece_demo$N
td <- tidy.svyder(result_nsece)

fe_td <- td[1:3, ]
fe_td$n_eff <- round(N / fe_td$der)

neff_df <- data.frame(
  Parameter = c("Intercept", "Poverty (CWC)", "Tiered Reim."),
  DER       = round(fe_td$der, 4),
  N_eff     = fe_td$n_eff
)
neff_df
```

The poverty coefficient ($\text{DER} \approx 2.6$) has an effective sample
size of about $N / 2.6$, reflecting substantial information loss from the
survey design.  The between-state parameters have effective sample sizes
much larger than $N$ because the hierarchical prior *adds* information
beyond what the data alone provide --- their DER below 1.0 reflects this
prior-augmented precision.

```{r fig-neff, fig.height = 4, fig.cap = "Design-effective sample size for the three fixed effects. The poverty coefficient has an effective sample size of roughly N/2.6, while the between-cluster parameters have effective sample sizes far exceeding N due to prior-augmented precision (DER < 1). The dashed grey line marks the nominal sample size N = 6785."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  neff_df$Parameter <- factor(neff_df$Parameter,
                               levels = c("Intercept", "Poverty (CWC)",
                                          "Tiered Reim."))
  neff_df$fill_col <- c("between", "within", "between")

  ggplot(neff_df, aes(x = Parameter, y = N_eff, fill = fill_col)) +
    geom_col(width = 0.5, colour = "white", linewidth = 0.4) +
    geom_hline(yintercept = N, linetype = "dashed", colour = "grey40",
               linewidth = 0.6) +
    annotate("text", x = 3.4, y = N, label = sprintf("N = %d", N),
             hjust = 1.1, vjust = -0.5, colour = "grey40", size = 3.5) +
    scale_fill_manual(
      values = c(within = pal["tier_ia"], between = pal["tier_ib"]),
      labels = c(within = "Within-cluster", between = "Between-cluster"),
      name = "Information source"
    ) +
    scale_y_continuous(labels = scales::comma_format()) +
    labs(x = NULL, y = expression(N[eff]^design),
         title = "Design-Effective Sample Size by Parameter") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())
}
```


## Algorithm vs Closed-Form: When Do They Diverge?

The decomposition formulas (Theorems 1 and 2) are derived under simplifying
assumptions: balanced groups, a conjugate normal model, and diagonal
observed information.  In practice:

- **The closed-form formulas correctly identify** which parameters are
  design-sensitive and the qualitative ordering of DER values across
  parameter types.

- **The closed-form formulas may not predict exact magnitudes** when groups
  are unbalanced, the model is non-conjugate (e.g., logistic), or there
  is substantial coupling between $\beta$ and $\theta$ in the observed
  information matrix.

For this reason, svyder always computes **exact DER values** via the full
sandwich variance (Algorithm 1 in the paper).  The decomposition formulas
serve as *interpretive tools* --- they explain *why* a parameter has its
observed DER value, but the algorithm provides the definitive answer.

```{r fig-exact-vs-approx, fig.height = 5, fig.cap = "Agreement between exact (Algorithm 1) and predicted (closed-form) DER values. Points near the 1:1 line indicate good agreement. The fixed-effect predictions (Theorem 1) are very accurate. Random-effect predictions (Theorem 2) show moderate scatter due to the non-conjugate logistic model and unbalanced groups."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  check <- der_theorem_check(result_nsece)
  check$predicted <- ifelse(!is.na(check$der_theorem1),
                            check$der_theorem1,
                            check$der_theorem2)
  check$type_label <- ifelse(check$param_type == "re",
                              "Random effects (Theorem 2)",
                       ifelse(check$param_type == "fe_within",
                              "FE within (Theorem 1)",
                              "FE between (Theorem 1)"))

  rng <- range(c(check$der_empirical, check$predicted), na.rm = TRUE)

  ggplot(check, aes(x = predicted, y = der_empirical,
                    colour = type_label)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey50", linewidth = 0.5) +
    geom_point(alpha = 0.7, size = 2) +
    scale_colour_manual(
      values = c("FE within (Theorem 1)" = pal["tier_ia"],
                 "FE between (Theorem 1)" = pal["tier_ib"],
                 "Random effects (Theorem 2)" = pal["tier_ii"]),
      name = NULL
    ) +
    labs(x = "Predicted DER (closed-form)", y = "Empirical DER (Algorithm 1)",
         title = "Exact vs Closed-Form DER Predictions") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())
}
```


## DER Distribution Across Random Effects

With 51 random effects in the NSECE model, it is instructive to examine the
distribution of DER values and relate them to the per-group shrinkage
factors:

```{r fig-re-distribution, fig.height = 5, fig.cap = "Distribution of DER values across 51 state-level random effects (top) and the relationship between per-group shrinkage factor B_j and DER (bottom). All random-effect DER values are well below the threshold (dashed line), confirming that hierarchical shrinkage shields them from design effects. The positive correlation between B_j and DER reflects the decomposition: DER_j = B_j x DEFF_j x kappa_j."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  re_der <- result_nsece$der[4:54]
  re_B   <- result_nsece$B_j

  re_df <- data.frame(
    state  = seq_along(re_der),
    der    = as.numeric(re_der),
    B_j    = re_B
  )

  p1 <- ggplot(re_df, aes(x = der)) +
    geom_histogram(bins = 20, fill = pal["tier_ii"], colour = "white",
                   linewidth = 0.3, alpha = 0.85) +
    geom_vline(xintercept = 1.2, linetype = "dashed",
               colour = pal["threshold"], linewidth = 0.6) +
    labs(x = "DER", y = "Count",
         subtitle = "Distribution of random-effect DER values") +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())

  p2 <- ggplot(re_df, aes(x = B_j, y = der)) +
    geom_point(colour = pal["tier_ii"], alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey40",
                linetype = "dashed", linewidth = 0.6) +
    labs(x = expression("Shrinkage factor " * B[j]),
         y = "DER",
         subtitle = expression("DER" [j] * " vs shrinkage factor " * B[j])) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())

  if (requireNamespace("patchwork", quietly = TRUE)) {
    library(patchwork)
    p1 / p2 +
      plot_annotation(
        title = "Random-Effect DER: Distribution and Shrinkage Relationship",
        theme = theme(plot.title = element_text(size = 13, face = "bold"))
      )
  } else {
    print(p1)
    print(p2)
  }
}
```


## Key Takeaways

The DER framework rests on five key insights:

1. **DER is parameter-specific.**
   Unlike the classical DEFF, which is a single number for the entire
   survey, DER captures the design sensitivity of each individual parameter.
   Parameters in the same model can differ by a factor of 100 in their
   design sensitivity.

2. **Hierarchical shrinkage shields random effects.**
   The prior $\theta_j \sim N(0, \sigma_\theta^2)$ directly attenuates the
   design effect through the shrinkage factor $B$.  Stronger shrinkage
   means more protection.

3. **The conservation law means you cannot have it all.**
   Protecting random effects comes at the cost of increased design
   sensitivity for fixed effects.  The total design sensitivity is
   conserved: $\text{DER}_\mu + \text{DER}_\theta = \text{DEFF}$.

4. **Selective correction preserves the benefits of Bayesian modelling.**
   By correcting only the flagged parameters (typically a small minority),
   svyder preserves the variance gains from hierarchical shrinkage for the
   vast majority of parameters.

5. **The algorithm is exact; the formulas are interpretive.**
   The sandwich computation (Algorithm 1) provides exact DER values for
   any model.  The decomposition theorems explain *why* each parameter has
   its observed DER value, connecting survey design theory to Bayesian
   hierarchical modelling.


## What's Next?

| Vignette | What you will learn |
|:---------|:--------------------|
| *Getting Started* | Quick tutorial, installation, complete API walkthrough |
| *Advanced Workflows* | Custom sandwich matrices, brms/rstanarm integration, multi-level clustering comparisons |
| *Simulation Studies* | Verifying DER coverage properties through Monte Carlo simulation |

For a hands-on tutorial with the full API, see
`vignette("getting-started", package = "svyder")`.
